---
title: "Question 1 – ABC Fund vs Benchmark and Peers"
author: "Precious Nhamo"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "default"
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)

# Load required packages
pacman::p_load(
  readr, dplyr, tidyr, ggplot2, lubridate, tbl2xts,
  PerformanceAnalytics, RcppRoll, zoo, fmxdat
)

# Load data and create analysis dataset
Benchmark <- read_rds("data/Capped_SWIX.rds")
Active_Managers <- read_rds("data/Active_Managers.rds")
ABC_Fund <- read_rds("data/ABC_Fund.rds")

start_date <- as.Date("2012-07-31")
end_date <- as.Date("2025-10-31")

analysis_data <- ABC_Fund %>%
  filter(date >= start_date, date <= end_date) %>%
  select(date, ABC_Returns = Returns) %>%
  left_join(
    Benchmark %>% filter(date >= start_date, date <= end_date) %>% select(date, Benchmark_Returns = Returns),
    by = "date"
  ) %>%
  left_join(
    Active_Managers %>% 
      filter(date >= start_date, date <= end_date) %>%
      group_by(date) %>%
      summarise(Peers_Avg_Returns = mean(Returns, na.rm = TRUE), .groups = "drop"),
    by = "date"
  )
```

## Rolling 3-year Beta vs Benchmark

```{r rolling-beta}
# Prepare returns data
Ra_xts <- analysis_data %>%
  arrange(date) %>%
  select(date, ABC_Fund = ABC_Returns, Peer_Average = Peers_Avg_Returns) %>%
  mutate(across(-date, ~ coalesce(., 0))) %>%
  tbl_xts(cols_to_xts = c("ABC_Fund", "Peer_Average"))

Rb_xts <- analysis_data %>%
  arrange(date) %>%
  select(date, Benchmark = Benchmark_Returns) %>%
  mutate(Benchmark = coalesce(Benchmark, 0)) %>%
  tbl_xts(cols_to_xts = "Benchmark")

# Plot rolling beta
PerformanceAnalytics::chart.RollingRegression(
  Ra = Ra_xts, Rb = Rb_xts, width = 36,
  attribute = "Beta", legend.loc = "topleft",
  main = "Rolling 3-year Beta vs Benchmark"
)
```

**Analysis:** ABC Fund shows consistently lower beta than peer average, indicating less market risk exposure. Both series demonstrate relative stability over time with ABC maintaining beta around 0.6-0.7 range.

## Beta Distribution: ABC vs Peers

```{r beta-distribution}
# Calculate rolling betas for comparison
Benchmark_roll <- Benchmark %>%
  filter(date >= start_date, date <= end_date) %>%
  arrange(date) %>%
  mutate(
    bench_roll = RcppRoll::roll_prod(1 + Returns, n = 36, fill = NA, align = "right")^(12/36) - 1,
    Y = format(date, "%Y")
  ) %>%
  select(date, Y, bench_roll)

Active_Beta <- Active_Managers %>%
  filter(date >= start_date, date <= end_date) %>%
  arrange(date) %>%
  group_by(Name) %>%
  mutate(
    active_roll = RcppRoll::roll_prod(1 + Returns, n = 36, fill = NA, align = "right")^(12/36) - 1,
    Y = format(date, "%Y")
  ) %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y, Name) %>%
  summarise(AM_Beta = coef(lm(active_roll ~ bench_roll))[2], .groups = "drop")

ABC_Beta <- ABC_Fund %>%
  filter(date >= start_date, date <= end_date) %>%
  arrange(date) %>%
  mutate(
    ABC_roll = RcppRoll::roll_prod(1 + Returns, n = 36, fill = NA, align = "right")^(12/36) - 1,
    Y = format(date, "%Y")
  ) %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y) %>%
  summarise(ABC_Beta = coef(lm(ABC_roll ~ bench_roll))[2], .groups = "drop")

# Create boxplot
Main_box <- Active_Beta %>%
  left_join(ABC_Beta, by = "Y") %>%
  filter(Y >= "2016")

Main_box_plot <- Main_box %>%
  ggplot() +
  geom_boxplot(aes(x = factor(Y), y = AM_Beta), fill = "lightblue", alpha = 0.7) +
  geom_point(aes(x = factor(Y), y = ABC_Beta), color = "red", size = 3, shape = 18) +
  labs(
    title = "3-Year Rolling Beta Distribution",
    subtitle = "ABC Fund (red) vs Peer Distribution",
    x = "Year", y = "Beta"
  ) +
  theme_minimal()

Main_box_plot
```

**Analysis:** ABC Fund consistently positions in the lower quartile of peer beta distribution, confirming systematic lower market risk exposure compared to industry peers.

## Performance Statistics

```{r performance-stats}
returns_xts <- merge(Rb_xts, Ra_xts, join = "inner")

# Key performance metrics
PerformanceAnalytics::table.Stats(returns_xts)
```

```{r risk-metrics}
# Risk-adjusted performance
PerformanceAnalytics::table.SFM(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"], Rf = 0
)
```

**Analysis:** ABC Fund demonstrates higher alpha (0.0056 vs 0.0001) and lower beta (0.616 vs 0.845) compared to peers. However, lower R-squared (0.576 vs 0.951) suggests more idiosyncratic risk.

## Cumulative Performance

```{r cumulative-returns}
cum_df <- analysis_data %>%
  arrange(date) %>%
  transmute(
    date, Benchmark = Benchmark_Returns,
    ABC_Fund = ABC_Returns, Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  group_by(Series) %>%
  mutate(
    ret = coalesce(ret, 0),
    CumRet = cumprod(1 + ret)
  )

ggplot(cum_df, aes(x = date, y = CumRet, color = Series)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Returns Comparison", y = "Growth of R1", x = "") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

**Analysis:** ABC Fund shows strong outperformance against both benchmark and peer average over the full period, with particularly strong performance in recent years.

## Rolling Returns & Volatility

```{r rolling-metrics}
# Rolling returns
roll_df <- analysis_data %>%
  arrange(date) %>%
  transmute(date, Benchmark = Benchmark_Returns, 
            ABC_Fund = ABC_Returns, Peer_Average = Peers_Avg_Returns) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  group_by(Series) %>%
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right")^(12/36) - 1
  ) %>%
  filter(!is.na(RollRets))

ggplot(roll_df, aes(x = date, y = RollRets, color = Series)) +
  geom_line() +
  labs(title = "Rolling 3-Year Annualized Returns", y = "Return", x = "") +
  theme_minimal()

# Rolling volatility
plot_dlog <- analysis_data %>%
  arrange(date) %>%
  transmute(date, Benchmark = Benchmark_Returns,
            ABC_Fund = ABC_Returns, Peer_Average = Peers_Avg_Returns) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  group_by(Series) %>%
  mutate(
    logret = log(1 + ret),
    RollSD = RcppRoll::roll_sd(logret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  filter(!is.na(RollSD))

ggplot(plot_dlog, aes(x = date, y = RollSD, color = Series)) +
  geom_line() +
  labs(title = "Rolling 3-Year Annualized Volatility", y = "Volatility", x = "") +
  theme_minimal()
```

**Analysis:** ABC Fund shows more stable rolling returns with lower volatility compared to benchmark, demonstrating better risk-adjusted performance characteristics.

## Risk-Return Profile

```{r risk-return}
PerformanceAnalytics::chart.RiskReturnScatter(
  returns_xts, 
  main = "Risk-Return Comparison"
)
```

**Analysis:** The scatter plot confirms ABC Fund's position in the favorable upper-left quadrant - delivering higher returns with lower risk compared to both benchmark and peer average.

## Summary

- **Risk Profile:** ABC Fund maintains systematically lower beta (0.62) vs peers (0.85)
- **Performance:** Positive alpha generation (0.07 annualized) with superior risk-adjusted returns
- **Consistency:** Stable outperformance across multiple time horizons
- **Efficiency:** Better risk-return tradeoff demonstrated in scatter plot positioning
```

