
---
title: "Q2"
output: html_document
---

To answer the first part of Question 2 on whether smaller stocks are more volatile than larger stocks, I focus on the last decade(2015-2024) of daily returns and compare volatility across size groups(Small , Mid , Large Caps). I work with the `Hold_Rets_Sectors` data set, which contains daily returns, holdings weights, size group (`Group`) and sector labels for JSE stocks.

In the lecturer's notes, volatility (risk) is defined and measured using the standard deviation of returns, and it is emphasised that:

* Standard deviation should be annualised when comparing different time periods.
* The usual shortcut ( \text{SD} \times \sqrt{12} ) is only theoretically correct when using **logarithmic** returns, because annual log return is the sum of monthly log returns.

I proceed as follows:

1. I construct value weighted indices for each size group (Small Caps, Mid Caps and Large Caps).
2. I convert their monthly simple returns to log returns using (\log(1 + r)).
3. I compute a 36-month rolling standard deviation of these monthly log returns.
4. I annualise this rolling standard deviation by multiplying by (\sqrt{12}).

This yields a rolling annualised standard deviation of returns for each size group .If the Small Cap series has a consistently higher annualised standard deviation, I can conclude that smaller stocks have been more volatile than larger stocks locally.

### Load libraries and raw data

```{r load-data-Q2, echo = TRUE, message = FALSE, warning = FALSE}
# Load required packages
pacman::p_load(
  readr,
  tidyverse,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  rmsfuns,
  fmxdat,
  DescTools,
  RcppRoll,
  knitr,
  kableExtra
)

# Load the holdings and returns data with sector and size group labels
Hold_Rets <- read_rds("data/Hold_Rets_Sectors.rds") %>%
  mutate(
    # Make the size and sector groups explicit instead of leaving them as NA
    Group  = if_else(is.na(Group),  "Unclassified", Group),
    Sector = if_else(is.na(Sector), "Unclassified", Sector)
  )

head(Hold_Rets)
```

Here I explicitly rename `NA` groups to `"Unclassified"` so they do not silently disappear from the data. Later, when I compare volatility across size groups, I restrict the analysis to the proper size categories (Small_Caps, Mid_Caps, Large_Caps) and treat "Unclassified" separately if needed.

### Data Preparation and Index Construction

```{r prep-and-index-Q2, echo = TRUE}
options(dplyr.summarise.inform = FALSE)

# Filter data for the past decade and main size groups
Hold_Rets_decade <- 
  Hold_Rets %>% 
  filter(Group %in% c("Small_Caps", "Mid_Caps", "Large_Caps","Unclassified")) %>% 
  mutate(year = year(date))

# Construct daily value-weighted size index returns
idx_daily <- 
  Hold_Rets_decade %>% 
  group_by(date, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(Tickers = Group) %>%      # so Tickers is "all groups"
  arrange(Tickers, date)

# Convert to monthly returns
idx <- 
  idx_daily %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Tickers, YM) %>% 
  # pick last date of the month and compute monthly return
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Tickers, date)

# Monthly value-weighted returns by Sector and Group (size)
idx_sector_size <- 
  Hold_Rets_decade %>% 
  group_by(date, Sector, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Sector, Group, YM) %>% 
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Sector, Group, date)

head(idx)
```

### Panel 1: All Cumulative Returns Graphs

```{r cumulative-returns-panel, echo = TRUE}
# Cumulative returns by size group
gg_cum <- 
  idx %>% 
  arrange(date) %>% 
  group_by(Tickers) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Tickers)) +
  scale_colour_manual(
    values = c(
      "Large_Caps"   = "#1f77b4",
      "Mid_Caps"     = "#2ca02c",
      "Small_Caps"   = "#d62728",
      "Unclassified" = "grey40"
    )) +
  labs(
    title    = "Cumulative Returns by Cap Size Group",
    subtitle = "Growth of 1 rand invested in each group (rebased to 1 at first observation)",
    caption  = "Note: Each size cap is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  )

gg_cum

# Log cumulative plot by size group
gg_cum_log <- 
  gg_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_log

# Cumulative returns by sector and size
idx_sector_size_cum <- 
  idx_sector_size %>% 
  arrange(Sector, Group, date) %>% 
  group_by(Sector, Group) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup()

gg_cum_sector <- 
  idx_sector_size_cum %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Group)) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title    = "Cumulative Returns by Sector and Size Group",
    subtitle = "Growth of 1 rand invested in each sector-size group (rebased to 1 at first observation)",
    caption  = "Note: Each sector-size group is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

gg_cum_sector

# Log cumulative plot by sector
gg_cum_sector_log <- 
  gg_cum_sector +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum_sector$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_sector_log

# Time series chart of sector returns using PerformanceAnalytics
sector_xts <- 
  idx_sector_size %>% 
  filter(Group %in% c("Large_Caps", "Mid_Caps", "Small_Caps")) %>% 
  select(date, Sector, Group, ret) %>% 
  mutate(Sector_Group = paste(Sector, Group, sep = "_")) %>% 
  select(-Sector, -Group) %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Sector_Group)

# Chart time series of selected sector-size combinations
chart.TimeSeries(
  sector_xts[, c("Industrials_Large_Caps", "Resources_Large_Caps", "Financials_Large_Caps")],
  main = "Large Cap Returns by Sector",
  legend.loc = "topright",
  colorset = 1:3,
  lwd = 2,
  ylab = "Monthly Returns"
)
```

### Panel 2: All Volatility Analysis Graphs

```{r volatility-panel, echo = TRUE}
# Annualized returns bar plot
dfplot <- 
  bind_rows(
    # 6 months - do NOT annualise 
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% 
      mutate(Freq = "A"),
    
    # 1 year
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 12) - 1) %>% 
      mutate(Freq = "B"),
    
    # 3 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 36) - 1) %>% 
      mutate(Freq = "C"),
    
    # 5 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 60) - 1) %>% 
      mutate(Freq = "D")
  )

to_string <- as_labeller(c(
  `A` = "6 Months", 
  `B` = "1 Year", 
  `C` = "3 Years", 
  `D` = "5 Years"
))

g_ann <- 
  dfplot %>% 
  ggplot() + 
  geom_bar(aes(Tickers, mu, fill = Tickers), stat = "identity") + 
  facet_wrap(~Freq, labeller = to_string, nrow = 1) + 
  labs(
    x = "",
    y = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualized terms."
  ) + 
  fmx_fills() + 
  geom_label(
    aes(Tickers, mu, label = paste0(round(mu, 4) * 100, "%")),
    size = ggpts(8),
    alpha = 0.35,
    fontface = "bold",
    nudge_y = 0.002
  ) + 
  theme_fmx(
    CustomCaption = TRUE,
    title.size = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size = ggpts(30),
    axis.size = ggpts(37),
    legend.size = ggpts(35),
    legend.pos = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

# Rolling returns analysis by size
plotdf <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll <- 
  plotdf %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Main volatility analysis - rolling SD by size
plot_dlog <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(ret = log(1 + ret)) %>% 
  mutate(
    RollSD = RcppRoll::roll_sd(ret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(RollSD)) %>% 
  mutate(
    SizeBucket = if_else(Tickers %in% c("Small_Caps", "Large_Caps"),
                         "focus", "other"),
    Tickers    = forcats::fct_relevel(Tickers, "Small_Caps", "Large_Caps")
  )

g_sd <- 
  plot_dlog %>% 
  ggplot() +
  geom_line(
    aes(date, RollSD, colour = Tickers, alpha = SizeBucket, size = SizeBucket)
  ) +
  scale_alpha_manual(values = c(focus = 1,   other = 0.25), guide = "none") +
  scale_size_manual( values = c(focus = 1.8, other = 0.7),   guide = "none") +
  labs(
    title   = "Rolling 3 Year Annualized SD by Size Bucket",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year SD (Ann.)",
    caption = "Note:\nAnnualized SD computed on log monthly returns."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_sd, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = FALSE)

# Rolling returns analysis by sector and size
plotdf_sector <- 
  idx_sector_size %>% 
  group_by(Sector, Group) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll_sector <- 
  plotdf_sector %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Group), alpha = 0.7, size = 1.25) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll_sector, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Sector-level volatility analysis
plot_sd_sector <- 
  Hold_Rets_decade %>%
  group_by(date, Sector, Group) %>%
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(YM = format(date, "%Y%B")) %>%
  arrange(Sector, Group, date) %>%
  group_by(Sector, Group, YM) %>%
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>%
  arrange(Sector, Group, date) %>%
  group_by(Sector, Group) %>%
  mutate(
    ret_log = log1p(ret),
    RollSD  = RcppRoll::roll_sd(ret_log, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  ungroup() %>%
  filter(!is.na(RollSD))

g_sd_sector <- 
  plot_sd_sector %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Group), size = 1.3, alpha = 0.9) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualised SD by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3-year SD (Ann.)"
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5)
  ) +
  fmx_cols()

finplot(
  g_sd_sector,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)
```

### Panel 3: All Drawdown Analysis Graphs

```{r drawdown-panel, echo = TRUE}
# Convert to xts for PerformanceAnalytics
idxxts <- 
  idx %>% 
  filter(Tickers != "Unclassified") %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Tickers)

# Main drawdown chart by size group
chart.Drawdown(
  idxxts[, c("Large_Caps", "Mid_Caps", "Small_Caps")],
  geometric  = TRUE,
  legend.loc = "bottomleft",
  colorset   = 1:3,
  lwd        = 3,
  main       = "Peak to trough drawdowns by size group",
  ylab       = "Drawdown from peak"
)

# Sector drawdowns function
draw_sector_dd <- function(sec) {
  
  groups_to_plot <- c("Large_Caps", "Mid_Caps", "Small_Caps")
  
  sec_xts <- 
    idx_sector_size %>% 
    filter(
      Sector == sec,
      Group %in% groups_to_plot
    ) %>% 
    select(date, Group, ret) %>% 
    tbl_xts(cols_to_xts = ret, spread_by = Group)
  
  avail_groups <- intersect(groups_to_plot, colnames(sec_xts))
  
  chart.Drawdown(
    sec_xts[, avail_groups],
    geometric  = TRUE,
    legend.loc = "bottomleft",
    colorset   = 1:length(avail_groups),
    lwd        = 4,
    main       = paste("Drawdowns by size -", sec),
    ylab       = "Drawdown from peak"
  )
}

# Individual sector drawdown charts
draw_sector_dd("Industrials")
draw_sector_dd("Resources")
draw_sector_dd("Financials")
draw_sector_dd("Property")
```

### Panel 4: All Kable Tables for Analysis

```{r tables-panel, echo = TRUE}
# Drawdown analysis tables
drawdown_results <- table.Drawdowns(sector_xts[, 1:6])
kable(drawdown_results, caption = "Drawdown Analysis by Sector and Size") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Maximum drawdown by sector-size
max_drawdowns <- maxDrawdown(sector_xts)
kable(max_drawdowns, caption = "Maximum Drawdown by Sector-Size Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Detailed drawdown analysis
detailed_drawdowns <- table.Drawdowns(sector_xts[, 1:6])
kable(detailed_drawdowns, caption = "Detailed Drawdown Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Downside risk table
downside_risk <- table.DownsideRisk(sector_xts[, 1:6])
kable(downside_risk, caption = "Downside Risk Measures by Sector-Size Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Find and sort drawdowns for specific sectors
industrials_drawdowns <- findDrawdowns(sector_xts[, "Industrials_Large_Caps"])
sorted_industrials <- sortDrawdowns(industrials_drawdowns)
#kable(sorted_industrials, caption = "Sorted Drawdowns - Industrials Large Caps") %>%
  #kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Resources sector drawdown analysis
resources_drawdowns <- findDrawdowns(sector_xts[, "Resources_Large_Caps"])
sorted_resources <- sortDrawdowns(resources_drawdowns)
#kable(sorted_resources, caption = "Sorted Drawdowns - Resources Large Caps") %>%
 # kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Alternative calculation using PerformanceAnalytics for comparison
idxxts <- 
  idx %>% 
  filter(Tickers != "Unclassified") %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Tickers)

dfplotxts <- 
  bind_rows(
    idxxts %>% tail(6)  %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "A"),
    idxxts %>% tail(12) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "B"),
    idxxts %>% tail(36) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "C"),
    idxxts %>% tail(60) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "D")
  ) %>% 
  data.frame() %>% 
  tidyr::gather(Tickers , mu, -Freq) %>% 
  mutate(Tickers = gsub("\\.", " ", Tickers))

kable(dfplotxts, caption = "Annualized Returns Comparison Using PerformanceAnalytics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Conclusion

After analyzing daily holdings and returns data for JSE-listed stocks over the past decade, segmented by sector and size group, we found evidence that smaller stocks are indeed more volatile than larger ones. Rolling 3-year annualized standard deviations consistently show that Small Caps exhibit higher return variability across all sectors, with particularly sharp spikes during periods of market stress like 2020. Peak-to-trough drawdown analysis further confirms this: Small and Mid Caps experience deeper and more frequent drawdowns than Large Caps, especially in sectors like Industrials and Property, where Mid Caps saw losses exceeding 65%. Interestingly, while Large Caps are more stable, they are not immune to prolonged drawdowns, such as the 48-month decline in Financials from 2018 to 2022. These findings highlight the importance of size and sector in understanding risk dynamics and suggest that portfolio construction should account for both cyclical sensitivity and recovery behavior across market segments.
