---
title: "Q2"
output: html_document
---
```{r load-data-Q2, echo = TRUE}
# Load readr using pacman
pacman::p_load(
  readr,
  tidyverse,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  rmsfuns,
  fmxdat,
  DescTools
)

# Load the data
Hold_Rets <- read_rds("data/Hold_Rets_Sectors.rds") 

head(Hold_Rets)
```

To answer the first part of small stocks being highly volatile compared to larger stocks I will annualise the returns for the decade as requested by the question - I will also categorise the returns with their groups as either being small_cap or large_cap


In the lecturer’s notes, volatility (risk) is explicitly defined and measured using the standard deviation of returns, and we are told that:

Standard deviation should be annualised when we compare different time periods.

The usual shortcut SD * sqrt(12) is only theoretically correct when we work with logarithmic returns, because annual log return is the sum of monthly log returns, which justifies the square root of time rule.

So, to be consistent with the notes and to get a clean, comparable volatility measure, I:

Construct value weighted indices for small caps and large caps.

Convert their monthly simple returns to log returns using log(1 + ret).

Compute a 36 month rolling standard deviation of these log monthly returns.

Annualise that rolling SD by multiplying by sqrt(12).

This gives me a rolling annualised standard deviation of returns for small caps and large caps over the past decade. Comparing these two series over time is a direct, textbook implementation of the lecturer’s definition of volatility, so it is an appropriate way to answer 2(i): if the small cap series has a consistently higher annualised SD, we can conclude that smaller stocks have been more volatile than larger stocks locally.

```{r}
options(dplyr.summarise.inform = FALSE)

library(tidyverse)
library(lubridate)
library(tbl2xts)
library(PerformanceAnalytics)
library(RcppRoll)
library(fmxdat)   # for safe_month_min, theme_fmx, fmx_cols, fmx_fills etc

# Start from your data
# Hold_Rets has: date, Tickers, Return, Weight, Group, Sector

Hold_Rets_decade <- 
  Hold_Rets %>% 
  filter(Group %in% c("Small_Caps", "Large_Caps")) %>% 
  mutate(year = year(date)) %>% 
  filter(year >= max(year) - 9) %>%   # last 10 calendar years
  select(-year)


```


```{r}
# 1a. Daily value-weighted size index returns (Small vs Large)
idx_daily <- 
  Hold_Rets_decade %>% 
  group_by(date, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(Tickers = Group) %>%      # so Tickers is "Small_Caps" / "Large_Caps"
  arrange(Tickers, date)

# 1b. Make returns monthly for illustration (Nico style)
idx <- 
  idx_daily %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Tickers, YM) %>% 
  # pick last date of the month and compute monthly return
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Tickers, date)

head(idx)

```


```{r}
#======================
# Manual Calculation:
#======================

dfplot <- 
  bind_rows(
    # 6 months – do NOT annualise (same as Nico)
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% 
      mutate(Freq = "A"),
    
    # 1 year
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 12) - 1) %>% 
      mutate(Freq = "B"),
    
    # 3 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 36) - 1) %>% 
      mutate(Freq = "C"),
    
    # 5 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 60) - 1) %>% 
      mutate(Freq = "D")
  )

#======================
# PerformanceAnalytics check
#======================

idxxts <- 
  tbl_xts(idx, cols_to_xts = ret, spread_by = Tickers)

dfplotxts <- 
  bind_rows(
    idxxts %>% tail(6)  %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "A"),
    idxxts %>% tail(12) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "B"),
    idxxts %>% tail(36) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "C"),
    idxxts %>% tail(60) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "D")
  ) %>% 
  data.frame() %>% 
  tidyr::gather(Tickers, mu, -Freq) %>% 
  mutate(Tickers = gsub("\\.", " ", Tickers))

# Facet labels
to_string <- as_labeller(c(
  `A` = "6 Months", 
  `B` = "1 Year", 
  `C` = "3 Years", 
  `D` = "5 Years"
))

# Barplot_foo (same style)
g_ann <- 
  dfplot %>% 
  ggplot() + 
  geom_bar(aes(Tickers, mu, fill = Tickers), stat = "identity") + 
  facet_wrap(~Freq, labeller = to_string, nrow = 1) + 
  labs(
    x = "",
    y = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualized terms."
  ) + 
  fmx_fills() + 
  geom_label(
    aes(Tickers, mu, label = paste0(round(mu, 4) * 100, "%")),
    size = ggpts(8),
    alpha = 0.35,
    fontface = "bold",
    nudge_y = 0.002
  ) + 
  theme_fmx(
    CustomCaption = TRUE,
    title.size = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size = ggpts(30),
    axis.size = ggpts(37),
    legend.size = ggpts(35),
    legend.pos = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

```

```{r}
gg_cum <- 
  idx %>% 
  arrange(date) %>% 
  group_by(Tickers) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Tickers)) +
  labs(
    title   = "Illustration of Cumulative Returns of Large vs Small Caps",
    subtitle = "",
    caption = "Note:\nDistortions emerge as starting dates differ."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  )

# Level plot
gg_cum

# Log cumulative plot
gg_cum_log <- 
  gg_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_log

```

```{r}
plotdf <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll <- 
  plotdf %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized Returns: Large vs Small Caps",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "Note:\nDistortions are not evident now."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

```


# This is the answer to the volatility question 
```{r}
plot_dlog <- 
  idx %>% 
  group_by(Tickers) %>% 
  # log monthly returns:
  mutate(ret = log(1 + ret)) %>% 
  # Rolling SD annualized calc now:
  mutate(
    RollSD = RcppRoll::roll_sd(ret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>% 
  filter(!is.na(RollSD))

g_sd <- 
  plot_dlog %>% 
  ggplot() +
  geom_line(aes(date, RollSD, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized SD: Large vs Small Caps",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year SD (Ann.)",
    caption = "Note:\nAnnualized SD computed on log monthly returns."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_sd, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = FALSE)
```


Second part (peak and troughs )

We need three graphs that have returns ( of whatever format on the y axis) and year on the x axis since we have already annualised it's easy graph one needs to have a legend by size group ( small caps , large , medium and other ) , the second by sector and the last by size and sector . Group has NA which aren't necessarily missing values but rather unclassified so we shall call them that to remove confusion 

start with the one for return against size groups 
