---
title: "Q4"
output: html_document
---



### Q4: Volatility and GARCH estimates for ZAR

#### 0. Libraries and data


```{r Q4-libs-data, echo = TRUE}
pacman::p_load(
  tidyverse,
  devtools,
  rugarch,
  rmgarch,
  forecast,
  tbl2xts,
  lubridate,
  PerformanceAnalytics,
  ggthemes,
  robustbase,
  kableExtra,
  RcppRoll
)

# Currency data (all quoted vs USD)
cncy       <- read_rds("data/currencies.rds")
cncy_Carry <- read_rds("data/cncy_Carry.rds")   # G10 carry index
cncy_value <- read_rds("data/cncy_value.rds")   # FX PPP value index
cncyIV     <- read_rds("data/cncyIV.rds")       # FX implied vol index
bbdxy      <- read_rds("data/bbdxy.rds")        # Bloomberg Dollar spot index

```



### 1. Cross sectional volatility: ZAR vs G10 vs others

```{r Q4-vol-panel, echo = TRUE}
# Construct daily log returns and scaled returns for all currencies
cncy_rts <- cncy %>%  
  group_by(Name) %>%  
  arrange(date, .by_group = TRUE) %>% 
  mutate(
    dlogret   = log(Price) - log(lag(Price)),
    scaledret = dlogret - mean(dlogret, na.rm = TRUE)
  ) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>% 
  mutate(Name = gsub("_Cncy", "", Name))

# Define G10 set in the cleaned Name space
g10_vec <- c(
  "Australia_Inv",  # AUD
  "Canada",         # CAD
  "EU_Inv",         # EUR
  "Japan",          # JPY
  "NZ_Inv",         # NZD
  "Norway",         # NOK
  "Sweden",         # SEK
  "UK_Inv"          # GBP
  # USD is the base here
)

# Volatility measures since 2015
vol_cncy <- cncy_rts %>% 
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%
  summarise(
    std_dev      = sd(dlogret, na.rm = TRUE),
    mad          = mean(abs(dlogret - mean(dlogret, na.rm = TRUE)), na.rm = TRUE),
    downside_dev = sqrt(mean(ifelse(dlogret < 0, dlogret^2, 0), na.rm = TRUE)),
    d_ratio      = (n() - sum(ifelse(dlogret < 0, dlogret, 0))) / 
                   (n() + sum(ifelse(dlogret > 0, dlogret, 0))),
    .groups = "drop"
  ) %>% 
  mutate(
    Group = case_when(
      Name == "SouthAfrica" ~ "ZAR",
      Name %in% g10_vec     ~ "G10",
      TRUE                  ~ "Other"
    )
  )

# Rank by standard deviation
ranked_data <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number()) %>% 
  select(rank, Name, Group, everything())

kableExtra::kable(
  ranked_data,
  caption = "Volatility ranking of currencies (log returns, since 2015)",
  digits = 6
)
```



```{r Q4-vol-plot, echo = TRUE}
# Identify top 10 by volatility
vol_with_rank <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number())

# Keep: top 10 *or* any G10 currency
plot_set <- vol_with_rank %>% 
  filter(rank <= 10 | Group == "G10")

plot_set %>% 
  mutate(Name = fct_reorder(Name, std_dev)) %>% 
  ggplot() +
  geom_col(aes(x = Name, y = std_dev, fill = Group), alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "ZAR"   = "red",
      "G10"   = "grey40",
      "Other" = "grey80"
    )
  ) +
  coord_flip() +
  labs(
    title    = "Cross sectional volatility of currencies",
    subtitle = "Top 10 currencies by volatility plus G10 majors (since 2015)",
    x = "",
    y = "Standard deviation",
    fill = ""
  ) +
  fmxdat::theme_fmx()

```

---


```{r Q4-zar-persistence, echo = TRUE}
# Simple returns for ZAR only
zar_rts <- cncy %>%  
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%  
  mutate(ret = Price / lag(Price) - 1) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>%  
  mutate(Name = gsub("_Cncy", "", Name)) %>% 
  filter(Name == "SouthAfrica") %>% 
  filter(!is.na(ret)) %>% 
  select(-Name, -Price) %>% 
  tbl_xts()

# Construct the three series: returns, squared returns, absolute returns
Plotdata <- cbind(zar_rts, zar_rts^2, abs(zar_rts))
colnames(Plotdata) <- c("Returns", "Returns_Sqd", "Returns_Abs")

Plotdata <- Plotdata %>% 
  xts_tbl() %>% 
  gather(ReturnType, Returns, -date)

ggplot(Plotdata) + 
  geom_line(aes(x = date, y = Returns, colour = ReturnType, alpha = 0.5)) + 
  ggtitle("Return Type Persistence: ZAR") + 
  facet_wrap(~ReturnType, nrow = 3, ncol = 1, scales = "free") + 
  guides(alpha = "none", colour = "none") + 
  fmxdat::theme_fmx()
```


```{r Q4-zar-acf, echo = TRUE}
forecast::Acf(zar_rts,   main = "ACF: ZAR return")
forecast::Acf(zar_rts^2, main = "ACF: Squared ZAR return")
forecast::Acf(abs(zar_rts), main = "ACF: Absolute ZAR return")

Box.test(coredata(zar_rts^2), type = "Ljung-Box", lag = 12)
```


---

### 3. Fitting GARCH(1,1) to ZAR (rugarch, copy of Practical 6)

```{r Q4-zar-garch-fit, echo = TRUE}
# Use the same object name as Nico: porteqw now holds ZAR returns
porteqw <- zar_rts

# GARCH(1,1) specification as in the notes
garch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[1],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[1]
)

garchfit1 <- ugarchfit(spec = garch11, data = porteqw)

garchfit1
```


```{r Q4-zar-garch-table, echo = TRUE}
garch_tab <- garchfit1@fit$matcoef
kableExtra::kable(garch_tab, caption = "GARCH(1,1) estimates for ZAR")
```



Conditional variance plot, identical structure:


```{r Q4-zar-sigma-plot, echo = TRUE}
sigma <- sigma(garchfit1) %>% xts_tbl() 
colnames(sigma) <- c("date", "sigma") 
sigma <- sigma %>% mutate(date = as.Date(date))

gg <- ggplot() + 
  geom_line(
    data = Plotdata %>% 
      filter(ReturnType == "Returns_Sqd") %>% 
      select(date, Returns) %>% 
      unique() %>% 
      mutate(Returns = sqrt(Returns)),
    aes(x = date, y = Returns)
  ) + 
  geom_line(
    data = sigma,
    aes(x = date, y = sigma),
    color = "red", size = 2, alpha = 0.8
  ) + 
  labs(
    title    = "Comparison: Returns sigma vs sigma from GARCH",
    subtitle = "ZAR: note the smoothing effect of GARCH as noise is controlled for.",
    x = "", y = "Estimated volatility",
    caption = "Source: Financial econometrics practical | Calculations: Own"
  ) + 
  fmxdat::theme_fmx(CustomCaption = TRUE)

fmxdat::finplot(gg, y.pct = TRUE, y.pct_acc = 1)

```


News impact curve for the ZAR GARCH model:


```{r Q4-zar-nic, echo = TRUE}
ni <- newsimpact(z = NULL, garchfit1)

plot(
  ni$zx, ni$zy,
  ylab = ni$yexpr,
  xlab = ni$xexpr,
  type = "l",
  main = "News Impact Curve: ZAR sGARCH(1,1)"
)

```



### 4. Alternative GARCH forms for ZAR (GJR and EGARCH) and NIC comparison

```{r Q4-zar-gjr-egarch, echo = TRUE}
# GJR GARCH with student t
gjrgarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[2],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit2 <- ugarchfit(spec = gjrgarch11, data = as.matrix(porteqw))

kableExtra::kable(
  garchfit2@fit$matcoef,
  caption = "GJR GARCH(1,1) with student t distribution for ZAR"
)

# EGARCH(1,1) with student t
egarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[3],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit3 <- ugarchfit(spec = egarch11, data = as.matrix(porteqw))

# Sign bias and info criteria, as in Nico’s notes
signbias(garchfit1)
infocriteria(garchfit1)
infocriteria(garchfit2)
infocriteria(garchfit3)

```



News impact curves for all three, identical helper function:


```{r Q4-zar-nic-compare, echo = TRUE}
NICurveMaker <- function(Fit, Name) {
  NI <- newsimpact(z = NULL, Fit)
  NI <- cbind(NI$zx, NI$zy)
  colnames(NI) <- c(paste0("Epsilon_", Name), paste0("Sigma_", Name))
  NI %>% data.frame() %>% tibble::as_tibble()
}

NI1 <- NICurveMaker(garchfit1, "GARCH11")
NI2 <- NICurveMaker(garchfit2, "GJR")
NI3 <- NICurveMaker(garchfit3, "EGARCH")

NI <- cbind(NI1, NI2, NI3) %>% 
  gather(Model, Sigma, starts_with("Sigma")) %>% 
  rename(Epsilon = Epsilon_GARCH11) %>% 
  select(-Epsilon_GJR, -Epsilon_EGARCH)

ggplot(NI) + 
  geom_line(aes(x = Epsilon, y = Sigma, colour = Model)) + 
  ggtitle("News Impact Curves: ZAR") + 
  theme_hc()

```


---

### 5. GO GARCH and time varying correlation between ZAR and Dollar strength

Now we bring in the multivariate flavour, matching Practical 7 style, but on a simple two series system: ZAR and the Dollar index (bbdxy). This speaks directly to the “benefits when the Dollar is strong” statement.


```{r Q4-zar-go-garch-data, echo = TRUE}
# Dollar index log returns
g10_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(G10 = log(Price) - log(lag(Price))) %>% 
  filter(date > dplyr::first(date)) %>% 
  select(date, G10)

# ZAR log returns (from earlier cncy_rts object)
zar_log_rts <- cncy_rts %>% 
  filter(Name == "SouthAfrica") %>% 
  rename(ZAR = dlogret) %>% 
  select(date, ZAR)

# Merge and convert to xts
xts_rtn <- left_join(g10_rts, zar_log_rts, by = "date") %>% 
  drop_na() %>% 
  tbl_xts()

```



Helper for renaming pairwise correlations (same as Practical 7):


```{r Q4-renamingdcc, echo = TRUE}
renamingdcc <- function(ReturnSeries, DCC.TV.Cor) {
  dates <- zoo::index(ReturnSeries)
  N     <- ncol(ReturnSeries)
  names <- colnames(ReturnSeries)
  
  pairs      <- expand.grid(names, names)
  pair_names <- paste0(pairs$Var1, "_", pairs$Var2)
  
  df <- as.data.frame(DCC.TV.Cor)
  colnames(df) <- pair_names
  df$date      <- dates
  
  df %>% 
    tidyr::pivot_longer(
      cols      = -date,
      names_to  = "Pairs",
      values_to = "Rho"
    ) %>% 
    dplyr::arrange(date)
}

```



GO GARCH specification and fit, directly from Nico’s multivariate notes:

```{r Q4-zar-go-garch-fit, echo = TRUE}
# Univariate GJR GARCH spec for each series
uspec <- ugarchspec(
  variance.model = list(
    model      = "gjrGARCH",
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = "sstd"
)

multi_univ_garch_spec <- multispec(replicate(ncol(xts_rtn), uspec))

spec.go <- gogarchspec(
  multi_univ_garch_spec,
  distribution.model = "mvnorm",
  ica                = "fastica"
)

cl <- makePSOCKcluster(10)

multf <- multifit(multi_univ_garch_spec, xts_rtn, cluster = cl)

fit.gogarch <- gogarchfit(
  spec    = spec.go,
  data    = xts_rtn,
  solver  = "hybrid",
  cluster = cl,
  gfun    = "tanh",
  maxiter1 = 40000,
  epsilon  = 1e-08,
  rseed    = 100
)

stopCluster(cl)

print(fit.gogarch)

```


Time varying correlations, same gymnastics as in the notes:


```{r Q4-zar-go-garch-tv-cor, echo = TRUE}
gog.time.var.cor <- rcor(fit.gogarch)
gog.time.var.cor <- aperm(gog.time.var.cor, c(3, 2, 1))
dim(gog.time.var.cor) <- c(
  nrow(gog.time.var.cor),
  ncol(gog.time.var.cor)^2
)

gog.time.var.cor <- renamingdcc(
  ReturnSeries = xts_rtn,
  DCC.TV.Cor   = gog.time.var.cor
)

g2 <- ggplot(
  gog.time.var.cor %>% 
    filter(grepl("ZAR_", Pairs), !grepl("_ZAR", Pairs))
) +
  geom_line(aes(x = date, y = Rho, colour = Pairs)) +
  theme_hc() +
  ggtitle("Go GARCH: time varying correlation between ZAR and Dollar index")

print(g2)

```


---

### 6. Linking to G10 carry and PPP value (for narrative)

For the second statement in the question, you can add very short extra chunks to relate ZAR returns to:

- G10 carry index `cncy_Carry`  
- PPP value index `cncy_value`  
- FX implied vol `cncyIV`  

For example, simple correlations over the sample:


```{r Q4-zar-carry-value-links, echo = TRUE}
# Simple daily log returns for ZAR, carry, value, and Dollar index
zar_daily <- cncy_rts %>% filter(Name == "SouthAfrica") %>% select(date, ZAR = dlogret)

carry_rts <- cncy_Carry %>% 
  arrange(date) %>% 
  mutate(Carry = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Carry)) %>% 
  select(date, Carry)

value_rts <- cncy_value %>% 
  arrange(date) %>% 
  mutate(Value = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Value)) %>% 
  select(date, Value)

dxy_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(DXY = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(DXY)) %>% 
  select(date, DXY)

link_df <- zar_daily %>% 
  inner_join(carry_rts, by = "date") %>% 
  inner_join(value_rts, by = "date") %>% 
  inner_join(dxy_rts,   by = "date")

round(cor(link_df %>% select(-date), use = "complete.obs"), 3)

```

