---
title: "Q5"
output: html_document
---


# QUESTION 5: Global Balanced Index Fund Portfolio Construction

We want to construct a global balanced index fund portfolio using a mix of traded global indexes.


## Load all required packages at the top

```{r}
# Load all required packages
pacman::p_load(
  tidyverse,      
  xts,            
  tbl2xts,        
  RiskPortfolios, 
  PerformanceAnalytics,
  PortfolioAnalytics,   
  lubridate,      
  kableExtra,   
  quadprog,       # For quadratic optimization
  cluster,        # For hierarchical clustering
  readr,          
  dplyr         
)


devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")
```


## DATA LOADING AND PREPARATION

```{r}

MAA_data <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI equity indices
msci_data <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine all datasets 
df_full <- bind_rows(
  msci_data %>% rename(Tickers = Name),
  MAA_data %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)  

# Create monthly returns from 2011
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       
  group_by(Tickers) %>%
  # Keep only assets with at least 5 years of data 
  filter(n_distinct(year(date)) >= 5) %>% 
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  filter(date == dplyr::last(date)) %>%
  # Calculate monthly returns
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   

# Convert to wide format matrix for portfolio optimization
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

## PORTFOLIO OPTIMIZER FUNCTION
```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,    # 10 years of monthly data
                      LB        = 0.01,         # Minimum 1% per asset
                      UB        = 0.35,         # Maximum 35% per single asset
                      bond_UB   = 0.25,         # Bonds max 25% of portfolio
                      eq_UB     = 0.60,         # Equities max 60%
                      com_UB    = 0.15) {       # Commodities max 15%
  
  # 1. SELECT LOOKBACK WINDOW
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)  # Last 'lookback_months' observations
  
  # Safety check - ensure we have enough data
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  #  Remove assets with insufficient history
  X <- data.matrix(window_df[, -1])  
  
  # Keep only assets with at least 60 non-NA observations (5 years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. ESTIMATE COVARIANCE AND RETURNS
  # Using Ledoit-Wolf shrinkage for better covariance estimation
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  # Calculate geometric mean returns (more appropriate for compounding)
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)  # Number of valid assets
  
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  # Create indices for each asset class for constraint application
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. CONSTRAINTS MATRIX
  # Base constraints: sum(weights)=1, lower bounds, upper bounds
  Amat <- cbind(
    rep(1, N),         # Sum of weights = 1 (fully invested)
    diag(N),           # Lower bounds: w_i >= LB
    -diag(N)           # Upper bounds: w_i <= UB (via -w_i >= -UB)
  )
  
  bvec <- c(
    1,                 # Fully invested constraint
    rep(LB, N),        # Lower bounds
    -rep(UB, N)        # Upper bounds (negative due to inequality direction)
  )
  
  #function to add group constraints
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # Initialize constraint vector
    g_vec[idx] <- -1              # -sum(w_group) >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  

  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # First constraint (sum to 1) is equality
  
  # 6. SOLVE QUADRATIC PROGRAM
  #  min (1/2) w' Sigma w - mu' w
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,    # Covariance matrix
    dvec = Mu,       # Expected returns  
    Amat = Amat,     # Constraints matrix
    bvec = bvec,     # Constraints vector
    meq  = meq       # Number of equality constraints
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

## 3. REBALANCING 

```{r}
# Get quarter-end dates (Jan, Apr, Jul, Oct) for rebalancing (Why ?? )
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%  # Quarter-end months
  pull(date)
```

## ROLLING OPTIMIZATION 
```{r}
# Wrapper function for rolling optimization
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 10-year lookback
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Execute rolling optimization across all quarter-end dates
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

# Display 
kable(Result_roll %>% head(13))
```

## 5. PORTFOLIO WEIGHTS VISUALIZATION
```{r}
# Prepare data for stacked bar chart
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%                    # Convert to xts format
  .[endpoints(., "months")]        # Keep only month-end points

# Create stacked bar chart of portfolio weights over time
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund",
  ylab = "Portfolio Weight",
  xlab = "Date"
)
```

## 6. CLUSTERING ANALYSIS FOR COMOVEMENT (prac 5 )

```{r}
# CLUSTERING BASED COMOVEMENT ANALYSIS
# Remove date column for clustering
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Estimate shrunk correlation matrix (Ledoit-Wolf)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")  # Ledoit-Wolf shrinkage
)

# Convert to correlation matrix
corr_clust <- cov2cor(Sigma_clust)


# Distance = sqrt((1 - correlation)/2)
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward's method
hc <- cluster::agnes(dist(distmat), method = "ward")

# Plot dendrogram to visualize clusters
plot(hc, 
     which.plot = 2, 
     main = "Hierarchical Clustering of Global Indices",
     xlab = "Indices", 
     sub = "",
     cex = 0.7)  # Smaller text for better fit
```

