---
title: "Financial Econometrics 871 Practical Exam"
output: github_document
editor_options: 
  chunk_output_type: console
---


**SU Number:** 22660348  
**Author:** Precious Nhamo  
**Course:** Financial Econometrics 871  
**Dates:** 18–19 November 2025  

Repo: <https://github.com/prexymtha/FMX_25_22660348>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo   = TRUE,   # show code
  message = FALSE,
  warning = FALSE
)
```





### Question 1 

I struggled with this question....

We can compute the rolling 3-year beta for each active manager and the ABC Fund against the benchmark.Then, we can create a scatterplot of returns vs. risk (beta) and overlay boxplots to show the distribution of betas for peers and highlight ABC Fund.


To see how much the fund differs from the rest of the industry in terms of what drives the return .We can perform a factor analysis or compare the R-squared from a market model to see how much of the returns are explained by the market.


Assess relative performance over time We can plot the cumulative returns of ABC Fund, peers, and the benchmark.

We can compute a measure of dispersion (cross-sectional standard deviation of returns) and then plot the fund's performance against this dispersion.

```{r load-data-Q1, echo = TRUE}
# Load readr for rds data sets 
pacman::p_load(readr, fmxdat)



# Load the data
Benchmark <- read_rds("data/Capped_SWIX.rds")
Active_Managers <- read_rds("data/Active_Managers.rds")
ABC_Fund <- read_rds("data/ABC_Fund.rds")

```

Since we have a benchmark , ABC fund industry peer we can't analyse all individual industry peers - we will calculate the mean or average of the peers 

```{r}
# Load required libraries with pacman
pacman::p_load(
  dplyr,
  ggplot2,
  lubridate,
  tidyr,
  roll,
  zoo
)

# 1. Set common sample based on ABC Fund dates
start_date <- as.Date("2012-07-31")
end_date   <- as.Date("2025-10-31")

# 2. Filter each data set to the common period
benchmark_common <- 
  Benchmark %>% 
  filter(date >= start_date, date <= end_date)

active_managers_common <- 
  Active_Managers %>% 
  filter(date >= start_date, date <= end_date)

abc_data <- 
  ABC_Fund %>% 
  filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# 3. Prepare benchmark data
bench_data <- 
  benchmark_common %>% 
  select(date, Benchmark_Returns = Returns)

# 4. Prepare peer summary by date
peers_data <- 
  active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns    = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count          = n(),
    .groups              = "drop"
  )

# 5. Merge into a single analysis data set
analysis_data <- 
  abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data,  by = "date")

# Optional quick check
dplyr::glimpse(analysis_data)


```

We will be using analysis data to continue with our questions 



```{r}
pacman::p_load(
  dplyr,
  tbl2xts,
  PerformanceAnalytics
)

# Ra: assets we are analysing (ABC + peer average)
Ra_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>% 
  mutate(across(-date, ~ coalesce(., 0))) %>% 
  tbl_xts(
    cols_to_xts = c("ABC_Fund", "Peer_Average")
  )

# Rb: benchmark we compare against
Rb_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    Benchmark = Benchmark_Returns
  ) %>% 
  mutate(Benchmark = coalesce(Benchmark, 0)) %>% 
  tbl_xts(
    cols_to_xts = "Benchmark"
  )

```

```{r}
# Rolling 3 year betas (36 months) vs benchmark
PerformanceAnalytics::chart.RollingRegression(
  Ra         = Ra_xts,
  Rb         = Rb_xts,
  width      = 36,
  attribute  = "Beta",
  legend.loc = "topleft",
  main       = "Rolling 3 year beta vs benchmark: ABC vs peer average"
)

# Full sample betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = Ra_xts,
  Rb = Rb_xts
)


```

Box-Scatter Plot ( I struggled with this part)

```{r}
pacman::p_load(
  dplyr,
  tidyr,
  ggplot2,
  roll,
  fmxdat
)

# 1. Compute rolling 3 year betas (36 months) vs benchmark

# make sure data is ordered
beta_data <- 
  analysis_data %>% 
  arrange(date)

# rolling regression: y on x with width = 36
# ABC vs Benchmark
roll_abc <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$ABC_Returns,
    width = 36
  )

# Peer average vs Benchmark
roll_peer <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$Peers_Avg_Returns,
    width = 36
  )

# extract slope (beta) from coefficients[, 2]
rolling_betas <- 
  tibble(
    date        = beta_data$date,
    ABC_Fund    = as.numeric(roll_abc$coefficients[, 2]),
    Peer_Average = as.numeric(roll_peer$coefficients[, 2])
  ) %>% 
  pivot_longer(
    cols      = c(ABC_Fund, Peer_Average),
    names_to  = "Series",
    values_to = "Beta"
  ) %>% 
  filter(!is.na(Beta))   # drop early windows with no full 36 months

# 2. Boxplot with scatter (jitter) on top

ggplot(rolling_betas, aes(x = Series, y = Beta)) +
  geom_jitter(width = 0.1, alpha = 0.4) +   # scatter points
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +  # box on top
  labs(
    title = "Distribution of rolling 3 year betas vs benchmark",
    x     = NULL,
    y     = "Rolling 3 year beta"
  ) +
  theme_fmx()

```






### Question 2

To answer the first part of Question 2 on whether smaller stocks are more volatile than larger stocks, I focus on the last decade(2015-2024) of daily returns and compare volatility across size groups(Small , Mid , Large Caps). I work with the `Hold_Rets_Sectors` data set, which contains daily returns, holdings weights, size group (`Group`) and sector labels for JSE stocks.

In the lecturer's notes, volatility (risk) is defined and measured using the standard deviation of returns, and it is emphasised that:

* Standard deviation should be annualised when comparing different time periods.
* The usual shortcut ( \text{SD} \times \sqrt{12} ) is only theoretically correct when using **logarithmic** returns, because annual log return is the sum of monthly log returns.

I proceed as follows:

1. I construct value weighted indices for each size group (Small Caps, Mid Caps and Large Caps).
2. I convert their monthly simple returns to log returns using (\log(1 + r)).
3. I compute a 36-month rolling standard deviation of these monthly log returns.
4. I annualise this rolling standard deviation by multiplying by (\sqrt{12}).

This yields a rolling annualised standard deviation of returns for each size group .If the Small Cap series has a consistently higher annualised standard deviation, I can conclude that smaller stocks have been more volatile than larger stocks locally.

### Load libraries and raw data

```{r load-data-Q2, echo = TRUE, message = FALSE, warning = FALSE}
# Load required packages
pacman::p_load(
  readr,
  tidyverse,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  rmsfuns,
  fmxdat,
  DescTools,
  RcppRoll
)

# Load the holdings and returns data with sector and size group labels
Hold_Rets <- read_rds("data/Hold_Rets_Sectors.rds") %>%
  mutate(
    # Make the size and sector groups explicit instead of leaving them as NA
    Group  = if_else(is.na(Group),  "Unclassified", Group),
    Sector = if_else(is.na(Sector), "Unclassified", Sector)
  )

head(Hold_Rets)
```

Here I explicitly rename `NA` groups to `"Unclassified"` so they do not silently disappear from the data. Later, when I compare volatility across size groups, I restrict the analysis to the proper size categories (Small_Caps, Mid_Caps, Large_Caps) and treat "Unclassified" separately if needed.

### Data Preparation and Index Construction

```{r prep-and-index-Q2, echo = TRUE}
options(dplyr.summarise.inform = FALSE)

# Filter data for the past decade and main size groups
Hold_Rets_decade <- 
  Hold_Rets %>% 
  filter(Group %in% c("Small_Caps", "Mid_Caps", "Large_Caps","Unclassified")) %>% 
  mutate(year = year(date))

# Construct daily value-weighted size index returns
idx_daily <- 
  Hold_Rets_decade %>% 
  group_by(date, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(Tickers = Group) %>%      # so Tickers is "all groups"
  arrange(Tickers, date)

# Convert to monthly returns
idx <- 
  idx_daily %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Tickers, YM) %>% 
  # pick last date of the month and compute monthly return
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Tickers, date)

head(idx)
```

###Performance Analysis - Returns

```{r performance-analysis-Q2, echo = TRUE}
# Prepare data for multi-horizon performance measures
dfplot <- 
  bind_rows(
    # 6 months - do NOT annualise 
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% 
      mutate(Freq = "A"),
    
    # 1 year
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 12) - 1) %>% 
      mutate(Freq = "B"),
    
    # 3 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 36) - 1) %>% 
      mutate(Freq = "C"),
    
    # 5 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 60) - 1) %>% 
      mutate(Freq = "D")
  )

# Visualization functions and plots
to_string <- as_labeller(c(
  `A` = "6 Months", 
  `B` = "1 Year", 
  `C` = "3 Years", 
  `D` = "5 Years"
))

# Annualized returns bar plot
g_ann <- 
  dfplot %>% 
  ggplot() + 
  geom_bar(aes(Tickers, mu, fill = Tickers), stat = "identity") + 
  facet_wrap(~Freq, labeller = to_string, nrow = 1) + 
  labs(
    x = "",
    y = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualized terms."
  ) + 
  fmx_fills() + 
  geom_label(
    aes(Tickers, mu, label = paste0(round(mu, 4) * 100, "%")),
    size = ggpts(8),
    alpha = 0.35,
    fontface = "bold",
    nudge_y = 0.002
  ) + 
  theme_fmx(
    CustomCaption = TRUE,
    title.size = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size = ggpts(30),
    axis.size = ggpts(37),
    legend.size = ggpts(35),
    legend.pos = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

# Cumulative returns plots
gg_cum <- 
  idx %>% 
  arrange(date) %>% 
  group_by(Tickers) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Tickers)) +
  scale_colour_manual(
    values = c(
      "Large_Caps"   = "#1f77b4",
      "Mid_Caps"     = "#2ca02c",
      "Small_Caps"   = "#d62728",
      "Unclassified" = "grey40"
    )) +
  labs(
    title    = "Cumulative Returns by Cap Size Group",
    subtitle = "Growth of 1 rand invested in each group (rebased to 1 at first observation)",
    caption  = "Note: Each size cap is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  )

gg_cum

# Log cumulative plot
gg_cum_log <- 
  gg_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_log
```

### Volatility Analysis

```{r volatility-analysis-Q2, echo = TRUE}
# Rolling returns analysis
plotdf <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll <- 
  plotdf %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Volatility analysis - answer to the volatility question
plot_dlog <- 
  idx %>% 
  group_by(Tickers) %>% 
  # 1. Convert simple monthly returns to log returns
  mutate(ret = log(1 + ret)) %>% 
  # 2. Rolling 36-month SD of log returns, annualised
  mutate(
    RollSD = RcppRoll::roll_sd(ret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(RollSD)) %>% 
  # 3. Flag the "focus" buckets and relevel for nicer colours
  mutate(
    SizeBucket = if_else(Tickers %in% c("Small_Caps", "Large_Caps"),
                         "focus", "other"),
    Tickers    = forcats::fct_relevel(Tickers, "Small_Caps", "Large_Caps")
  )

g_sd <- 
  plot_dlog %>% 
  ggplot() +
  geom_line(
    aes(date, RollSD, colour = Tickers, alpha = SizeBucket, size = SizeBucket)
  ) +
  # Make Small/Large bold, others dim
  scale_alpha_manual(values = c(focus = 1,   other = 0.25), guide = "none") +
  scale_size_manual( values = c(focus = 1.8, other = 0.7),   guide = "none") +
  labs(
    title   = "Rolling 3 Year Annualized SD by Size Bucket",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year SD (Ann.)",
    caption = "Note:\nAnnualized SD computed on log monthly returns."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_sd, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = FALSE)
```

### Sector-Level Analysis

```{r sector-analysis-Q2, echo = TRUE}
# Monthly value-weighted returns by Sector and Group (size)
idx_sector_size <- 
  Hold_Rets_decade %>% 
  group_by(date, Sector, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Sector, Group, YM) %>% 
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Sector, Group, date)

# Rolling returns analysis by sector and size
plotdf_sector <- 
  idx_sector_size %>% 
  group_by(Sector, Group) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll_sector <- 
  plotdf_sector %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Group), alpha = 0.7, size = 1.25) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll_sector, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Sector-level volatility analysis
plot_sd_sector <- 
  Hold_Rets_decade %>%
  # value-weighted daily returns per Sector × Size Group
  group_by(date, Sector, Group) %>%
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(YM = format(date, "%Y%B")) %>%
  arrange(Sector, Group, date) %>%
  # collapse to month-end and compute monthly returns
  group_by(Sector, Group, YM) %>%
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>%
  arrange(Sector, Group, date) %>%
  group_by(Sector, Group) %>%
  # log monthly returns and rolling 36-month SD, annualised
  mutate(
    ret_log = log1p(ret),
    RollSD  = RcppRoll::roll_sd(ret_log, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  ungroup() %>%
  filter(!is.na(RollSD))

g_sd_sector <- 
  plot_sd_sector %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Group), size = 1.3, alpha = 0.9) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualised SD by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3-year SD (Ann.)"
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5)
  ) +
  fmx_cols()

finplot(
  g_sd_sector,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)

# Cumulative returns by sector and size
idx_sector_size_cum <- 
  idx_sector_size %>% 
  arrange(Sector, Group, date) %>% 
  group_by(Sector, Group) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup()

gg_cum_sector <- 
  idx_sector_size_cum %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Group)) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title    = "Cumulative Returns by Sector and Size Group",
    subtitle = "Growth of 1 rand invested in each sector-size group (rebased to 1 at first observation)",
    caption  = "Note: Each sector-size group is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

gg_cum_sector

# Log cumulative plot by sector
gg_cum_sector_log <- 
  gg_cum_sector +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum_sector$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_sector_log
```

### Drawdown Analysis

```{r drawdown-analysis-Q2, echo = TRUE}
# Convert to xts for PerformanceAnalytics
idxxts <- 
  idx %>% 
  filter(Tickers != "Unclassified") %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Tickers)

# Main drawdown chart
chart.Drawdown(
  idxxts[, c("Large_Caps", "Mid_Caps", "Small_Caps")],
  geometric  = TRUE,
  legend.loc = "bottomleft",
  colorset   = 1:3,
  lwd        = 3,
  main       = "Peak to trough drawdowns by size group",
  ylab       = "Drawdown from peak"
)

# Sector drawdowns
draw_sector_dd <- function(sec) {
  
  groups_to_plot <- c("Large_Caps", "Mid_Caps", "Small_Caps")
  
  sec_xts <- 
    idx_sector_size %>% 
    filter(
      Sector == sec,
      Group %in% groups_to_plot    # drop Unclassified
    ) %>% 
    select(date, Group, ret) %>% 
    tbl_xts(cols_to_xts = ret, spread_by = Group)
  
  # Only use the groups that actually exist for this sector
  avail_groups <- intersect(groups_to_plot, colnames(sec_xts))
  
  chart.Drawdown(
    sec_xts[, avail_groups],
    geometric  = TRUE,
    legend.loc = "bottomleft",
    colorset   = 1:length(avail_groups),
    lwd        = 4,
    main       = paste("Drawdowns by size -", sec),
    ylab       = "Drawdown from peak"
  )
}

# Individual sector drawdown charts
draw_sector_dd("Industrials")
draw_sector_dd("Resources")
draw_sector_dd("Financials")
draw_sector_dd("Property")
```

### Statistical Analysis Tables 

(Added extra from data-camp maybe unnecessary)

```{r statistical-tables-Q2, echo = TRUE}
# Time series chart of sector returns using PerformanceAnalytics
sector_xts <- 
  idx_sector_size %>% 
  filter(Group %in% c("Large_Caps", "Mid_Caps", "Small_Caps")) %>% 
  select(date, Sector, Group, ret) %>% 
  mutate(Sector_Group = paste(Sector, Group, sep = "_")) %>% 
  select(-Sector, -Group) %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Sector_Group)

# Chart time series of selected sector-size combinations
chart.TimeSeries(
  sector_xts[, c("Industrials_Large_Caps", "Resources_Large_Caps", "Financials_Large_Caps")],
  main = "Large Cap Returns by Sector",
  legend.loc = "topright",
  colorset = 1:3,
  lwd = 2,
  ylab = "Monthly Returns"
)

# Drawdown analysis tables
drawdown_results <- table.Drawdowns(sector_xts[, 1:6])
drawdown_results

# Maximum drawdown by sector-size
max_drawdowns <- maxDrawdown(sector_xts)
max_drawdowns

# Detailed drawdown analysis
detailed_drawdowns <- table.Drawdowns(sector_xts[, 1:6])
detailed_drawdowns

# Downside risk table
downside_risk <- table.DownsideRisk(sector_xts[, 1:6])
downside_risk

# Find and sort drawdowns for specific sectors
industrials_drawdowns <- findDrawdowns(sector_xts[, "Industrials_Large_Caps"])
sorted_industrials <- sortDrawdowns(industrials_drawdowns)
sorted_industrials

# Resources sector drawdown analysis
resources_drawdowns <- findDrawdowns(sector_xts[, "Resources_Large_Caps"])
sorted_resources <- sortDrawdowns(resources_drawdowns)
sorted_resources
```

### Conclusion


Using a  Rolling 3-year annualized standard deviations consistently show that Small Caps exhibit higher return variability across all sectors, with particularly sharp spikes during periods of market stress like 2020. Peak-to-trough drawdown analysis further confirms this: Small and Mid Caps experience deeper and more frequent drawdowns than Large Caps, especially in sectors like Industrials and Property, where Mid Caps saw losses exceeding 65%. Interestingly, while Large Caps are more stable, they are not immune to prolonged drawdowns, such as the 48-month decline in Financials from 2018 to 2022. These findings highlight the importance of size and sector in understanding risk dynamics and suggest that portfolio construction should account for both cyclical sensitivity and recovery behavior across market segments.

Seeing that we have individual stocks ( maybe we could have assessed the most volatility and which sectors they are found by their sizes but I was unsure how to do that , plus I felt my solution was way all over the place trying to include all code so I excluded it)














### Question 3 

### Question 4: Volatility and GARCH estimates

#### 5.4.1 Aim of the question

- We are assessing whether the South African Rand (ZAR) has been one of the most volatile currency over the past few years.(since it says one of the most we assess it relative to other currencies)
- ZAR has generally performed well during periods where the G10 currency carry trades have been favourable and currency valuations cheap ( we can split this for two and then do the joint )
- Globally, it has been one of the currencies that most benefit during periods where the Dollar is comparatively strong,
 indicating a risk-on sentiment
- Use your discretion on what you use to measure volatility, which currencies you compare
 and how you arrive at your conclusions. Be creative in using tables, graphs, stratification and
 statistics to argue your points.


- Use your discretion on what you use to measure volatility, which currencies you compare
 and how you arrive at your conclusions. Be creative in using tables, graphs, stratification and
 statistics to argue your points.








### Q4: Volatility and GARCH estimates for ZAR

#### 0. Libraries and data


```{r Q4-libs-data, echo = TRUE}
pacman::p_load(
  tidyverse,
  devtools,
  rugarch,
  rmgarch,
  forecast,
  tbl2xts,
  lubridate,
  PerformanceAnalytics,
  ggthemes,
  robustbase,
  kableExtra,
  RcppRoll
)

# Currency data (all quoted vs USD)
cncy       <- read_rds("data/currencies.rds")
cncy_Carry <- read_rds("data/cncy_Carry.rds")   # G10 carry index
cncy_value <- read_rds("data/cncy_value.rds")   # FX PPP value index
cncyIV     <- read_rds("data/cncyIV.rds")       # FX implied vol index
bbdxy      <- read_rds("data/bbdxy.rds")        # Bloomberg Dollar spot index

```



### 1. Cross sectional volatility: ZAR vs G10 vs others

```{r Q4-vol-panel, echo = TRUE}
# Construct daily log returns and scaled returns for all currencies
cncy_rts <- cncy %>%  
  group_by(Name) %>%  
  arrange(date, .by_group = TRUE) %>% 
  mutate(
    dlogret   = log(Price) - log(lag(Price)),
    scaledret = dlogret - mean(dlogret, na.rm = TRUE)
  ) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>% 
  mutate(Name = gsub("_Cncy", "", Name))

# Define G10 set in the cleaned Name space
g10_vec <- c(
  "Australia_Inv",  # AUD
  "Canada",         # CAD
  "EU_Inv",         # EUR
  "Japan",          # JPY
  "NZ_Inv",         # NZD
  "Norway",         # NOK
  "Sweden",         # SEK
  "UK_Inv"          # GBP
  # USD is the base here
)

# Volatility measures since 2015
vol_cncy <- cncy_rts %>% 
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%
  summarise(
    std_dev      = sd(dlogret, na.rm = TRUE),
    mad          = mean(abs(dlogret - mean(dlogret, na.rm = TRUE)), na.rm = TRUE),
    downside_dev = sqrt(mean(ifelse(dlogret < 0, dlogret^2, 0), na.rm = TRUE)),
    d_ratio      = (n() - sum(ifelse(dlogret < 0, dlogret, 0))) / 
                   (n() + sum(ifelse(dlogret > 0, dlogret, 0))),
    .groups = "drop"
  ) %>% 
  mutate(
    Group = case_when(
      Name == "SouthAfrica" ~ "ZAR",
      Name %in% g10_vec     ~ "G10",
      TRUE                  ~ "Other"
    )
  )

# Rank by standard deviation
ranked_data <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number()) %>% 
  select(rank, Name, Group, everything())

kableExtra::kable(
  ranked_data,
  caption = "Volatility ranking of currencies (log returns, since 2015)",
  digits = 6
)
```



```{r Q4-vol-plot, echo = TRUE}
# Identify top 10 by volatility
vol_with_rank <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number())

# Keep: top 10 *or* any G10 currency
plot_set <- vol_with_rank %>% 
  filter(rank <= 10 | Group == "G10")

plot_set %>% 
  mutate(Name = fct_reorder(Name, std_dev)) %>% 
  ggplot() +
  geom_col(aes(x = Name, y = std_dev, fill = Group), alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "ZAR"   = "red",
      "G10"   = "grey40",
      "Other" = "grey80"
    )
  ) +
  coord_flip() +
  labs(
    title    = "Cross sectional volatility of currencies",
    subtitle = "Top 10 currencies by volatility plus G10 majors (since 2015)",
    x = "",
    y = "Standard deviation",
    fill = ""
  ) +
  fmxdat::theme_fmx()

```

---


```{r Q4-zar-persistence, echo = TRUE}
# Simple returns for ZAR only
zar_rts <- cncy %>%  
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%  
  mutate(ret = Price / lag(Price) - 1) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>%  
  mutate(Name = gsub("_Cncy", "", Name)) %>% 
  filter(Name == "SouthAfrica") %>% 
  filter(!is.na(ret)) %>% 
  select(-Name, -Price) %>% 
  tbl_xts()

# Construct the three series: returns, squared returns, absolute returns
Plotdata <- cbind(zar_rts, zar_rts^2, abs(zar_rts))
colnames(Plotdata) <- c("Returns", "Returns_Sqd", "Returns_Abs")

Plotdata <- Plotdata %>% 
  xts_tbl() %>% 
  gather(ReturnType, Returns, -date)

ggplot(Plotdata) + 
  geom_line(aes(x = date, y = Returns, colour = ReturnType, alpha = 0.5)) + 
  ggtitle("Return Type Persistence: ZAR") + 
  facet_wrap(~ReturnType, nrow = 3, ncol = 1, scales = "free") + 
  guides(alpha = "none", colour = "none") + 
  fmxdat::theme_fmx()
```


```{r Q4-zar-acf, echo = TRUE}
forecast::Acf(zar_rts,   main = "ACF: ZAR return")
forecast::Acf(zar_rts^2, main = "ACF: Squared ZAR return")
forecast::Acf(abs(zar_rts), main = "ACF: Absolute ZAR return")

Box.test(coredata(zar_rts^2), type = "Ljung-Box", lag = 12)
```


---

### 3. Fitting GARCH(1,1) to ZAR (rugarch, copy of Practical 6)

```{r Q4-zar-garch-fit, echo = TRUE}
# Use the same object name as Nico: porteqw now holds ZAR returns
porteqw <- zar_rts

# GARCH(1,1) specification as in the notes
garch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[1],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[1]
)

garchfit1 <- ugarchfit(spec = garch11, data = porteqw)

garchfit1
```


```{r Q4-zar-garch-table, echo = TRUE}
garch_tab <- garchfit1@fit$matcoef
kableExtra::kable(garch_tab, caption = "GARCH(1,1) estimates for ZAR")
```



Conditional variance plot, identical structure:


```{r Q4-zar-sigma-plot, echo = TRUE}
sigma <- sigma(garchfit1) %>% xts_tbl() 
colnames(sigma) <- c("date", "sigma") 
sigma <- sigma %>% mutate(date = as.Date(date))

gg <- ggplot() + 
  geom_line(
    data = Plotdata %>% 
      filter(ReturnType == "Returns_Sqd") %>% 
      select(date, Returns) %>% 
      unique() %>% 
      mutate(Returns = sqrt(Returns)),
    aes(x = date, y = Returns)
  ) + 
  geom_line(
    data = sigma,
    aes(x = date, y = sigma),
    color = "red", size = 2, alpha = 0.8
  ) + 
  labs(
    title    = "Comparison: Returns sigma vs sigma from GARCH",
    subtitle = "ZAR: note the smoothing effect of GARCH as noise is controlled for.",
    x = "", y = "Estimated volatility",
    caption = "Source: Financial econometrics practical | Calculations: Own"
  ) + 
  fmxdat::theme_fmx(CustomCaption = TRUE)

fmxdat::finplot(gg, y.pct = TRUE, y.pct_acc = 1)

```


News impact curve for the ZAR GARCH model:


```{r Q4-zar-nic, echo = TRUE}
ni <- newsimpact(z = NULL, garchfit1)

plot(
  ni$zx, ni$zy,
  ylab = ni$yexpr,
  xlab = ni$xexpr,
  type = "l",
  main = "News Impact Curve: ZAR sGARCH(1,1)"
)

```



### 4. Alternative GARCH forms for ZAR (GJR and EGARCH) and NIC comparison

```{r Q4-zar-gjr-egarch, echo = TRUE}
# GJR GARCH with student t
gjrgarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[2],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit2 <- ugarchfit(spec = gjrgarch11, data = as.matrix(porteqw))

kableExtra::kable(
  garchfit2@fit$matcoef,
  caption = "GJR GARCH(1,1) with student t distribution for ZAR"
)

# EGARCH(1,1) with student t
egarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[3],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit3 <- ugarchfit(spec = egarch11, data = as.matrix(porteqw))

# Sign bias and info criteria, as in Nico’s notes
signbias(garchfit1)
infocriteria(garchfit1)
infocriteria(garchfit2)
infocriteria(garchfit3)

```



News impact curves for all three, identical helper function:


```{r Q4-zar-nic-compare, echo = TRUE}
NICurveMaker <- function(Fit, Name) {
  NI <- newsimpact(z = NULL, Fit)
  NI <- cbind(NI$zx, NI$zy)
  colnames(NI) <- c(paste0("Epsilon_", Name), paste0("Sigma_", Name))
  NI %>% data.frame() %>% tibble::as_tibble()
}

NI1 <- NICurveMaker(garchfit1, "GARCH11")
NI2 <- NICurveMaker(garchfit2, "GJR")
NI3 <- NICurveMaker(garchfit3, "EGARCH")

NI <- cbind(NI1, NI2, NI3) %>% 
  gather(Model, Sigma, starts_with("Sigma")) %>% 
  rename(Epsilon = Epsilon_GARCH11) %>% 
  select(-Epsilon_GJR, -Epsilon_EGARCH)

ggplot(NI) + 
  geom_line(aes(x = Epsilon, y = Sigma, colour = Model)) + 
  ggtitle("News Impact Curves: ZAR") + 
  theme_hc()

```


---

### 5. GO GARCH and time varying correlation between ZAR and Dollar strength

Now we bring in the multivariate flavour, matching Practical 7 style, but on a simple two series system: ZAR and the Dollar index (bbdxy). This speaks directly to the “benefits when the Dollar is strong” statement.


```{r Q4-zar-go-garch-data, echo = TRUE}
# Dollar index log returns
g10_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(G10 = log(Price) - log(lag(Price))) %>% 
  filter(date > dplyr::first(date)) %>% 
  select(date, G10)

# ZAR log returns (from earlier cncy_rts object)
zar_log_rts <- cncy_rts %>% 
  filter(Name == "SouthAfrica") %>% 
  rename(ZAR = dlogret) %>% 
  select(date, ZAR)

# Merge and convert to xts
xts_rtn <- left_join(g10_rts, zar_log_rts, by = "date") %>% 
  drop_na() %>% 
  tbl_xts()

```



Helper for renaming pairwise correlations (same as Practical 7):


```{r Q4-renamingdcc, echo = TRUE}
renamingdcc <- function(ReturnSeries, DCC.TV.Cor) {
  dates <- zoo::index(ReturnSeries)
  N     <- ncol(ReturnSeries)
  names <- colnames(ReturnSeries)
  
  pairs      <- expand.grid(names, names)
  pair_names <- paste0(pairs$Var1, "_", pairs$Var2)
  
  df <- as.data.frame(DCC.TV.Cor)
  colnames(df) <- pair_names
  df$date      <- dates
  
  df %>% 
    tidyr::pivot_longer(
      cols      = -date,
      names_to  = "Pairs",
      values_to = "Rho"
    ) %>% 
    dplyr::arrange(date)
}

```



GO GARCH specification and fit, directly from Nico’s multivariate notes:

```{r Q4-zar-go-garch-fit, echo = TRUE}
# Univariate GJR GARCH spec for each series
uspec <- ugarchspec(
  variance.model = list(
    model      = "gjrGARCH",
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = "sstd"
)

multi_univ_garch_spec <- multispec(replicate(ncol(xts_rtn), uspec))

spec.go <- gogarchspec(
  multi_univ_garch_spec,
  distribution.model = "mvnorm",
  ica                = "fastica"
)

cl <- makePSOCKcluster(10)

multf <- multifit(multi_univ_garch_spec, xts_rtn, cluster = cl)

fit.gogarch <- gogarchfit(
  spec    = spec.go,
  data    = xts_rtn,
  solver  = "hybrid",
  cluster = cl,
  gfun    = "tanh",
  maxiter1 = 40000,
  epsilon  = 1e-08,
  rseed    = 100
)

stopCluster(cl)

print(fit.gogarch)

```


Time varying correlations, same gymnastics as in the notes:


```{r Q4-zar-go-garch-tv-cor, echo = TRUE}
gog.time.var.cor <- rcor(fit.gogarch)
gog.time.var.cor <- aperm(gog.time.var.cor, c(3, 2, 1))
dim(gog.time.var.cor) <- c(
  nrow(gog.time.var.cor),
  ncol(gog.time.var.cor)^2
)

gog.time.var.cor <- renamingdcc(
  ReturnSeries = xts_rtn,
  DCC.TV.Cor   = gog.time.var.cor
)

g2 <- ggplot(
  gog.time.var.cor %>% 
    filter(grepl("ZAR_", Pairs), !grepl("_ZAR", Pairs))
) +
  geom_line(aes(x = date, y = Rho, colour = Pairs)) +
  theme_hc() +
  ggtitle("Go GARCH: time varying correlation between ZAR and Dollar index")

print(g2)

```


---

### 6. Linking to G10 carry and PPP value (for narrative)

For the second statement in the question, you can add very short extra chunks to relate ZAR returns to:

- G10 carry index `cncy_Carry`  
- PPP value index `cncy_value`  
- FX implied vol `cncyIV`  

For example, simple correlations over the sample:


```{r Q4-zar-carry-value-links, echo = TRUE}
# Simple daily log returns for ZAR, carry, value, and Dollar index
zar_daily <- cncy_rts %>% filter(Name == "SouthAfrica") %>% select(date, ZAR = dlogret)

carry_rts <- cncy_Carry %>% 
  arrange(date) %>% 
  mutate(Carry = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Carry)) %>% 
  select(date, Carry)

value_rts <- cncy_value %>% 
  arrange(date) %>% 
  mutate(Value = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Value)) %>% 
  select(date, Value)

dxy_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(DXY = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(DXY)) %>% 
  select(date, DXY)

link_df <- zar_daily %>% 
  inner_join(carry_rts, by = "date") %>% 
  inner_join(value_rts, by = "date") %>% 
  inner_join(dxy_rts,   by = "date")

round(cor(link_df %>% select(-date), use = "complete.obs"), 3)

```


 QUESTION 5 
  
  ## 0. Libraries
  
```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")

```



---
  
  ## 1. Load data and build monthly return panel
  
  This uses monthly data from 2011 onwards and drops assets with less than five calendar years of data.

```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")
library(readr)
library(dplyr)
# Load and filter MAA data
MAA <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI data
msci <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine benchmarks into a single tidy panel
df_full <- bind_rows(
  msci %>% rename(Tickers = Name),
  MAA  %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)

# Monthly data after 2010
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       # data after 2010
  group_by(Tickers) %>%
  # Keep only assets with at least five distinct calendar years of data
  filter(n_distinct(year(date)) >= 5) %>%
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  # Use last observation in each month
  filter(date == dplyr::last(date)) %>%
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   # drop first NA return in each series

# Wide return matrix (monthly)
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

---
  
  ## 2. Single period optimiser: `optim_foo()`
  
  This function

* takes a look back window in months up to a given rebalance date

* keeps only assets with at least five years of data in that window

* builds Sigma and mu using `RiskPortfolios::covEstimation()` and geometric means

* maps tickers to asset classes according to the MAA table

* bonds and credit: government rates plus corporate credit
* commodity: BCOMTR
* equity: all MSCI indices

* builds `Amat` and `bvec` with group constraints that match the exam rules

* solves with `quadprog::solve.QP()` and returns weights as a tibble.

```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,     # at least ten years
                      LB        = 0.01,
                      UB        = 0.35,          # single asset at most 35%
                      bond_UB   = 0.25,          # bonds and credit at most 25%
                      eq_UB     = 0.60,          # equity at most 60%
                      com_UB    = 0.15) {        # commodities at most 15%
  
  # 1. Select look back window (assumes one row per month)
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)
  
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  # 2. Drop date and enforce at least five years of data per asset
  X <- data.matrix(window_df[, -1])
  
  # require at least 60 non NA monthly observations (about five years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. Estimate covariance (Sigma) and geometric mean returns (Mu)
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)
  
  # 4. Asset class mapping for constraints
  # Based on the MAA image -table ( provided in the data set):
  # - LGAGTRUH, LUAGTRUU, LEATTREU: rates (government bonds)
  # - LGCPTREH, LUACTRUU, LP05TREH: credit (corporate bonds)
  # All of these form the "bonds and credit instruments" bucket.
  # - BCOMTR: commodity
  # - MSCI indices: equity
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. Base constraints (long only, fully invested, single asset caps)
  #    Amat^T w >= bvec
  Amat <- cbind(
    rep(1, N),         # sum w = 1
    diag(N),           # w_i >= LB
    -diag(N)           # -w_i >= -UB  gives w_i <= UB
  )
  
  bvec <- c(
    1,                 # fully invested
    rep(LB, N),        # lower bound
    -rep(UB, N)        # upper bound
  )
  
  # 6. Group constraints: sum over group of w_i <= cap
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # rows correspond to assets
    g_vec[idx] <- -1              # -sum_{i in group} w_i >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  
  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # first constraint (sum w = 1) is equality
  
  # 7. Solve quadratic program:
  #    min (1/2) w' Sigma w - mu' w   subject to constraints
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,
    dvec = Mu,
    Amat = Amat,
    bvec = bvec,
    meq  = meq
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

---
  
  ## 3. Quarterly rebalancing dates
  
  We work with quarter end month ends (January, April, July, October).

```{r}
# Quarter end dates from the monthly series
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  pull(date)
```

---
  
  ## 4. Rolling optimiser and results
  
```{r}
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 120 months equals ten years
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Ten year look back as required
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

kable(Result_roll %>% head(13))
```

---
  
  ## 5. Stacked bar chart of portfolio weights over time
  
```{r}
# Prepare xts object of weights at rebalance dates
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%
  .[endpoints(., "months")]

# Plot stacked weights
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund"
)

# StackBar of monthly weights (Note the stand-out rebalance weights...):
#Result_roll %>% bar_xts() %>% .[endpoints(.,'months')] %>% chart.StackedBar()

```

# Clustering 
Using the Practical 5 (https://www.fmx.nfkatzke.com/posts/2020-08-17-practical-5/), I cluster the indices on a Ledoit–Wolf shrunk correlation matrix to identify statistical comovement groups. The dendrogram suggests, for example, that [equity indices] cluster closely, while [bond/credit] and commodities sit in distinct branches.
```{r}
## 6. Cluster based comovement analysis (Practical 5)

# Use the same monthly return matrix, drop date
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Ledoit–Wolf covariance and correlation matrix (as in Practical 5)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")
)

corr_clust <- cov2cor(Sigma_clust)

# 6.2 Distance matrix using the Dower metric: d_ij = sqrt( (1 - rho_ij) / 2 )
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward linkage (AGNES)
library(cluster)

hc <- cluster::agnes(dist(distmat), method = "ward")

# Basic dendrogram (you could beautify with the ggdendro helpers from Practical 5)
plot(hc, which.plot = 2, main = "Hierarchical clustering of global indices",
     xlab = "", sub = "")

```



