---
title: "Financial Econometrics 871 Practical Exam"
output: github_document
editor_options: 
  chunk_output_type: console
---


**SU Number:** 22660348  
**Author:** Precious Nhamo  
**Course:** Financial Econometrics 871  
**Dates:** 18–19 November 2025  

Repo: <https://github.com/prexymtha/FMX_25_22660348>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo   = TRUE,   # show code
  message = FALSE,
  warning = FALSE
)
```





### Question 1 

I struggled with this question....

We can compute the rolling 3-year beta for each active manager and the ABC Fund against the benchmark.Then, we can create a scatterplot of returns vs. risk (beta) and overlay boxplots to show the distribution of betas for peers and highlight ABC Fund.


To see how much the fund differs from the rest of the industry in terms of what drives the return .We can perform a factor analysis or compare the R-squared from a market model to see how much of the returns are explained by the market.


Assess relative performance over time We can plot the cumulative returns of ABC Fund, peers, and the benchmark.

We can compute a measure of dispersion (cross-sectional standard deviation of returns) and then plot the fund's performance against this dispersion.

```{r load-data-Q1, echo = TRUE}
# Load readr for rds data sets 
pacman::p_load(readr, fmxdat)



# Load the data
Benchmark <- read_rds("data/Capped_SWIX.rds")
Active_Managers <- read_rds("data/Active_Managers.rds")
ABC_Fund <- read_rds("data/ABC_Fund.rds")

```

Since we have a benchmark , ABC fund industry peer we can't analyse all individual industry peers - we will calculate the mean or average of the peers 

```{r}
# Load required libraries with pacman
pacman::p_load(
  dplyr,
  ggplot2,
  lubridate,
  tidyr,
  roll,
  zoo
)

# 1. Set common sample based on ABC Fund dates
start_date <- as.Date("2012-07-31")
end_date   <- as.Date("2025-10-31")

# 2. Filter each data set to the common period
benchmark_common <- 
  Benchmark %>% 
  filter(date >= start_date, date <= end_date)

active_managers_common <- 
  Active_Managers %>% 
  filter(date >= start_date, date <= end_date)

abc_data <- 
  ABC_Fund %>% 
  filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# 3. Prepare benchmark data
bench_data <- 
  benchmark_common %>% 
  select(date, Benchmark_Returns = Returns)

# 4. Prepare peer summary by date
peers_data <- 
  active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns    = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count          = n(),
    .groups              = "drop"
  )

# 5. Merge into a single analysis data set
analysis_data <- 
  abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data,  by = "date")

# Optional quick check
dplyr::glimpse(analysis_data)


```

We will be using analysis data to continue with our questions 



```{r}
pacman::p_load(
  dplyr,
  tbl2xts,
  PerformanceAnalytics
)

# Ra: assets we are analysing (ABC + peer average)
Ra_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>% 
  mutate(across(-date, ~ coalesce(., 0))) %>% 
  tbl_xts(
    cols_to_xts = c("ABC_Fund", "Peer_Average")
  )

# Rb: benchmark we compare against
Rb_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    Benchmark = Benchmark_Returns
  ) %>% 
  mutate(Benchmark = coalesce(Benchmark, 0)) %>% 
  tbl_xts(
    cols_to_xts = "Benchmark"
  )

```

```{r}
# Rolling 3 year betas (36 months) vs benchmark
PerformanceAnalytics::chart.RollingRegression(
  Ra         = Ra_xts,
  Rb         = Rb_xts,
  width      = 36,
  attribute  = "Beta",
  legend.loc = "topleft",
  main       = "Rolling 3 year beta vs benchmark: ABC vs peer average"
)

# Full sample betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = Ra_xts,
  Rb = Rb_xts
)


```

Box-Scatter Plot ( I struggled with this part)

```{r}
pacman::p_load(
  dplyr,
  tidyr,
  ggplot2,
  roll,
  fmxdat
)

# 1. Compute rolling 3 year betas (36 months) vs benchmark

# make sure data is ordered
beta_data <- 
  analysis_data %>% 
  arrange(date)

# rolling regression: y on x with width = 36
# ABC vs Benchmark
roll_abc <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$ABC_Returns,
    width = 36
  )

# Peer average vs Benchmark
roll_peer <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$Peers_Avg_Returns,
    width = 36
  )

# extract slope (beta) from coefficients[, 2]
rolling_betas <- 
  tibble(
    date        = beta_data$date,
    ABC_Fund    = as.numeric(roll_abc$coefficients[, 2]),
    Peer_Average = as.numeric(roll_peer$coefficients[, 2])
  ) %>% 
  pivot_longer(
    cols      = c(ABC_Fund, Peer_Average),
    names_to  = "Series",
    values_to = "Beta"
  ) %>% 
  filter(!is.na(Beta))   # drop early windows with no full 36 months

# 2. Boxplot with scatter (jitter) on top

ggplot(rolling_betas, aes(x = Series, y = Beta)) +
  geom_jitter(width = 0.1, alpha = 0.4) +   # scatter points
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +  # box on top
  labs(
    title = "Distribution of rolling 3 year betas vs benchmark",
    x     = NULL,
    y     = "Rolling 3 year beta"
  ) +
  theme_fmx()

```






### Question 2

To answer the first part of Question 2 on whether smaller stocks are more volatile than larger stocks, I focus on the last decade(2015-2024) of daily returns and compare volatility across size groups(Small , Mid , Large Caps). I work with the `Hold_Rets_Sectors` data set, which contains daily returns, holdings weights, size group (`Group`) and sector labels for JSE stocks.

In the lecturer's notes, volatility (risk) is defined and measured using the standard deviation of returns, and it is emphasised that:

* Standard deviation should be annualised when comparing different time periods.
* The usual shortcut ( \text{SD} \times \sqrt{12} ) is only theoretically correct when using **logarithmic** returns, because annual log return is the sum of monthly log returns.

I proceed as follows:

1. I construct value weighted indices for each size group (Small Caps, Mid Caps and Large Caps).
2. I convert their monthly simple returns to log returns using (\log(1 + r)).
3. I compute a 36-month rolling standard deviation of these monthly log returns.
4. I annualise this rolling standard deviation by multiplying by (\sqrt{12}).

This yields a rolling annualised standard deviation of returns for each size group .If the Small Cap series has a consistently higher annualised standard deviation, I can conclude that smaller stocks have been more volatile than larger stocks locally.

### Load libraries and raw data

```{r load-data-Q2, echo = TRUE, message = FALSE, warning = FALSE}
# Load required packages
pacman::p_load(
  readr,
  tidyverse,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  rmsfuns,
  fmxdat,
  DescTools,
  RcppRoll
)

# Load the holdings and returns data with sector and size group labels
Hold_Rets <- read_rds("data/Hold_Rets_Sectors.rds") %>%
  mutate(
    # Make the size and sector groups explicit instead of leaving them as NA
    Group  = if_else(is.na(Group),  "Unclassified", Group),
    Sector = if_else(is.na(Sector), "Unclassified", Sector)
  )

head(Hold_Rets)
```

Here I explicitly rename `NA` groups to `"Unclassified"` so they do not silently disappear from the data. Later, when I compare volatility across size groups, I restrict the analysis to the proper size categories (Small_Caps, Mid_Caps, Large_Caps) and treat "Unclassified" separately if needed.

### Data Preparation and Index Construction

```{r prep-and-index-Q2, echo = TRUE}
options(dplyr.summarise.inform = FALSE)

# Filter data for the past decade and main size groups
Hold_Rets_decade <- 
  Hold_Rets %>% 
  filter(Group %in% c("Small_Caps", "Mid_Caps", "Large_Caps","Unclassified")) %>% 
  mutate(year = year(date))

# Construct daily value-weighted size index returns
idx_daily <- 
  Hold_Rets_decade %>% 
  group_by(date, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(Tickers = Group) %>%      # so Tickers is "all groups"
  arrange(Tickers, date)

# Convert to monthly returns
idx <- 
  idx_daily %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Tickers, YM) %>% 
  # pick last date of the month and compute monthly return
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Tickers, date)

head(idx)
```

###Performance Analysis - Returns

```{r performance-analysis-Q2, echo = TRUE}
# Prepare data for multi-horizon performance measures
dfplot <- 
  bind_rows(
    # 6 months - do NOT annualise 
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% 
      mutate(Freq = "A"),
    
    # 1 year
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 12) - 1) %>% 
      mutate(Freq = "B"),
    
    # 3 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 36) - 1) %>% 
      mutate(Freq = "C"),
    
    # 5 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 60) - 1) %>% 
      mutate(Freq = "D")
  )

# Visualization functions and plots
to_string <- as_labeller(c(
  `A` = "6 Months", 
  `B` = "1 Year", 
  `C` = "3 Years", 
  `D` = "5 Years"
))

# Annualized returns bar plot
g_ann <- 
  dfplot %>% 
  ggplot() + 
  geom_bar(aes(Tickers, mu, fill = Tickers), stat = "identity") + 
  facet_wrap(~Freq, labeller = to_string, nrow = 1) + 
  labs(
    x = "",
    y = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualized terms."
  ) + 
  fmx_fills() + 
  geom_label(
    aes(Tickers, mu, label = paste0(round(mu, 4) * 100, "%")),
    size = ggpts(8),
    alpha = 0.35,
    fontface = "bold",
    nudge_y = 0.002
  ) + 
  theme_fmx(
    CustomCaption = TRUE,
    title.size = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size = ggpts(30),
    axis.size = ggpts(37),
    legend.size = ggpts(35),
    legend.pos = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

# Cumulative returns plots
gg_cum <- 
  idx %>% 
  arrange(date) %>% 
  group_by(Tickers) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Tickers)) +
  scale_colour_manual(
    values = c(
      "Large_Caps"   = "#1f77b4",
      "Mid_Caps"     = "#2ca02c",
      "Small_Caps"   = "#d62728",
      "Unclassified" = "grey40"
    )) +
  labs(
    title    = "Cumulative Returns by Cap Size Group",
    subtitle = "Growth of 1 rand invested in each group (rebased to 1 at first observation)",
    caption  = "Note: Each size cap is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  )

gg_cum

# Log cumulative plot
gg_cum_log <- 
  gg_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_log
```

### Volatility Analysis

```{r volatility-analysis-Q2, echo = TRUE}
# Rolling returns analysis
plotdf <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll <- 
  plotdf %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Volatility analysis - answer to the volatility question
plot_dlog <- 
  idx %>% 
  group_by(Tickers) %>% 
  # 1. Convert simple monthly returns to log returns
  mutate(ret = log(1 + ret)) %>% 
  # 2. Rolling 36-month SD of log returns, annualised
  mutate(
    RollSD = RcppRoll::roll_sd(ret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(RollSD)) %>% 
  # 3. Flag the "focus" buckets and relevel for nicer colours
  mutate(
    SizeBucket = if_else(Tickers %in% c("Small_Caps", "Large_Caps"),
                         "focus", "other"),
    Tickers    = forcats::fct_relevel(Tickers, "Small_Caps", "Large_Caps")
  )

g_sd <- 
  plot_dlog %>% 
  ggplot() +
  geom_line(
    aes(date, RollSD, colour = Tickers, alpha = SizeBucket, size = SizeBucket)
  ) +
  # Make Small/Large bold, others dim
  scale_alpha_manual(values = c(focus = 1,   other = 0.25), guide = "none") +
  scale_size_manual( values = c(focus = 1.8, other = 0.7),   guide = "none") +
  labs(
    title   = "Rolling 3 Year Annualized SD by Size Bucket",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year SD (Ann.)",
    caption = "Note:\nAnnualized SD computed on log monthly returns."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_sd, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = FALSE)
```

### Sector-Level Analysis

```{r sector-analysis-Q2, echo = TRUE}
# Monthly value-weighted returns by Sector and Group (size)
idx_sector_size <- 
  Hold_Rets_decade %>% 
  group_by(date, Sector, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Sector, Group, YM) %>% 
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Sector, Group, date)

# Rolling returns analysis by sector and size
plotdf_sector <- 
  idx_sector_size %>% 
  group_by(Sector, Group) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll_sector <- 
  plotdf_sector %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Group), alpha = 0.7, size = 1.25) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualized Returns by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll_sector, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

# Sector-level volatility analysis
plot_sd_sector <- 
  Hold_Rets_decade %>%
  # value-weighted daily returns per Sector × Size Group
  group_by(date, Sector, Group) %>%
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(YM = format(date, "%Y%B")) %>%
  arrange(Sector, Group, date) %>%
  # collapse to month-end and compute monthly returns
  group_by(Sector, Group, YM) %>%
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>%
  arrange(Sector, Group, date) %>%
  group_by(Sector, Group) %>%
  # log monthly returns and rolling 36-month SD, annualised
  mutate(
    ret_log = log1p(ret),
    RollSD  = RcppRoll::roll_sd(ret_log, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  ungroup() %>%
  filter(!is.na(RollSD))

g_sd_sector <- 
  plot_sd_sector %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Group), size = 1.3, alpha = 0.9) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title   = "Rolling 3 Year Annualised SD by Size and Sector",
    subtitle = "",
    x       = "",
    y       = "Rolling 3-year SD (Ann.)"
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5)
  ) +
  fmx_cols()

finplot(
  g_sd_sector,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)

# Cumulative returns by sector and size
idx_sector_size_cum <- 
  idx_sector_size %>% 
  arrange(Sector, Group, date) %>% 
  group_by(Sector, Group) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup()

gg_cum_sector <- 
  idx_sector_size_cum %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Group)) +
  facet_wrap(~ Sector, nrow = 2) +
  labs(
    title    = "Cumulative Returns by Sector and Size Group",
    subtitle = "Growth of 1 rand invested in each sector-size group (rebased to 1 at first observation)",
    caption  = "Note: Each sector-size group is normalised to 1 at its first observation."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

gg_cum_sector

# Log cumulative plot by sector
gg_cum_sector_log <- 
  gg_cum_sector +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum_sector$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_sector_log
```

### Drawdown Analysis

```{r drawdown-analysis-Q2, echo = TRUE}
# Convert to xts for PerformanceAnalytics
idxxts <- 
  idx %>% 
  filter(Tickers != "Unclassified") %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Tickers)

# Main drawdown chart
chart.Drawdown(
  idxxts[, c("Large_Caps", "Mid_Caps", "Small_Caps")],
  geometric  = TRUE,
  legend.loc = "bottomleft",
  colorset   = 1:3,
  lwd        = 3,
  main       = "Peak to trough drawdowns by size group",
  ylab       = "Drawdown from peak"
)

# Sector drawdowns
draw_sector_dd <- function(sec) {
  
  groups_to_plot <- c("Large_Caps", "Mid_Caps", "Small_Caps")
  
  sec_xts <- 
    idx_sector_size %>% 
    filter(
      Sector == sec,
      Group %in% groups_to_plot    # drop Unclassified
    ) %>% 
    select(date, Group, ret) %>% 
    tbl_xts(cols_to_xts = ret, spread_by = Group)
  
  # Only use the groups that actually exist for this sector
  avail_groups <- intersect(groups_to_plot, colnames(sec_xts))
  
  chart.Drawdown(
    sec_xts[, avail_groups],
    geometric  = TRUE,
    legend.loc = "bottomleft",
    colorset   = 1:length(avail_groups),
    lwd        = 4,
    main       = paste("Drawdowns by size -", sec),
    ylab       = "Drawdown from peak"
  )
}

# Individual sector drawdown charts
draw_sector_dd("Industrials")
draw_sector_dd("Resources")
draw_sector_dd("Financials")
draw_sector_dd("Property")
```

### Statistical Analysis Tables 

(Added extra from data-camp maybe unnecessary)

```{r statistical-tables-Q2, echo = TRUE}
# Time series chart of sector returns using PerformanceAnalytics
sector_xts <- 
  idx_sector_size %>% 
  filter(Group %in% c("Large_Caps", "Mid_Caps", "Small_Caps")) %>% 
  select(date, Sector, Group, ret) %>% 
  mutate(Sector_Group = paste(Sector, Group, sep = "_")) %>% 
  select(-Sector, -Group) %>% 
  tbl_xts(cols_to_xts = ret, spread_by = Sector_Group)

# Chart time series of selected sector-size combinations
chart.TimeSeries(
  sector_xts[, c("Industrials_Large_Caps", "Resources_Large_Caps", "Financials_Large_Caps")],
  main = "Large Cap Returns by Sector",
  legend.loc = "topright",
  colorset = 1:3,
  lwd = 2,
  ylab = "Monthly Returns"
)

# Drawdown analysis tables
drawdown_results <- table.Drawdowns(sector_xts[, 1:6])
drawdown_results

# Maximum drawdown by sector-size
max_drawdowns <- maxDrawdown(sector_xts)
max_drawdowns

# Detailed drawdown analysis
detailed_drawdowns <- table.Drawdowns(sector_xts[, 1:6])
detailed_drawdowns

# Downside risk table
downside_risk <- table.DownsideRisk(sector_xts[, 1:6])
downside_risk

# Find and sort drawdowns for specific sectors
industrials_drawdowns <- findDrawdowns(sector_xts[, "Industrials_Large_Caps"])
sorted_industrials <- sortDrawdowns(industrials_drawdowns)
sorted_industrials

# Resources sector drawdown analysis
resources_drawdowns <- findDrawdowns(sector_xts[, "Resources_Large_Caps"])
sorted_resources <- sortDrawdowns(resources_drawdowns)
sorted_resources
```

### Conclusion


Using a  Rolling 3-year annualized standard deviations consistently show that Small Caps exhibit higher return variability across all sectors, with particularly sharp spikes during periods of market stress like 2020. Peak-to-trough drawdown analysis further confirms this: Small and Mid Caps experience deeper and more frequent drawdowns than Large Caps, especially in sectors like Industrials and Property, where Mid Caps saw losses exceeding 65%. Interestingly, while Large Caps are more stable, they are not immune to prolonged drawdowns, such as the 48-month decline in Financials from 2018 to 2022. These findings highlight the importance of size and sector in understanding risk dynamics and suggest that portfolio construction should account for both cyclical sensitivity and recovery behavior across market segments.

Seeing that we have individual stocks ( maybe we could have assessed the most volatility and which sectors they are found by their sizes but I was unsure how to do that , plus I felt my solution was way all over the place trying to include all code so I excluded it)





### Question 3 

### Question 4: Volatility and GARCH estimates


We are evaluating whether the South African Rand (ZAR) has been among the most volatile currencies in recent years, which requires comparing its behavior to other currencies. Historically, the ZAR tends to perform well during periods when G10 currency carry trades are favorable and when currency valuations are relatively cheap—these can be analyzed separately and jointly. Additionally, the ZAR often benefits during times of Dollar strength, reflecting global risk-on sentiment. You have full discretion in choosing how to measure volatility, which currencies to compare, and how to structure your analysis.

#### Libraries and data
```{r Q4-libs-data, echo = TRUE}
# Load all required packages 
pacman::p_load(
  tidyverse,     
  devtools,     
  rugarch,       # For GARCH modeling - this was crucial
  rmgarch,       # For multivariate GARCH
  forecast,      # For time series functions like ACF
  tbl2xts,      
  lubridate,     
  PerformanceAnalytics, 
  ggthemes,      
  robustbase,    
  kableExtra,   
  RcppRoll       
)

# Load currency data - all quoted vs USD

cncy       <- read_rds("data/currencies.rds")        # Raw currency prices
cncy_Carry <- read_rds("data/cncy_Carry.rds")        # G10 carry index - for statement 2
cncy_value <- read_rds("data/cncy_value.rds")        # FX PPP value index - for statement 2  
cncyIV     <- read_rds("data/cncyIV.rds")            # FX implied vol index
bbdxy      <- read_rds("data/bbdxy.rds")             # Bloomberg Dollar spot index - for Dollar strength analysis
```

### Cross sectional volatility: ZAR vs G10 vs others
I calculate returns for all currencies to measure volatility
```{r Q4-vol-panel, echo = TRUE}
# First, I need to calculate returns for all currencies to measure volatility
# Using log returns 
cncy_rts <- cncy %>%  
  group_by(Name) %>%  # Calculate returns for each currency separately
  arrange(date, .by_group = TRUE) %>% # Ensure dates are in order
  mutate(
    dlogret   = log(Price) - log(lag(Price)),  # Daily log return
    scaledret = dlogret - mean(dlogret, na.rm = TRUE)  # Scaled return (de-meaned)
  ) %>% 
  filter(date > dplyr::first(date)) %>% # Remove first row with NA return
  ungroup() %>% 
  mutate(Name = gsub("_Cncy", "", Name)) # Clean up currency names

# Define G10 currencies 
# I looked up which currencies are considered G10 and matched them to our data ( some not included in our data )
g10_vec <- c(
  "Australia_Inv",  # AUD
  "Canada",         # CAD
  "EU_Inv",         # EUR
  "Japan",          # JPY
  "NZ_Inv",         # NZD
  "Norway",         # NOK
  "Sweden",         # SEK
  "UK_Inv"          # GBP
  # USD is the base currency here, so not included
)

# Now calculate various volatility measures since 2015
# I'm using multiple measures to get a comprehensive view
vol_cncy <- cncy_rts %>% 
  filter(date > ymd(20150101)) %>% # Focus on recent period as mentioned in question
  group_by(Name) %>%
  summarise(
    std_dev      = sd(dlogret, na.rm = TRUE),  # Standard deviation - most common measure
    mad          = mean(abs(dlogret - mean(dlogret, na.rm = TRUE)), na.rm = TRUE), # Mean absolute deviation
    downside_dev = sqrt(mean(ifelse(dlogret < 0, dlogret^2, 0), na.rm = TRUE)), # Downside deviation (semi-deviation)
    d_ratio      = (n() - sum(ifelse(dlogret < 0, dlogret, 0))) / 
                   (n() + sum(ifelse(dlogret > 0, dlogret, 0))), # Not sure if this is the correct formulation
    .groups = "drop"
  ) %>% 
  mutate(
    # Create groups for comparison: ZAR vs G10 vs Other emerging markets
    Group = case_when(
      Name == "SouthAfrica" ~ "ZAR",      # Our currency of interest
      Name %in% g10_vec     ~ "G10",      # Major developed currencies
      TRUE                  ~ "Other"     # Everything else
    )
  )

# Rank currencies by volatility to see where ZAR stands
ranked_data <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% # Sort from highest to lowest volatility
  mutate(rank = row_number()) %>% # Add rank column
  select(rank, Name, Group, everything()) # Reorder columns


kableExtra::kable(
  ranked_data,
  caption = "Volatility ranking of currencies (log returns, since 2015)",
  digits = 6
)
```


SA is the 8th most volatile currency so we can conclude that the statement is true.


```{r Q4-vol-plot, echo = TRUE}
# Create visualization to make the volatility comparison clear
# First, add ranks to our volatility data
vol_with_rank <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number())

# We want to show: top 10 most volatile currencies OR any G10 currency
# This ensures we see ZAR's position and how G10 currencies compare
plot_set <- vol_with_rank %>% 
  filter(rank <= 10 | Group == "G10")

# Create the plot - using bars ordered by volatility
plot_set %>% 
  mutate(Name = fct_reorder(Name, std_dev)) %>% # Order bars by volatility
  ggplot() +
  geom_col(aes(x = Name, y = std_dev, fill = Group), alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "ZAR"   = "red",      # Highlight ZAR in red
      "G10"   = "grey40",   # G10 in dark grey
      "Other" = "grey80"    # Others in light grey
    )
  ) +
  coord_flip() + # Horizontal bars are easier to read with many categories
  labs(
    title    = "Cross sectional volatility of currencies",
    subtitle = "Top 10 currencies by volatility plus G10 majors (since 2015)",
    x = "",
    y = "Standard deviation",
    fill = ""
  ) +
  fmxdat::theme_fmx() 
```

Again , SA is more volatile than the G10 currencies.

### 2. Analyzing ZAR return persistence
```{r Q4-zar-persistence, echo = TRUE}
# Now i focus specifically on ZAR for deeper analysis
# Using simple returns this time instead of log returns for GARCH modeling
zar_rts <- cncy %>%  
  filter(date > ymd(20150101)) %>% # Consistent time period
  group_by(Name) %>%  
  mutate(ret = Price / lag(Price) - 1) %>% # Simple return calculation
  filter(date > dplyr::first(date)) %>% # Remove NA
  ungroup() %>%  
  mutate(Name = gsub("_Cncy", "", Name)) %>% # Clean names
  filter(Name == "SouthAfrica") %>% # Focus on ZAR only
  filter(!is.na(ret)) %>% # Remove any remaining NAs
  select(-Name, -Price) %>% # Keep only date and returns
  tbl_xts() # Convert to xts format for time series analysis

# I  Create three different return series to examine persistence (practical 6 ):
# Raw returns, squared returns (variance proxy), absolute returns (volatility proxy)
Plotdata <- cbind(zar_rts, zar_rts^2, abs(zar_rts))
colnames(Plotdata) <- c("Returns", "Returns_Sqd", "Returns_Abs")

# Convert back to tidy format for ggplot
Plotdata <- Plotdata %>% 
  xts_tbl() %>% 
  gather(ReturnType, Returns, -date)

# Plot to visually inspect persistence patterns
ggplot(Plotdata) + 
  geom_line(aes(x = date, y = Returns, colour = ReturnType, alpha = 0.5)) + 
  ggtitle("Return Type Persistence: ZAR") + 
  facet_wrap(~ReturnType, nrow = 3, ncol = 1, scales = "free") + # Separate plots for each type
  guides(alpha = "none", colour = "none") + # Remove legends
  fmxdat::theme_fmx()
```

```{r Q4-zar-acf, echo = TRUE}
# Autocorrelation analysis - prep for GARCH
# If returns show no autocorrelation but squared returns do, it suggests volatility clustering
forecast::Acf(zar_rts,   main = "ACF: ZAR return")           # Check raw return persistence
forecast::Acf(zar_rts^2, main = "ACF: Squared ZAR return")   # Check volatility persistence
forecast::Acf(abs(zar_rts), main = "ACF: Absolute ZAR return") # Alternative volatility measure

# Formal test for ARCH effects - significant p-value means volatility clustering exists
Box.test(coredata(zar_rts^2), type = "Ljung-Box", lag = 12)
```

“X-squared = 101.98, df = 12, p-value = 2.22e-16.” The small p-value means we reject the null hypothesis (no autocorrelation) in the squared returns. This suggests the presence of volatility clustering, and using a GARCH model could help capture the time-varying risk.


### Fitting GARCH(1,1) to ZAR (rugarch, copy of Practical 6)
```{r Q4-zar-garch-fit, echo = TRUE}
# Following the practical example structure for consistency
porteqw <- zar_rts # Renaming to match practical example

# Specify GARCH(1,1) model - this took some time to get right
garch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[1], # Standard GARCH
    garchOrder = c(1, 1) # 1 ARCH term, 1 GARCH term
  ),
  mean.model = list(
    armaOrder    = c(1, 0), # AR(1) model for mean
    include.mean = TRUE     # Include constant in mean equation
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[1] # Normal distribution
)

# Fit the model - this can take a moment
garchfit1 <- ugarchfit(spec = garch11, data = porteqw)

# Display results
garchfit1
```

```{r Q4-zar-garch-table, echo = TRUE}
# Extract coefficient table 
garch_tab <- garchfit1@fit$matcoef
kableExtra::kable(garch_tab, caption = "GARCH(1,1) estimates for ZAR")
```

```{r Q4-zar-sigma-plot, echo = TRUE}
# Plot the estimated conditional volatility from GARCH
sigma <- sigma(garchfit1) %>% xts_tbl() # Extract conditional sigma
colnames(sigma) <- c("date", "sigma") 
sigma <- sigma %>% mutate(date = as.Date(date))

# Compare GARCH volatility with simple squared returns volatility
gg <- ggplot() + 
  # Plot rolling volatility from squared returns (naive approach)
  geom_line(
    data = Plotdata %>% 
      filter(ReturnType == "Returns_Sqd") %>% 
      select(date, Returns) %>% 
      unique() %>% 
      mutate(Returns = sqrt(Returns)), # Convert variance to volatility
    aes(x = date, y = Returns)
  ) + 
  # Overlay GARCH volatility in red
  geom_line(
    data = sigma,
    aes(x = date, y = sigma),
    color = "red", size = 2, alpha = 0.8
  ) + 
  labs(
    title    = "Comparison: Returns sigma vs sigma from GARCH",
    subtitle = "ZAR: note the smoothing effect of GARCH as noise is controlled for.",
    x = "", y = "Estimated volatility",
    caption = "Source: Financial econometrics practical | Calculations: Own"
  ) + 
  fmxdat::theme_fmx(CustomCaption = TRUE)


fmxdat::finplot(gg, y.pct = TRUE, y.pct_acc = 1)
```

```{r Q4-zar-nic, echo = TRUE}
# News Impact Curve - shows how news (shocks) affect future volatility
ni <- newsimpact(z = NULL, garchfit1)

# Plot the NIC - I struggled to understand this at first
# It shows asymmetry: how positive vs negative shocks impact volatility differently
plot(
  ni$zx, ni$zy,
  ylab = ni$yexpr,
  xlab = ni$xexpr,
  type = "l",
  main = "News Impact Curve: ZAR sGARCH(1,1)"
)


```

### Alternative GARCH forms for ZAR (GJR and EGARCH) and NIC comparison
```{r Q4-zar-gjr-egarch, echo = TRUE}
# Standard GARCH assumes symmetric volatility response
# But in finance, negative shocks often increase volatility more (leverage effect)
# So i try asymmetric GARCH models

# GJR-GARCH with student t distribution - accounts for leverage effect
gjrgarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[2], # GJR-GARCH
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3] # Student t
)

garchfit2 <- ugarchfit(spec = gjrgarch11, data = as.matrix(porteqw))

# Display results
kableExtra::kable(
  garchfit2@fit$matcoef,
  caption = "GJR GARCH(1,1) with student t distribution for ZAR"
)

# EGARCH - another asymmetric model with exponential form
egarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[3], # EGARCH
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3] # Student t
)

garchfit3 <- ugarchfit(spec = egarch11, data = as.matrix(porteqw))

# Model diagnostics and comparison
signbias(garchfit1) # Test for sign bias - if significant, asymmetric models are better
infocriteria(garchfit1) # Information criteria for model selection
infocriteria(garchfit2)
infocriteria(garchfit3)
```

```{r Q4-zar-nic-compare, echo = TRUE}
# Compare News Impact Curves across all three models
#function to extract NIC data - copied from practical
NICurveMaker <- function(Fit, Name) {
  NI <- newsimpact(z = NULL, Fit)
  NI <- cbind(NI$zx, NI$zy)
  colnames(NI) <- c(paste0("Epsilon_", Name), paste0("Sigma_", Name))
  NI %>% data.frame() %>% tibble::as_tibble()
}

# Extract NIC data for all three models
NI1 <- NICurveMaker(garchfit1, "GARCH11")
NI2 <- NICurveMaker(garchfit2, "GJR")
NI3 <- NICurveMaker(garchfit3, "EGARCH")

# Combine and reshape for plotting
NI <- cbind(NI1, NI2, NI3) %>% 
  gather(Model, Sigma, starts_with("Sigma")) %>% 
  rename(Epsilon = Epsilon_GARCH11) %>% 
  select(-Epsilon_GJR, -Epsilon_EGARCH)

# Plot comparison - this clearly shows asymmetry differences
ggplot(NI) + 
  geom_line(aes(x = Epsilon, y = Sigma, colour = Model)) + 
  ggtitle("News Impact Curves: ZAR") + 
  theme_hc()


```



### GO GARCH and time varying correlation between ZAR and Dollar strength
```{r Q4-zar-go-garch-data, echo = TRUE}
# Now  i  examine the second statement about Dollar strength using prac 7 - multivariate analysis
#I analyse the  correlation between ZAR and Dollar index

# Calculate Dollar index log returns
g10_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(G10 = log(Price) - log(lag(Price))) %>% # Log returns for Dollar index
  filter(date > dplyr::first(date)) %>% 
  select(date, G10)

# Get ZAR log returns from our earlier calculation
zar_log_rts <- cncy_rts %>% 
  filter(Name == "SouthAfrica") %>% 
  rename(ZAR = dlogret) %>% 
  select(date, ZAR)

# Merge and convert to xts for multivariate GARCH
xts_rtn <- left_join(g10_rts, zar_log_rts, by = "date") %>% 
  drop_na() %>% # Remove any missing values
  tbl_xts() # Convert to xts format
```



```{r Q4-renamingdcc, echo = TRUE}
# function from practical to rename DCC correlation outputs
# This was complex - had to carefully study how the correlation matrices are structured ( struggled a bit )
renamingdcc <- function(ReturnSeries, DCC.TV.Cor) {
  dates <- zoo::index(ReturnSeries)
  N     <- ncol(ReturnSeries)
  names <- colnames(ReturnSeries)
  
  # Create all possible pairs
  pairs      <- expand.grid(names, names)
  pair_names <- paste0(pairs$Var1, "_", pairs$Var2)
  
  df <- as.data.frame(DCC.TV.Cor)
  colnames(df) <- pair_names
  df$date      <- dates
  
  # Reshape to long format for plotting
  df %>% 
    tidyr::pivot_longer(
      cols      = -date,
      names_to  = "Pairs",
      values_to = "Rho"
    ) %>% 
    dplyr::arrange(date)
}
```



```{r Q4-zar-go-garch-fit, echo = TRUE}
# GO-GARCH model for multivariate volatility
# This allows us to estimate time-varying correlation

# First specify univariate GARCH models for each series
uspec <- ugarchspec(
  variance.model = list(
    model      = "gjrGARCH", # Using GJR-GARCH for asymmetry
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = "sstd" # Skewed student t for fat tails and asymmetry
)

# Replicate specification for both series
multi_univ_garch_spec <- multispec(replicate(ncol(xts_rtn), uspec))

# GO-GARCH specification 
spec.go <- gogarchspec(
  multi_univ_garch_spec,
  distribution.model = "mvnorm", # Multivariate normal
  ica                = "fastica" # Independent component analysis
)

# Use parallel processing - this speeds up estimation significantly
cl <- makePSOCKcluster(10)

# Fit univariate models first
multf <- multifit(multi_univ_garch_spec, xts_rtn, cluster = cl)

# Fit GO-GARCH model - this can take a while
fit.gogarch <- gogarchfit(
  spec    = spec.go,
  data    = xts_rtn,
  solver  = "hybrid",
  cluster = cl,
  gfun    = "tanh",
  maxiter1 = 40000,
  epsilon  = 1e-08,
  rseed    = 100
)

stopCluster(cl) # Always stop cluster when done

print(fit.gogarch)
```



```{r Q4-zar-go-garch-tv-cor, echo = TRUE}
# Extract time-varying correlations
gog.time.var.cor <- rcor(fit.gogarch)
gog.time.var.cor <- aperm(gog.time.var.cor, c(3, 2, 1)) # Rearrange dimensions
dim(gog.time.var.cor) <- c(
  nrow(gog.time.var.cor),
  ncol(gog.time.var.cor)^2
)

# Use renamingdcc function to rename and reshape
gog.time.var.cor <- renamingdcc(
  ReturnSeries = xts_rtn,
  DCC.TV.Cor   = gog.time.var.cor
)

# Plot the time-varying correlation between ZAR and Dollar index
g2 <- ggplot(
  gog.time.var.cor %>% 
    filter(grepl("ZAR_", Pairs), !grepl("_ZAR", Pairs)) # Get ZAR pairs but not ZAR_ZAR
) +
  geom_line(aes(x = date, y = Rho, colour = Pairs)) +
  theme_hc() +
  ggtitle("Go GARCH: time varying correlation between ZAR and Dollar index")

print(g2)
```

---

###  G10 carry and PPP value (for narrative) 

I added this but was unsure about it 
```{r Q4-zar-carry-value-links, echo = TRUE}
# Finally, let's examine the relationships mentioned in the second statement
# Calculate returns for all relevant series

# ZAR returns (already calculated)
zar_daily <- cncy_rts %>% filter(Name == "SouthAfrica") %>% select(date, ZAR = dlogret)

# G10 carry index returns
carry_rts <- cncy_Carry %>% 
  arrange(date) %>% 
  mutate(Carry = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Carry)) %>% 
  select(date, Carry)

# PPP value index returns  
value_rts <- cncy_value %>% 
  arrange(date) %>% 
  mutate(Value = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Value)) %>% 
  select(date, Value)

# Dollar index returns (alternative calculation)
dxy_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(DXY = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(DXY)) %>% 
  select(date, DXY)

# Combine all series
link_df <- zar_daily %>% 
  inner_join(carry_rts, by = "date") %>% 
  inner_join(value_rts, by = "date") %>% 
  inner_join(dxy_rts,   by = "date")

# Calculate correlation matrix - this gives us the relationships
round(cor(link_df %>% select(-date), use = "complete.obs"), 3)
```

ZAR ranks as the 8th most volatile currency among 41 currencies in the sample (starting year = 2015 , if we change it to 2016/2017 it moves up to number 5), and is more volatile than all G10 currencies. Univariate GARCH models confirm strong and persistent volatility clustering for ZAR. The asymmetric GJR GARCH model further suggests that positive shocks to ZAR returns increase volatility more than negative shocks of the same magnitude.


ZAR shows moderate relationships with carry and value factors, with correlations of about −0.33 with the G10 carry index and +0.25 with the value index. However, the claim that ZAR benefits from periods of strong Dollar performance is not supported by the data. ZAR returns are positively correlated with the Dollar index (about +0.54), which in this quotation convention means that ZAR typically weakens when the Dollar strengthens. GO GARCH results indicate that this relationship is time varying, so in this sample the statement is not strongly supported.


The GJR GARCH model indicates that ZAR volatility responds more strongly to positive shocks than to negative shocks (γ ≈ −0.05), which is the opposite of the usual leverage effect found in equity markets. This points to a different volatility mechanism for ZAR, consistent with the idea that good news rallies and periods of strong inflows can be associated with elevated and persistent volatility in an emerging market currency.





 QUESTION 5 
  
  ## 0. Libraries
  
```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")

```



---
  
  ## 1. Load data and build monthly return panel
  
  This uses monthly data from 2011 onwards and drops assets with less than five calendar years of data.

```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")
library(readr)
library(dplyr)
# Load and filter MAA data
MAA <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI data
msci <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine benchmarks into a single tidy panel
df_full <- bind_rows(
  msci %>% rename(Tickers = Name),
  MAA  %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)

# Monthly data after 2010
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       # data after 2010
  group_by(Tickers) %>%
  # Keep only assets with at least five distinct calendar years of data
  filter(n_distinct(year(date)) >= 5) %>%
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  # Use last observation in each month
  filter(date == dplyr::last(date)) %>%
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   # drop first NA return in each series

# Wide return matrix (monthly)
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

---
  
  ## 2. Single period optimiser: `optim_foo()`
  
  This function

* takes a look back window in months up to a given rebalance date

* keeps only assets with at least five years of data in that window

* builds Sigma and mu using `RiskPortfolios::covEstimation()` and geometric means

* maps tickers to asset classes according to the MAA table

* bonds and credit: government rates plus corporate credit
* commodity: BCOMTR
* equity: all MSCI indices

* builds `Amat` and `bvec` with group constraints that match the exam rules

* solves with `quadprog::solve.QP()` and returns weights as a tibble.

```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,     # at least ten years
                      LB        = 0.01,
                      UB        = 0.35,          # single asset at most 35%
                      bond_UB   = 0.25,          # bonds and credit at most 25%
                      eq_UB     = 0.60,          # equity at most 60%
                      com_UB    = 0.15) {        # commodities at most 15%
  
  # 1. Select look back window (assumes one row per month)
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)
  
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  # 2. Drop date and enforce at least five years of data per asset
  X <- data.matrix(window_df[, -1])
  
  # require at least 60 non NA monthly observations (about five years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. Estimate covariance (Sigma) and geometric mean returns (Mu)
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)
  
  # 4. Asset class mapping for constraints
  # Based on the MAA image -table ( provided in the data set):
  # - LGAGTRUH, LUAGTRUU, LEATTREU: rates (government bonds)
  # - LGCPTREH, LUACTRUU, LP05TREH: credit (corporate bonds)
  # All of these form the "bonds and credit instruments" bucket.
  # - BCOMTR: commodity
  # - MSCI indices: equity
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. Base constraints (long only, fully invested, single asset caps)
  #    Amat^T w >= bvec
  Amat <- cbind(
    rep(1, N),         # sum w = 1
    diag(N),           # w_i >= LB
    -diag(N)           # -w_i >= -UB  gives w_i <= UB
  )
  
  bvec <- c(
    1,                 # fully invested
    rep(LB, N),        # lower bound
    -rep(UB, N)        # upper bound
  )
  
  # 6. Group constraints: sum over group of w_i <= cap
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # rows correspond to assets
    g_vec[idx] <- -1              # -sum_{i in group} w_i >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  
  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # first constraint (sum w = 1) is equality
  
  # 7. Solve quadratic program:
  #    min (1/2) w' Sigma w - mu' w   subject to constraints
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,
    dvec = Mu,
    Amat = Amat,
    bvec = bvec,
    meq  = meq
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

---
  
  ## 3. Quarterly rebalancing dates
  
  We work with quarter end month ends (January, April, July, October).

```{r}
# Quarter end dates from the monthly series
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  pull(date)
```

---
  
  ## 4. Rolling optimiser and results
  
```{r}
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 120 months equals ten years
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Ten year look back as required
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

kable(Result_roll %>% head(13))
```

---
  
  ## 5. Stacked bar chart of portfolio weights over time
  
```{r}
# Prepare xts object of weights at rebalance dates
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%
  .[endpoints(., "months")]

# Plot stacked weights
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund"
)

# StackBar of monthly weights (Note the stand-out rebalance weights...):
#Result_roll %>% bar_xts() %>% .[endpoints(.,'months')] %>% chart.StackedBar()

```

# Clustering 
Using the Practical 5 (https://www.fmx.nfkatzke.com/posts/2020-08-17-practical-5/), I cluster the indices on a Ledoit–Wolf shrunk correlation matrix to identify statistical comovement groups. The dendrogram suggests, for example, that [equity indices] cluster closely, while [bond/credit] and commodities sit in distinct branches.
```{r}
## 6. Cluster based comovement analysis (Practical 5)

# Use the same monthly return matrix, drop date
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Ledoit–Wolf covariance and correlation matrix (as in Practical 5)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")
)

corr_clust <- cov2cor(Sigma_clust)

# 6.2 Distance matrix using the Dower metric: d_ij = sqrt( (1 - rho_ij) / 2 )
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward linkage (AGNES)
library(cluster)

hc <- cluster::agnes(dist(distmat), method = "ward")

# Basic dendrogram (you could beautify with the ggdendro helpers from Practical 5)
plot(hc, which.plot = 2, main = "Hierarchical clustering of global indices",
     xlab = "", sub = "")

```



