---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Financial Econometrics 871 Practical Exam

**SU Number:** 22660348  
**Author:** Precious Nhamo  
**Course:** Financial Econometrics 871  
**Dates:** 18–19 November 2025



---

### Question 4: Volatility and GARCH estimates

#### 5.4.1 Aim of the question

- We are assessing whether the South African Rand (ZAR) has been one of the most volatile currency over the past few years.(since it says one of the most we assess it relative to other currencies)
- ZAR has generally performed well during periods where the G10 currency carry trades have been favourable and currency valuations cheap ( we can split this for two and then do the joint )
- Globally, it has been one of the currencies that most benefit during periods where the Dollar is comparatively strong,
 indicating a risk-on sentiment
- Use your discretion on what you use to measure volatility, which currencies you compare
 and how you arrive at your conclusions. Be creative in using tables, graphs, stratification and
 statistics to argue your points.



```{r load-data-Q4, echo = TRUE}

# Load required packages using pacman (installs if missing)
pacman::p_load(
  readr,     # for reading .rds files 
  tidyverse, # for data manipulation and visualization
  rugarch,   # for GARCH modeling (volatility analysis)
  lubridate, # for working with dates and times
  tbl2xts,   # for converting tibbles to xts time series
  zoo        # for time series infrastructure and manipulation
)

# Source any helper functions for Q5
list.files("Q5/code/", full.names = TRUE, recursive = TRUE) %>% 
  .[grepl("\\.R$", .)] %>%     # Filter only .R script files
  as.list() %>%                # Convert to list for walk()
  walk(~ source(.x))           # Source each script (e.g., custom functions)

# Load pre-processed data files (all relative to USD)
cncy       <- read_rds("data/currencies.rds")     # Daily spot rates for G10 currencies
cncy_Carry <- read_rds("data/cncy_Carry.rds")     # Deutsche Bank G10 Carry Index (DBHVG10U Index)
cncy_value <- read_rds("data/cncy_value.rds")     # Deutsche Bank FX PPP Value Index (DBPPPUSF Index)
cncyIV     <- read_rds("data/cncyIV.rds")         # Implied volatility (similar to VIX for currencies)
bbdxy      <- read_rds("data/bbdxy.rds")          # Bloomberg Dollar Spot Index (BBDXY)



```

Volatility of ZAR 

```{r}
# This step calculates various volatility measures for currency returns
# We use multiple measures because each captures different aspects of risk:
# - Standard deviation (sd) is the most common symmetric volatility measure
# - MAD is more robust to outliers than standard deviation
# - Downside deviation focuses only on negative returns, which aligns with investor concerns about losses
# - d-ratio provides a measure of return asymmetry

# Lets start with the currency returns
# Volatility of ZAR
# Currency returns following lecturer's pattern
cncy_rts <- cncy %>%  
  group_by(Name) %>%  
  arrange(date, .by_group = TRUE) %>%                 # ensure sorted
  mutate(
    dlogret   = log(Price) - log(lag(Price)),
    scaledret = dlogret - mean(dlogret, na.rm = TRUE)
  ) %>% 
  filter(date > dplyr::first(date)) %>%               # drop first date per currency
  ungroup() %>% 
  mutate(Name = gsub("_Cncy", "", Name))


# Calculate multiple volatility measures over the past few years
vol_cncy <- cncy_rts %>% 
    filter(date > as.Date("2015-01-01")) %>%  # Filter for recent period
    group_by(Name) %>%
    summarize(
        # Standard deviation - most commonly used measure due to mathematical properties
        # and compatibility with normal distribution assumptions
        std_dev = sd(dlogret, na.rm = TRUE),
        
        # Mean Absolute Deviation (MAD) - more robust to outliers than standard deviation
        mad = mean(abs(dlogret - mean(dlogret, na.rm = TRUE)), na.rm = TRUE),
        
        # Downside deviation (semi-deviation) - only considers returns below zero
        # More relevant for risk-averse investors who primarily care about losses
        downside_dev = sqrt(mean(ifelse(dlogret < 0, dlogret^2, 0), na.rm = TRUE)),
        
        # d-ratio - measures asymmetry between upside and downside
        # Values < 1 indicate more downside risk, values > 1 indicate more upside potential
        d_ratio = (n() - sum(ifelse(dlogret < 0, dlogret, 0))) / 
                  (n() + sum(ifelse(dlogret > 0, dlogret, 0)))
    )

# Rank currencies by standard deviation (most common volatility measure)
ranked_data <- vol_cncy %>%
  mutate(rank = rank(-std_dev, na.last = "keep")) %>%  # Rank by standard deviation (descending)
    arrange(rank) %>%  # Sort by rank
    slice_head(n = 20)  # Keep top 20 most volatile currencies

kableExtra::kable(ranked_data, caption = "Top 20 Most Volatile Currencies (2016-2021)")

# DISCLAIMER: 
# While we calculate multiple volatility measures, standard deviation (std_dev) remains
# the most widely used measure in practice due to(S1 lecture notes):
# 1. Mathematical convenience and well-understood properties
# 2. Compatibility with normal distribution assumptions in many financial models
# 3. Use in common risk-adjusted performance measures like Sharpe ratio
# 4. Historical precedent and industry standardization
#
# However, investors should consider that standard deviation treats positive and 
# negative returns equally, while most investors are primarily concerned with 
# downside risk. The additional measures provided (MAD, downside_dev, d-ratio)
# offer complementary perspectives on risk characteristics.
```






```{r}
# I now want to run a for loop to get the average garch sigma over the same period for all the countries and then rank them again and see if noise is influencing the ranking

#I really tried to get it to work but i couldnt manage

```


```{r}
#Lets plot the return type persistence

#Get the returns and ensure they are in xts
zar_rts <- cncy %>%  
    filter(date > ymd(20150101)) %>% 
    group_by(Name) %>%  
    mutate(ret = Price/lag(Price)-1) %>% 
     filter(date > dplyr::first(date)) %>%               # drop first date per currency
  ungroup() %>%  
    mutate(Name= gsub("_Cncy", "", Name)) %>% 
   filter(Name == "SouthAfrica") %>% 
     filter(!is.na(ret)) %>% 
    select(-Name, -Price) %>% 
    tbl_xts()

#Following the practical for ease of syntax
Plotdata = cbind(zar_rts, zar_rts^2, abs(zar_rts))
colnames(Plotdata) = c("Returns", "Returns_Sqd", "Returns_Abs")

Plotdata <- 
Plotdata %>% xts_tbl() %>% 
gather(ReturnType, Returns, -date)

ggplot(Plotdata) + 
geom_line(aes(x = date, y = Returns, colour = ReturnType, alpha = 0.5)) + 
    
ggtitle("Return Type Persistence: ZAR") + 
facet_wrap(~ReturnType, nrow = 3, ncol = 1, scales = "free") + 
    
guides(alpha = "none", colour = "none") + 
fmxdat::theme_fmx()

```

```{r}
forecast::Acf(zar_rts, main = "ACF: ZAR")
```

```{r}
forecast::Acf(zar_rts^2, main = "ACF: Squared ZAR")
```
```{r}
forecast::Acf(abs(zar_rts), main = "ACF: Absolute ZAR")
```

```{r}
Box.test(coredata(zar_rts^2), type = "Ljung-Box", lag = 12)
```

```{r}
# Now we can actually fit the univariate garch model
cncy_rts_xts <- cncy %>%  
    filter(date > ymd(20150101)) %>% 
    group_by(Name) %>%  
    mutate(ret = Price/lag(Price)-1) %>% 
     filter(date > dplyr::first(date)) %>%               # drop first date per currency
  ungroup() %>% 
    mutate(Name= gsub("_Cncy", "", Name)) %>% 
   filter(Name == "SouthAfrica") %>% 
      filter(!is.na(ret)) %>% 
    select(-Name, -Price) %>% 
    tbl_xts()


#now i follow the practical as to fit the model

garch11 <- 
  
  ugarchspec(
    
    variance.model = list(model = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[1], 
                          
    garchOrder = c(1, 1)), 
    
    mean.model = list(armaOrder = c(1, 0), include.mean = TRUE), 
    
    distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[1])

# Now to fit, I use as.matrix and the data - this way the plot functions we will use later will work.

garchfit1 = ugarchfit(spec = garch11,data = cncy_rts_xts) 
sigma <- sigma(garchfit1) %>% xts_tbl() 
```

```{r}
garch_tab <- garchfit1@fit$matcoef
kableExtra::kable(garch_tab)
```

```{r}
sigma <- sigma(garchfit1) %>% xts_tbl() 
colnames(sigma) <- c("date", "sigma") 
sigma <- sigma %>% mutate(date = as.Date(date))

gg <- 
  
ggplot() + 
  geom_line(data = Plotdata %>% filter(ReturnType == "Returns_Sqd") %>% select(date, Returns) %>% 
              
              unique %>% mutate(Returns = sqrt(Returns)), aes(x = date, y = Returns)) + 
  
  geom_line(data = sigma, aes(x = date, y = sigma), color = "red", size = 2, alpha = 0.8) + 
  
  # scale_y_continuous(limits = c(0, 0.35)) + 
  labs(title = "Comparison: Returns Sigma vs Sigma from Garch") + 
  
    fmxdat::theme_fmx()
```

```{r}
fmxdat::finplot(gg, y.pct = T, y.pct_acc = 1)
```


GO-GARCH

```{r}
#Now I use the bbdxy to compare the volatility of the rand to global currencies

g10_rts <- bbdxy %>% 
    mutate(G10 = log(Price)-log(lag(Price))) %>% 
   filter(date > dplyr::first(date)) %>%               # drop first date per currency
    select(date, G10)

zar_log_rts <- cncy_rts %>% 
    filter(Name == "SouthAfrica") %>% 
    rename("ZAR" = "dlogret") %>% 
    select(date, ZAR) 



xts_rtn<- left_join(g10_rts,zar_log_rts, by=  "date") %>% tbl_xts()
library(rmgarch)

```

```{r}
#Now set the specifications for the go garch
```{r}

```

```{r}
renamingdcc <- function(ReturnSeries, DCC.TV.Cor) {
  # ReturnSeries: xts matrix of returns
  # DCC.TV.Cor:   T x (N^2) matrix after reshaping the rcor output
  
  dates <- zoo::index(ReturnSeries)
  N     <- ncol(ReturnSeries)
  names <- colnames(ReturnSeries)
  
  # Build pair names in the same order as vectorised correlations
  pairs      <- expand.grid(names, names)
  pair_names <- paste0(pairs$Var1, "_", pairs$Var2)
  
  df <- as.data.frame(DCC.TV.Cor)
  colnames(df) <- pair_names
  df$date      <- dates
  
  df %>% 
    tidyr::pivot_longer(
      cols      = -date,
      names_to  = "Pairs",
      values_to = "Rho"
    ) %>% 
    dplyr::arrange(date)
}

# ------------------------------------------------------------------------------
# GO GARCH specification and estimation
# ------------------------------------------------------------------------------

# Univariate GARCH spec for each series
uspec <- ugarchspec(
  variance.model = list(
    model      = "gjrGARCH",
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = "sstd"
)

# Replicate univariate spec across all return series
multi_univ_garch_spec <- multispec(replicate(ncol(xts_rtn), uspec))

# GO GARCH spec (multivariate)
spec.go <- gogarchspec(
  multi_univ_garch_spec,
  distribution.model = "mvnorm",   # or "manig"
  ica                = "fastica"   # fastICA
)

# Parallel cluster for speed
cl <- makePSOCKcluster(10)

# Fit univariate models jointly
multf <- multifit(multi_univ_garch_spec, xts_rtn, cluster = cl)

# Fit GO GARCH model
fit.gogarch <- gogarchfit(
  spec    = spec.go,
  data    = xts_rtn,
  solver  = "hybrid",
  cluster = cl,
  gfun    = "tanh",
  maxiter1 = 40000,
  epsilon  = 1e-08,
  rseed    = 100
)

stopCluster(cl)

# ------------------------------------------------------------------------------
# Extract and tidy time varying correlations
# ------------------------------------------------------------------------------

gog.time.var.cor <- rcor(fit.gogarch)               # array: N x N x T
gog.time.var.cor <- aperm(gog.time.var.cor, c(3, 2, 1))
dim(gog.time.var.cor) <- c(
  nrow(gog.time.var.cor),
  ncol(gog.time.var.cor)^2
)

gog.time.var.cor <- renamingdcc(
  ReturnSeries = xts_rtn,
  DCC.TV.Cor   = gog.time.var.cor
)



```{r}
g2 <- ggplot(gog.time.var.cor %>% filter(grepl("ZAR_", Pairs), 
    !grepl("_ZAR", Pairs))) + geom_line(aes(x = date, y = Rho, 
    colour = Pairs)) + fmxdat::theme_fmx() + ggtitle("Go-GARCH: ZAR")

print(g2)

```

Lastly I plot the GO-GARCH’s correlation between G10 currencies and the Rand. This graph now provides more evidence for how volatile the Rand is with the correleations over just 3 years ranging between 0.7 and 0.2.

```{r}
detach("package:rmgarch", unload=TRUE)
```


 QUESTION 5 
  
  ## 0. Libraries
  
```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")

```



---
  
  ## 1. Load data and build monthly return panel
  
  This uses monthly data from 2011 onwards and drops assets with less than five calendar years of data.

```{r}
# Load and filter MAA data
MAA <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI data
msci <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine benchmarks into a single tidy panel
df_full <- bind_rows(
  msci %>% rename(Tickers = Name),
  MAA  %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)

# Monthly data after 2010
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       # data after 2010
  group_by(Tickers) %>%
  # Keep only assets with at least five distinct calendar years of data
  filter(n_distinct(year(date)) >= 5) %>%
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  # Use last observation in each month
  filter(date == last(date)) %>%
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   # drop first NA return in each series

# Wide return matrix (monthly)
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

---
  
  ## 2. Single period optimiser: `optim_foo()`
  
  This function

* takes a look back window in months up to a given rebalance date

* keeps only assets with at least five years of data in that window

* builds Sigma and mu using `RiskPortfolios::covEstimation()` and geometric means

* maps tickers to asset classes according to the MAA table

* bonds and credit: government rates plus corporate credit
* commodity: BCOMTR
* equity: all MSCI indices

* builds `Amat` and `bvec` with group constraints that match the exam rules

* solves with `quadprog::solve.QP()` and returns weights as a tibble.

```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,     # at least ten years
                      LB        = 0.01,
                      UB        = 0.35,          # single asset at most 35%
                      bond_UB   = 0.25,          # bonds and credit at most 25%
                      eq_UB     = 0.60,          # equity at most 60%
                      com_UB    = 0.15) {        # commodities at most 15%
  
  # 1. Select look back window (assumes one row per month)
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)
  
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  # 2. Drop date and enforce at least five years of data per asset
  X <- data.matrix(window_df[, -1])
  
  # require at least 60 non NA monthly observations (about five years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. Estimate covariance (Sigma) and geometric mean returns (Mu)
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)
  
  # 4. Asset class mapping for constraints
  # Based on the MAA image -table ( provided in the data set):
  # - LGAGTRUH, LUAGTRUU, LEATTREU: rates (government bonds)
  # - LGCPTREH, LUACTRUU, LP05TREH: credit (corporate bonds)
  # All of these form the "bonds and credit instruments" bucket.
  # - BCOMTR: commodity
  # - MSCI indices: equity
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. Base constraints (long only, fully invested, single asset caps)
  #    Amat^T w >= bvec
  Amat <- cbind(
    rep(1, N),         # sum w = 1
    diag(N),           # w_i >= LB
    -diag(N)           # -w_i >= -UB  gives w_i <= UB
  )
  
  bvec <- c(
    1,                 # fully invested
    rep(LB, N),        # lower bound
    -rep(UB, N)        # upper bound
  )
  
  # 6. Group constraints: sum over group of w_i <= cap
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # rows correspond to assets
    g_vec[idx] <- -1              # -sum_{i in group} w_i >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  
  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # first constraint (sum w = 1) is equality
  
  # 7. Solve quadratic program:
  #    min (1/2) w' Sigma w - mu' w   subject to constraints
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,
    dvec = Mu,
    Amat = Amat,
    bvec = bvec,
    meq  = meq
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

---
  
  ## 3. Quarterly rebalancing dates
  
  We work with quarter end month ends (January, April, July, October).

```{r}
# Quarter end dates from the monthly series
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  pull(date)
```

---
  
  ## 4. Rolling optimiser and results
  
```{r}
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 120 months equals ten years
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Ten year look back as required
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

kable(Result_roll %>% head(13))
```

---
  
  ## 5. Stacked bar chart of portfolio weights over time
  
```{r}
# Prepare xts object of weights at rebalance dates
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%
  .[endpoints(., "months")]

# Plot stacked weights
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund"
)

```

# Clustering 
Using the Practical 5 (https://www.fmx.nfkatzke.com/posts/2020-08-17-practical-5/), I cluster the indices on a Ledoit–Wolf shrunk correlation matrix to identify statistical comovement groups. The dendrogram suggests, for example, that [equity indices] cluster closely, while [bond/credit] and commodities sit in distinct branches.
```{r}
## 6. Cluster based comovement analysis (Practical 5)

# Use the same monthly return matrix, drop date
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Ledoit–Wolf covariance and correlation matrix (as in Practical 5)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")
)

corr_clust <- cov2cor(Sigma_clust)

# 6.2 Distance matrix using the Dower metric: d_ij = sqrt( (1 - rho_ij) / 2 )
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward linkage (AGNES)
library(cluster)

hc <- cluster::agnes(dist(distmat), method = "ward")

# Basic dendrogram (you could beautify with the ggdendro helpers from Practical 5)
plot(hc, which.plot = 2, main = "Hierarchical clustering of global indices",
     xlab = "", sub = "")

```

