---
title: "Financial Econometrics 871 Practical Exam"
output: github_document
editor_options: 
  chunk_output_type: console
---


**SU Number:** 22660348  
**Author:** Precious Nhamo  
**Course:** Financial Econometrics 871  
**Dates:** 18–19 November 2025  

Repo: <https://github.com/prexymtha/FMX_25_22660348>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo   = TRUE,   # show code
  message = FALSE,
  warning = FALSE
)
```





### Question 1 

I struggled with this question....

We can compute the rolling 3-year beta for each active manager and the ABC Fund against the benchmark.Then, we can create a scatterplot of returns vs. risk (beta) and overlay boxplots to show the distribution of betas for peers and highlight ABC Fund.


To see how much the fund differs from the rest of the industry in terms of what drives the return .We can perform a factor analysis or compare the R-squared from a market model to see how much of the returns are explained by the market.


Assess relative performance over time We can plot the cumulative returns of ABC Fund, peers, and the benchmark.

We can compute a measure of dispersion (cross-sectional standard deviation of returns) and then plot the fund's performance against this dispersion.




### Question 2

---
```{r load-data-Q2, echo = TRUE}
# Load readr using pacman
pacman::p_load(
  readr,
  tidyverse,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  rmsfuns,
  fmxdat,
  DescTools
)

# Load the data
Hold_Rets <- read_rds("data/Hold_Rets_Sectors.rds") 

head(Hold_Rets)
```

To answer the first part of small stocks being highly volatile compared to larger stocks I will annualise the returns for the decade as requested by the question - I will also categorise the returns with their groups as either being small_cap or large_cap


In the lecturer’s notes, volatility (risk) is explicitly defined and measured using the standard deviation of returns, and we are told that:

Standard deviation should be annualised when we compare different time periods.

The usual shortcut SD * sqrt(12) is only theoretically correct when we work with logarithmic returns, because annual log return is the sum of monthly log returns, which justifies the square root of time rule.

So, to be consistent with the notes and to get a clean, comparable volatility measure, I:

Construct value weighted indices for small caps and large caps.

Convert their monthly simple returns to log returns using log(1 + ret).

Compute a 36 month rolling standard deviation of these log monthly returns.

Annualise that rolling SD by multiplying by sqrt(12).

This gives me a rolling annualised standard deviation of returns for small caps and large caps over the past decade. Comparing these two series over time is a direct, textbook implementation of the lecturer’s definition of volatility, so it is an appropriate way to answer 2(i): if the small cap series has a consistently higher annualised SD, we can conclude that smaller stocks have been more volatile than larger stocks locally.

```{r}
options(dplyr.summarise.inform = FALSE)

library(tidyverse)
library(lubridate)
library(tbl2xts)
library(PerformanceAnalytics)
library(RcppRoll)
library(fmxdat)   # for safe_month_min, theme_fmx, fmx_cols, fmx_fills etc

# Start from your data
# Hold_Rets has: date, Tickers, Return, Weight, Group, Sector

Hold_Rets_decade <- 
  Hold_Rets %>% 
  filter(Group %in% c("Small_Caps", "Large_Caps")) %>% 
  mutate(year = year(date)) %>% 
  filter(year >= max(year) - 9) %>%   # last 10 calendar years
  select(-year)


```


```{r}
# 1a. Daily value-weighted size index returns (Small vs Large)
idx_daily <- 
  Hold_Rets_decade %>% 
  group_by(date, Group) %>% 
  summarise(
    ret = sum(Weight * Return, na.rm = TRUE) / sum(Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(Tickers = Group) %>%      # so Tickers is "Small_Caps" / "Large_Caps"
  arrange(Tickers, date)

# 1b. Make returns monthly for illustration (Nico style)
idx <- 
  idx_daily %>% 
  mutate(YM = format(date, "%Y%B")) %>% 
  arrange(date) %>% 
  group_by(Tickers, YM) %>% 
  # pick last date of the month and compute monthly return
  summarise(
    date = last(date),
    ret  = prod(1 + ret, na.rm = TRUE) - 1,
    .groups = "drop"
  ) %>% 
  arrange(Tickers, date)

head(idx)

```


```{r}
#======================
# Manual Calculation:
#======================

dfplot <- 
  bind_rows(
    # 6 months – do NOT annualise (same as Nico)
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% 
      mutate(Freq = "A"),
    
    # 1 year
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 12) - 1) %>% 
      mutate(Freq = "B"),
    
    # 3 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 36) - 1) %>% 
      mutate(Freq = "C"),
    
    # 5 years
    idx %>% 
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>% 
      group_by(Tickers) %>% 
      summarise(mu = prod(1 + ret, na.rm = TRUE) ^ (12 / 60) - 1) %>% 
      mutate(Freq = "D")
  )

#======================
# PerformanceAnalytics check
#======================

idxxts <- 
  tbl_xts(idx, cols_to_xts = ret, spread_by = Tickers)

dfplotxts <- 
  bind_rows(
    idxxts %>% tail(6)  %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "A"),
    idxxts %>% tail(12) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "B"),
    idxxts %>% tail(36) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "C"),
    idxxts %>% tail(60) %>% Return.annualized(scale = 12) %>% data.frame() %>% mutate(Freq = "D")
  ) %>% 
  data.frame() %>% 
  tidyr::gather(Tickers, mu, -Freq) %>% 
  mutate(Tickers = gsub("\\.", " ", Tickers))

# Facet labels
to_string <- as_labeller(c(
  `A` = "6 Months", 
  `B` = "1 Year", 
  `C` = "3 Years", 
  `D` = "5 Years"
))

# Barplot_foo (same style)
g_ann <- 
  dfplot %>% 
  ggplot() + 
  geom_bar(aes(Tickers, mu, fill = Tickers), stat = "identity") + 
  facet_wrap(~Freq, labeller = to_string, nrow = 1) + 
  labs(
    x = "",
    y = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualized terms."
  ) + 
  fmx_fills() + 
  geom_label(
    aes(Tickers, mu, label = paste0(round(mu, 4) * 100, "%")),
    size = ggpts(8),
    alpha = 0.35,
    fontface = "bold",
    nudge_y = 0.002
  ) + 
  theme_fmx(
    CustomCaption = TRUE,
    title.size = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size = ggpts(30),
    axis.size = ggpts(37),
    legend.size = ggpts(35),
    legend.pos = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

```

```{r}
gg_cum <- 
  idx %>% 
  arrange(date) %>% 
  group_by(Tickers) %>% 
  mutate(
    Rets = coalesce(ret, 0),
    CP   = cumprod(1 + Rets)
  ) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(date, CP, colour = Tickers)) +
  labs(
    title   = "Illustration of Cumulative Returns of Large vs Small Caps",
    subtitle = "",
    caption = "Note:\nDistortions emerge as starting dates differ."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  )

# Level plot
gg_cum

# Log cumulative plot
gg_cum_log <- 
  gg_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(gg_cum$labels$title, "\nLog Scaled"),
    y     = "Log Scaled Cumulative Returns"
  )

gg_cum_log

```

```{r}
plotdf <- 
  idx %>% 
  group_by(Tickers) %>% 
  mutate(
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right") ^ (12 / 36) - 1
  ) %>% 
  group_by(date) %>% 
  filter(any(!is.na(RollRets))) %>% 
  ungroup()

g_roll <- 
  plotdf %>% 
  ggplot() +
  geom_line(aes(date, RollRets, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized Returns: Large vs Small Caps",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year Returns (Ann.)",
    caption = "Note:\nDistortions are not evident now."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_roll, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = TRUE, y.pct_acc = 1)

```


# This is the answer to the volatility question 
```{r}
plot_dlog <- 
  idx %>% 
  group_by(Tickers) %>% 
  # log monthly returns:
  mutate(ret = log(1 + ret)) %>% 
  # Rolling SD annualized calc now:
  mutate(
    RollSD = RcppRoll::roll_sd(ret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>% 
  filter(!is.na(RollSD))

g_sd <- 
  plot_dlog %>% 
  ggplot() +
  geom_line(aes(date, RollSD, colour = Tickers), alpha = 0.7, size = 1.25) +
  labs(
    title   = "Rolling 3 Year Annualized SD: Large vs Small Caps",
    subtitle = "",
    x       = "",
    y       = "Rolling 3 year SD (Ann.)",
    caption = "Note:\nAnnualized SD computed on log monthly returns."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(g_sd, x.date.dist = "1 year", x.date.type = "%Y", x.vert = TRUE,
        y.pct = FALSE)
```


Second part (peak and troughs )

We need three graphs that have returns ( of whatever format on the y axis) and year on the x axis since we have already annualised it's easy graph one needs to have a legend by size group ( small caps , large , medium and other ) , the second by sector and the last by size and sector . Group has NA which aren't necessarily missing values but rather unclassified so we shall call them that to remove confusion 

start with the one for return against size groups 



### Question 3 

### Question 4: Volatility and GARCH estimates

#### 5.4.1 Aim of the question

- We are assessing whether the South African Rand (ZAR) has been one of the most volatile currency over the past few years.(since it says one of the most we assess it relative to other currencies)
- ZAR has generally performed well during periods where the G10 currency carry trades have been favourable and currency valuations cheap ( we can split this for two and then do the joint )
- Globally, it has been one of the currencies that most benefit during periods where the Dollar is comparatively strong,
 indicating a risk-on sentiment
- Use your discretion on what you use to measure volatility, which currencies you compare
 and how you arrive at your conclusions. Be creative in using tables, graphs, stratification and
 statistics to argue your points.


- Use your discretion on what you use to measure volatility, which currencies you compare
 and how you arrive at your conclusions. Be creative in using tables, graphs, stratification and
 statistics to argue your points.



### Q4: Volatility and GARCH estimates for ZAR

#### 0. Libraries and data


```{r Q4-libs-data, echo = TRUE}
pacman::p_load(
  tidyverse,
  devtools,
  rugarch,
  rmgarch,
  forecast,
  tbl2xts,
  lubridate,
  PerformanceAnalytics,
  ggthemes,
  robustbase,
  kableExtra,
  RcppRoll
)

# Currency data (all quoted vs USD)
cncy       <- read_rds("data/currencies.rds")
cncy_Carry <- read_rds("data/cncy_Carry.rds")   # G10 carry index
cncy_value <- read_rds("data/cncy_value.rds")   # FX PPP value index
cncyIV     <- read_rds("data/cncyIV.rds")       # FX implied vol index
bbdxy      <- read_rds("data/bbdxy.rds")        # Bloomberg Dollar spot index

```



### 1. Cross sectional volatility: ZAR vs G10 vs others

```{r Q4-vol-panel, echo = TRUE}
# Construct daily log returns and scaled returns for all currencies
cncy_rts <- cncy %>%  
  group_by(Name) %>%  
  arrange(date, .by_group = TRUE) %>% 
  mutate(
    dlogret   = log(Price) - log(lag(Price)),
    scaledret = dlogret - mean(dlogret, na.rm = TRUE)
  ) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>% 
  mutate(Name = gsub("_Cncy", "", Name))

# Define G10 set in the cleaned Name space
g10_vec <- c(
  "Australia_Inv",  # AUD
  "Canada",         # CAD
  "EU_Inv",         # EUR
  "Japan",          # JPY
  "NZ_Inv",         # NZD
  "Norway",         # NOK
  "Sweden",         # SEK
  "UK_Inv"          # GBP
  # USD is the base here
)

# Volatility measures since 2015
vol_cncy <- cncy_rts %>% 
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%
  summarise(
    std_dev      = sd(dlogret, na.rm = TRUE),
    mad          = mean(abs(dlogret - mean(dlogret, na.rm = TRUE)), na.rm = TRUE),
    downside_dev = sqrt(mean(ifelse(dlogret < 0, dlogret^2, 0), na.rm = TRUE)),
    d_ratio      = (n() - sum(ifelse(dlogret < 0, dlogret, 0))) / 
                   (n() + sum(ifelse(dlogret > 0, dlogret, 0))),
    .groups = "drop"
  ) %>% 
  mutate(
    Group = case_when(
      Name == "SouthAfrica" ~ "ZAR",
      Name %in% g10_vec     ~ "G10",
      TRUE                  ~ "Other"
    )
  )

# Rank by standard deviation
ranked_data <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number()) %>% 
  select(rank, Name, Group, everything())

kableExtra::kable(
  ranked_data,
  caption = "Volatility ranking of currencies (log returns, since 2015)",
  digits = 6
)
```



```{r Q4-vol-plot, echo = TRUE}
# Identify top 10 by volatility
vol_with_rank <- vol_cncy %>% 
  arrange(desc(std_dev)) %>% 
  mutate(rank = row_number())

# Keep: top 10 *or* any G10 currency
plot_set <- vol_with_rank %>% 
  filter(rank <= 10 | Group == "G10")

plot_set %>% 
  mutate(Name = fct_reorder(Name, std_dev)) %>% 
  ggplot() +
  geom_col(aes(x = Name, y = std_dev, fill = Group), alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "ZAR"   = "red",
      "G10"   = "grey40",
      "Other" = "grey80"
    )
  ) +
  coord_flip() +
  labs(
    title    = "Cross sectional volatility of currencies",
    subtitle = "Top 10 currencies by volatility plus G10 majors (since 2015)",
    x = "",
    y = "Standard deviation",
    fill = ""
  ) +
  fmxdat::theme_fmx()

```

---


```{r Q4-zar-persistence, echo = TRUE}
# Simple returns for ZAR only
zar_rts <- cncy %>%  
  filter(date > ymd(20150101)) %>% 
  group_by(Name) %>%  
  mutate(ret = Price / lag(Price) - 1) %>% 
  filter(date > dplyr::first(date)) %>% 
  ungroup() %>%  
  mutate(Name = gsub("_Cncy", "", Name)) %>% 
  filter(Name == "SouthAfrica") %>% 
  filter(!is.na(ret)) %>% 
  select(-Name, -Price) %>% 
  tbl_xts()

# Construct the three series: returns, squared returns, absolute returns
Plotdata <- cbind(zar_rts, zar_rts^2, abs(zar_rts))
colnames(Plotdata) <- c("Returns", "Returns_Sqd", "Returns_Abs")

Plotdata <- Plotdata %>% 
  xts_tbl() %>% 
  gather(ReturnType, Returns, -date)

ggplot(Plotdata) + 
  geom_line(aes(x = date, y = Returns, colour = ReturnType, alpha = 0.5)) + 
  ggtitle("Return Type Persistence: ZAR") + 
  facet_wrap(~ReturnType, nrow = 3, ncol = 1, scales = "free") + 
  guides(alpha = "none", colour = "none") + 
  fmxdat::theme_fmx()
```


```{r Q4-zar-acf, echo = TRUE}
forecast::Acf(zar_rts,   main = "ACF: ZAR return")
forecast::Acf(zar_rts^2, main = "ACF: Squared ZAR return")
forecast::Acf(abs(zar_rts), main = "ACF: Absolute ZAR return")

Box.test(coredata(zar_rts^2), type = "Ljung-Box", lag = 12)
```


---

### 3. Fitting GARCH(1,1) to ZAR (rugarch, copy of Practical 6)

```{r Q4-zar-garch-fit, echo = TRUE}
# Use the same object name as Nico: porteqw now holds ZAR returns
porteqw <- zar_rts

# GARCH(1,1) specification as in the notes
garch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[1],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[1]
)

garchfit1 <- ugarchfit(spec = garch11, data = porteqw)

garchfit1
```


```{r Q4-zar-garch-table, echo = TRUE}
garch_tab <- garchfit1@fit$matcoef
kableExtra::kable(garch_tab, caption = "GARCH(1,1) estimates for ZAR")
```



Conditional variance plot, identical structure:


```{r Q4-zar-sigma-plot, echo = TRUE}
sigma <- sigma(garchfit1) %>% xts_tbl() 
colnames(sigma) <- c("date", "sigma") 
sigma <- sigma %>% mutate(date = as.Date(date))

gg <- ggplot() + 
  geom_line(
    data = Plotdata %>% 
      filter(ReturnType == "Returns_Sqd") %>% 
      select(date, Returns) %>% 
      unique() %>% 
      mutate(Returns = sqrt(Returns)),
    aes(x = date, y = Returns)
  ) + 
  geom_line(
    data = sigma,
    aes(x = date, y = sigma),
    color = "red", size = 2, alpha = 0.8
  ) + 
  labs(
    title    = "Comparison: Returns sigma vs sigma from GARCH",
    subtitle = "ZAR: note the smoothing effect of GARCH as noise is controlled for.",
    x = "", y = "Estimated volatility",
    caption = "Source: Financial econometrics practical | Calculations: Own"
  ) + 
  fmxdat::theme_fmx(CustomCaption = TRUE)

fmxdat::finplot(gg, y.pct = TRUE, y.pct_acc = 1)

```


News impact curve for the ZAR GARCH model:


```{r Q4-zar-nic, echo = TRUE}
ni <- newsimpact(z = NULL, garchfit1)

plot(
  ni$zx, ni$zy,
  ylab = ni$yexpr,
  xlab = ni$xexpr,
  type = "l",
  main = "News Impact Curve: ZAR sGARCH(1,1)"
)

```



### 4. Alternative GARCH forms for ZAR (GJR and EGARCH) and NIC comparison

```{r Q4-zar-gjr-egarch, echo = TRUE}
# GJR GARCH with student t
gjrgarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[2],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit2 <- ugarchfit(spec = gjrgarch11, data = as.matrix(porteqw))

kableExtra::kable(
  garchfit2@fit$matcoef,
  caption = "GJR GARCH(1,1) with student t distribution for ZAR"
)

# EGARCH(1,1) with student t
egarch11 <- ugarchspec(
  variance.model = list(
    model      = c("sGARCH","gjrGARCH","eGARCH","fGARCH","apARCH")[3],
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = c("norm", "snorm", "std", "sstd", "ged", "sged", "nig", "ghyp", "jsu")[3]
)

garchfit3 <- ugarchfit(spec = egarch11, data = as.matrix(porteqw))

# Sign bias and info criteria, as in Nico’s notes
signbias(garchfit1)
infocriteria(garchfit1)
infocriteria(garchfit2)
infocriteria(garchfit3)

```



News impact curves for all three, identical helper function:


```{r Q4-zar-nic-compare, echo = TRUE}
NICurveMaker <- function(Fit, Name) {
  NI <- newsimpact(z = NULL, Fit)
  NI <- cbind(NI$zx, NI$zy)
  colnames(NI) <- c(paste0("Epsilon_", Name), paste0("Sigma_", Name))
  NI %>% data.frame() %>% tibble::as_tibble()
}

NI1 <- NICurveMaker(garchfit1, "GARCH11")
NI2 <- NICurveMaker(garchfit2, "GJR")
NI3 <- NICurveMaker(garchfit3, "EGARCH")

NI <- cbind(NI1, NI2, NI3) %>% 
  gather(Model, Sigma, starts_with("Sigma")) %>% 
  rename(Epsilon = Epsilon_GARCH11) %>% 
  select(-Epsilon_GJR, -Epsilon_EGARCH)

ggplot(NI) + 
  geom_line(aes(x = Epsilon, y = Sigma, colour = Model)) + 
  ggtitle("News Impact Curves: ZAR") + 
  theme_hc()

```


---

### 5. GO GARCH and time varying correlation between ZAR and Dollar strength

Now we bring in the multivariate flavour, matching Practical 7 style, but on a simple two series system: ZAR and the Dollar index (bbdxy). This speaks directly to the “benefits when the Dollar is strong” statement.


```{r Q4-zar-go-garch-data, echo = TRUE}
# Dollar index log returns
g10_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(G10 = log(Price) - log(lag(Price))) %>% 
  filter(date > dplyr::first(date)) %>% 
  select(date, G10)

# ZAR log returns (from earlier cncy_rts object)
zar_log_rts <- cncy_rts %>% 
  filter(Name == "SouthAfrica") %>% 
  rename(ZAR = dlogret) %>% 
  select(date, ZAR)

# Merge and convert to xts
xts_rtn <- left_join(g10_rts, zar_log_rts, by = "date") %>% 
  drop_na() %>% 
  tbl_xts()

```



Helper for renaming pairwise correlations (same as Practical 7):


```{r Q4-renamingdcc, echo = TRUE}
renamingdcc <- function(ReturnSeries, DCC.TV.Cor) {
  dates <- zoo::index(ReturnSeries)
  N     <- ncol(ReturnSeries)
  names <- colnames(ReturnSeries)
  
  pairs      <- expand.grid(names, names)
  pair_names <- paste0(pairs$Var1, "_", pairs$Var2)
  
  df <- as.data.frame(DCC.TV.Cor)
  colnames(df) <- pair_names
  df$date      <- dates
  
  df %>% 
    tidyr::pivot_longer(
      cols      = -date,
      names_to  = "Pairs",
      values_to = "Rho"
    ) %>% 
    dplyr::arrange(date)
}

```



GO GARCH specification and fit, directly from Nico’s multivariate notes:

```{r Q4-zar-go-garch-fit, echo = TRUE}
# Univariate GJR GARCH spec for each series
uspec <- ugarchspec(
  variance.model = list(
    model      = "gjrGARCH",
    garchOrder = c(1, 1)
  ),
  mean.model = list(
    armaOrder    = c(1, 0),
    include.mean = TRUE
  ),
  distribution.model = "sstd"
)

multi_univ_garch_spec <- multispec(replicate(ncol(xts_rtn), uspec))

spec.go <- gogarchspec(
  multi_univ_garch_spec,
  distribution.model = "mvnorm",
  ica                = "fastica"
)

cl <- makePSOCKcluster(10)

multf <- multifit(multi_univ_garch_spec, xts_rtn, cluster = cl)

fit.gogarch <- gogarchfit(
  spec    = spec.go,
  data    = xts_rtn,
  solver  = "hybrid",
  cluster = cl,
  gfun    = "tanh",
  maxiter1 = 40000,
  epsilon  = 1e-08,
  rseed    = 100
)

stopCluster(cl)

print(fit.gogarch)

```


Time varying correlations, same gymnastics as in the notes:


```{r Q4-zar-go-garch-tv-cor, echo = TRUE}
gog.time.var.cor <- rcor(fit.gogarch)
gog.time.var.cor <- aperm(gog.time.var.cor, c(3, 2, 1))
dim(gog.time.var.cor) <- c(
  nrow(gog.time.var.cor),
  ncol(gog.time.var.cor)^2
)

gog.time.var.cor <- renamingdcc(
  ReturnSeries = xts_rtn,
  DCC.TV.Cor   = gog.time.var.cor
)

g2 <- ggplot(
  gog.time.var.cor %>% 
    filter(grepl("ZAR_", Pairs), !grepl("_ZAR", Pairs))
) +
  geom_line(aes(x = date, y = Rho, colour = Pairs)) +
  theme_hc() +
  ggtitle("Go GARCH: time varying correlation between ZAR and Dollar index")

print(g2)

```


---

### 6. Linking to G10 carry and PPP value (for narrative)

For the second statement in the question, you can add very short extra chunks to relate ZAR returns to:

- G10 carry index `cncy_Carry`  
- PPP value index `cncy_value`  
- FX implied vol `cncyIV`  

For example, simple correlations over the sample:


```{r Q4-zar-carry-value-links, echo = TRUE}
# Simple daily log returns for ZAR, carry, value, and Dollar index
zar_daily <- cncy_rts %>% filter(Name == "SouthAfrica") %>% select(date, ZAR = dlogret)

carry_rts <- cncy_Carry %>% 
  arrange(date) %>% 
  mutate(Carry = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Carry)) %>% 
  select(date, Carry)

value_rts <- cncy_value %>% 
  arrange(date) %>% 
  mutate(Value = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(Value)) %>% 
  select(date, Value)

dxy_rts <- bbdxy %>% 
  arrange(date) %>% 
  mutate(DXY = log(Price) - log(lag(Price))) %>% 
  filter(!is.na(DXY)) %>% 
  select(date, DXY)

link_df <- zar_daily %>% 
  inner_join(carry_rts, by = "date") %>% 
  inner_join(value_rts, by = "date") %>% 
  inner_join(dxy_rts,   by = "date")

round(cor(link_df %>% select(-date), use = "complete.obs"), 3)

```


 QUESTION 5 
  
  ## 0. Libraries
  
```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")

```



---
  
  ## 1. Load data and build monthly return panel
  
  This uses monthly data from 2011 onwards and drops assets with less than five calendar years of data.

```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")
library(readr)
library(dplyr)
# Load and filter MAA data
MAA <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI data
msci <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine benchmarks into a single tidy panel
df_full <- bind_rows(
  msci %>% rename(Tickers = Name),
  MAA  %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)

# Monthly data after 2010
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       # data after 2010
  group_by(Tickers) %>%
  # Keep only assets with at least five distinct calendar years of data
  filter(n_distinct(year(date)) >= 5) %>%
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  # Use last observation in each month
  filter(date == dplyr::last(date)) %>%
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   # drop first NA return in each series

# Wide return matrix (monthly)
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

---
  
  ## 2. Single period optimiser: `optim_foo()`
  
  This function

* takes a look back window in months up to a given rebalance date

* keeps only assets with at least five years of data in that window

* builds Sigma and mu using `RiskPortfolios::covEstimation()` and geometric means

* maps tickers to asset classes according to the MAA table

* bonds and credit: government rates plus corporate credit
* commodity: BCOMTR
* equity: all MSCI indices

* builds `Amat` and `bvec` with group constraints that match the exam rules

* solves with `quadprog::solve.QP()` and returns weights as a tibble.

```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,     # at least ten years
                      LB        = 0.01,
                      UB        = 0.35,          # single asset at most 35%
                      bond_UB   = 0.25,          # bonds and credit at most 25%
                      eq_UB     = 0.60,          # equity at most 60%
                      com_UB    = 0.15) {        # commodities at most 15%
  
  # 1. Select look back window (assumes one row per month)
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)
  
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  # 2. Drop date and enforce at least five years of data per asset
  X <- data.matrix(window_df[, -1])
  
  # require at least 60 non NA monthly observations (about five years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. Estimate covariance (Sigma) and geometric mean returns (Mu)
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)
  
  # 4. Asset class mapping for constraints
  # Based on the MAA image -table ( provided in the data set):
  # - LGAGTRUH, LUAGTRUU, LEATTREU: rates (government bonds)
  # - LGCPTREH, LUACTRUU, LP05TREH: credit (corporate bonds)
  # All of these form the "bonds and credit instruments" bucket.
  # - BCOMTR: commodity
  # - MSCI indices: equity
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. Base constraints (long only, fully invested, single asset caps)
  #    Amat^T w >= bvec
  Amat <- cbind(
    rep(1, N),         # sum w = 1
    diag(N),           # w_i >= LB
    -diag(N)           # -w_i >= -UB  gives w_i <= UB
  )
  
  bvec <- c(
    1,                 # fully invested
    rep(LB, N),        # lower bound
    -rep(UB, N)        # upper bound
  )
  
  # 6. Group constraints: sum over group of w_i <= cap
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # rows correspond to assets
    g_vec[idx] <- -1              # -sum_{i in group} w_i >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  
  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # first constraint (sum w = 1) is equality
  
  # 7. Solve quadratic program:
  #    min (1/2) w' Sigma w - mu' w   subject to constraints
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,
    dvec = Mu,
    Amat = Amat,
    bvec = bvec,
    meq  = meq
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

---
  
  ## 3. Quarterly rebalancing dates
  
  We work with quarter end month ends (January, April, July, October).

```{r}
# Quarter end dates from the monthly series
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  pull(date)
```

---
  
  ## 4. Rolling optimiser and results
  
```{r}
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 120 months equals ten years
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Ten year look back as required
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

kable(Result_roll %>% head(13))
```

---
  
  ## 5. Stacked bar chart of portfolio weights over time
  
```{r}
# Prepare xts object of weights at rebalance dates
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%
  .[endpoints(., "months")]

# Plot stacked weights
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund"
)

# StackBar of monthly weights (Note the stand-out rebalance weights...):
#Result_roll %>% bar_xts() %>% .[endpoints(.,'months')] %>% chart.StackedBar()

```

# Clustering 
Using the Practical 5 (https://www.fmx.nfkatzke.com/posts/2020-08-17-practical-5/), I cluster the indices on a Ledoit–Wolf shrunk correlation matrix to identify statistical comovement groups. The dendrogram suggests, for example, that [equity indices] cluster closely, while [bond/credit] and commodities sit in distinct branches.
```{r}
## 6. Cluster based comovement analysis (Practical 5)

# Use the same monthly return matrix, drop date
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Ledoit–Wolf covariance and correlation matrix (as in Practical 5)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")
)

corr_clust <- cov2cor(Sigma_clust)

# 6.2 Distance matrix using the Dower metric: d_ij = sqrt( (1 - rho_ij) / 2 )
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward linkage (AGNES)
library(cluster)

hc <- cluster::agnes(dist(distmat), method = "ward")

# Basic dendrogram (you could beautify with the ggdendro helpers from Practical 5)
plot(hc, which.plot = 2, main = "Hierarchical clustering of global indices",
     xlab = "", sub = "")

```



