---
title: "Question 5: Global Balanced Index Fund"
output: html_document
---


## 0. Libraries

```{r}
pacman::p_load(
  tidyverse,
  xts,
  tbl2xts,
  RiskPortfolios,
  PerformanceAnalytics,
  PortfolioAnalytics,
  lubridate,
  kableExtra,
  quadprog
)

devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")

```

---

## 1. Load data and build monthly return panel

This uses monthly data from 2011 onwards and drops assets with less than five calendar years of data.

```{r}
# Load and filter MAA data
MAA <- read_rds("data/MAA.rds") %>%
  filter(Ticker %in% c("LUACTRUU Index", "LUAGTRUU Index",
                       "BCOMTR Index", "LP05TREH Index"))

# Load and filter MSCI data
msci <- read_rds("data/msci.rds") %>%
  filter(Name %in% c("MSCI ACWI", "MSCI USA", "MSCI RE",
                     "MSCI Jap", "MSCI China"))

# Combine benchmarks into a single tidy panel
df_full <- bind_rows(
  msci %>% rename(Tickers = Name),
  MAA  %>% select(-Name) %>% rename(Tickers = Ticker)
) %>%
  arrange(date)

# Monthly data after 2010
monthly_df <- df_full %>%
  filter(date >= ymd(20110101)) %>%       # data after 2010
  group_by(Tickers) %>%
  # Keep only assets with at least five distinct calendar years of data
  filter(n_distinct(year(date)) >= 5) %>%
  ungroup() %>%
  mutate(YM = format(date, "%Y%m")) %>%
  arrange(date) %>%
  group_by(Tickers, YM) %>%
  # Use last observation in each month
  filter(date == last(date)) %>%
  group_by(Tickers) %>%
  mutate(ret = Price / lag(Price) - 1) %>%
  ungroup() %>%
  select(date, Tickers, ret) %>%
  filter(!is.na(ret))   # drop first NA return in each series

# Wide return matrix (monthly)
return_mat <- monthly_df %>%
  spread(Tickers, ret)

return_mat_Nodate <- data.matrix(return_mat[, -1])
```

---

## 2. Single period optimiser: `optim_foo()`

This function

* takes a look back window in months up to a given rebalance date

* keeps only assets with at least five years of data in that window

* builds Sigma and mu using `RiskPortfolios::covEstimation()` and geometric means

* maps tickers to asset classes according to the MAA table

  * bonds and credit: government rates plus corporate credit
  * commodity: BCOMTR
  * equity: all MSCI indices

* builds `Amat` and `bvec` with group constraints that match the exam rules

* solves with `quadprog::solve.QP()` and returns weights as a tibble.

```{r}
optim_foo <- function(return_mat,
                      end_date,
                      lookback_months = 120,     # at least ten years
                      LB        = 0.01,
                      UB        = 0.35,          # single asset at most 35%
                      bond_UB   = 0.25,          # bonds and credit at most 25%
                      eq_UB     = 0.60,          # equity at most 60%
                      com_UB    = 0.15) {        # commodities at most 15%
  
  # 1. Select look back window (assumes one row per month)
  window_df <- return_mat %>%
    filter(date <= end_date) %>%
    arrange(date) %>%
    slice_tail(n = lookback_months)
  
  if (nrow(window_df) < lookback_months) {
    warning("Not enough history for lookback at ", end_date)
    return(NULL)
  }
  
  # 2. Drop date and enforce at least five years of data per asset
  X <- data.matrix(window_df[, -1])
  
  # require at least 60 non NA monthly observations (about five years)
  valid_cols    <- colSums(!is.na(X)) >= 60
  X_valid       <- X[, valid_cols, drop = FALSE]
  assets_valid  <- colnames(X_valid)
  
  if (length(assets_valid) == 0) {
    warning("No assets with >= 5 years of data in window ending ", end_date)
    return(NULL)
  }
  
  # 3. Estimate covariance (Sigma) and geometric mean returns (Mu)
  Sigma <- RiskPortfolios::covEstimation(X_valid)
  
  Mu <- window_df %>%
    select(date, all_of(assets_valid)) %>%
    summarise(across(-date, ~ prod(1 + .)^(1 / n()) - 1)) %>%
    purrr::as_vector()
  
  N <- length(assets_valid)
  
  # 4. Asset class mapping for constraints
  # Based on the MAA image -table ( provided in the data set):
  # - LGAGTRUH, LUAGTRUU, LEATTREU: rates (government bonds)
  # - LGCPTREH, LUACTRUU, LP05TREH: credit (corporate bonds)
  # All of these form the "bonds and credit instruments" bucket.
  # - BCOMTR: commodity
  # - MSCI indices: equity
  asset_class <- case_when(
    assets_valid == "BCOMTR Index" ~ "Commodity",
    assets_valid %in% c(
      "LGAGTRUH Index", "LUAGTRUU Index", "LEATTREU Index",
      "LGCPTREH Index", "LUACTRUU Index", "LP05TREH Index"
    ) ~ "BondCredit",
    TRUE ~ "Equity"
  )
  
  bond_idx <- which(asset_class == "BondCredit")
  eq_idx   <- which(asset_class == "Equity")
  com_idx  <- which(asset_class == "Commodity")
  
  # 5. Base constraints (long only, fully invested, single asset caps)
  #    Amat^T w >= bvec
  Amat <- cbind(
    rep(1, N),         # sum w = 1
    diag(N),           # w_i >= LB
    -diag(N)           # -w_i >= -UB  gives w_i <= UB
  )
  
  bvec <- c(
    1,                 # fully invested
    rep(LB, N),        # lower bound
    -rep(UB, N)        # upper bound
  )
  
  # 6. Group constraints: sum over group of w_i <= cap
  add_group_constraint <- function(idx, cap, Amat, bvec) {
    if (length(idx) == 0) return(list(Amat = Amat, bvec = bvec))
    
    g_vec <- rep(0, nrow(Amat))   # rows correspond to assets
    g_vec[idx] <- -1              # -sum_{i in group} w_i >= -cap
    Amat_new <- cbind(Amat, g_vec)
    bvec_new <- c(bvec, -cap)
    list(Amat = Amat_new, bvec = bvec_new)
  }
  
  tmp <- add_group_constraint(bond_idx, bond_UB, Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(eq_idx,   eq_UB,   Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  tmp <- add_group_constraint(com_idx,  com_UB,  Amat, bvec)
  Amat <- tmp$Amat; bvec <- tmp$bvec
  
  meq <- 1  # first constraint (sum w = 1) is equality
  
  # 7. Solve quadratic program:
  #    min (1/2) w' Sigma w - mu' w   subject to constraints
  w.opt <- quadprog::solve.QP(
    Dmat = Sigma,
    dvec = Mu,
    Amat = Amat,
    bvec = bvec,
    meq  = meq
  )$solution
  
  tibble(
    date   = end_date,
    stocks = assets_valid,
    weight = as.numeric(w.opt)
  )
}
```

---

## 3. Quarterly rebalancing dates

We work with quarter end month ends (January, April, July, October).

```{r}
# Quarter end dates from the monthly series
EOQ_datevec <- return_mat %>%
  select(date) %>%
  distinct() %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  pull(date)
```

---

## 4. Rolling optimiser and results

```{r}
Roll_optimizer <- function(return_mat,
                           EOQ_datevec,
                           LookBackSel = 120,    # 120 months equals ten years
                           ...) {
  optim_foo(return_mat,
            end_date        = EOQ_datevec,
            lookback_months = LookBackSel,
            ...)
}

# Ten year look back as required
Result_roll <- EOQ_datevec %>%
  map_df(~ Roll_optimizer(return_mat,
                          EOQ_datevec = .,
                          LookBackSel = 120))

kable(Result_roll %>% head(13))
```

---

## 5. Stacked bar chart of portfolio weights over time

```{r}
# Prepare xts object of weights at rebalance dates
bar_xts <- Result_roll %>%
  select(date, stocks, weight) %>%
  spread(stocks, weight) %>%
  tbl_xts() %>%
  .[endpoints(., "months")]

# Plot stacked weights
chart.StackedBar(
  bar_xts,
  main = "Quarterly Rebalanced Global Balanced Index Fund"
)


```

# Clustering 
Using the Practical 5 (https://www.fmx.nfkatzke.com/posts/2020-08-17-practical-5/), I cluster the indices on a Ledoit–Wolf shrunk correlation matrix to identify statistical comovement groups. The dendrogram suggests, for example, that [equity indices] cluster closely, while [bond/credit] and commodities sit in distinct branches.
```{r}
## 6. Cluster based comovement analysis (Practical 5)

# Use the same monthly return matrix, drop date
clust_mat <- return_mat %>%
  select(-date)

# 6.1 Ledoit–Wolf covariance and correlation matrix (as in Practical 5)
Sigma_clust <- RiskPortfolios::covEstimation(
  as.matrix(clust_mat),
  control = list(type = "lw")
)

corr_clust <- cov2cor(Sigma_clust)

# 6.2 Distance matrix using the Dower metric: d_ij = sqrt( (1 - rho_ij) / 2 )
distmat <- ((1 - corr_clust) / 2)^0.5

# 6.3 Hierarchical clustering with Ward linkage (AGNES)
library(cluster)

hc <- cluster::agnes(dist(distmat), method = "ward")

# Basic dendrogram (you could beautify with the ggdendro helpers from Practical 5)
plot(hc, which.plot = 2, main = "Hierarchical clustering of global indices",
     xlab = "", sub = "")

```

