
```{r}

# Q1: ABC Fund vs Benchmark and Peers – 
Libraries and data ----

pacman::p_load(
  readr,
  dplyr,
  tidyr,
  ggplot2,
  lubridate,
  tbl2xts,
  PerformanceAnalytics,
  RcppRoll,
  roll,
  zoo,
  fmxdat
)

```


Load Rmd helper functions (themes etc.) if you use them elsewhere
devtools::source_gist("https://gist.github.com/Nicktz/bd2614f8f8a551881a1dc3c11a1e7268")


```{r}
# Raw data
Benchmark       <- read_rds("data/Capped_SWIX.rds")
Active_Managers <- read_rds("data/Active_Managers.rds")
ABC_Fund        <- read_rds("data/ABC_Fund.rds")

```


I fix a common sample based on ABC Fund dates
```{r}
start_date <- as.Date("2012-07-31")
end_date   <- as.Date("2025-10-31")

# Filter each data set to the common period
benchmark_common <- 
  Benchmark %>% 
  filter(date >= start_date, date <= end_date)

active_managers_common <- 
  Active_Managers %>% 
  filter(date >= start_date, date <= end_date)

abc_data <- 
  ABC_Fund %>% 
  filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# Prepare benchmark data
bench_data <- 
  benchmark_common %>% 
  select(date, Benchmark_Returns = Returns)

# Peer summary by date (average peer return)
peers_data <- 
  active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns    = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count          = n(),
    .groups              = "drop"
  )

# Single analysis data set we use for most of Q1
analysis_data <- 
  abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data, by = "date")

glimpse(analysis_data)
```



### 2. Rolling regression vs benchmark (beta time series)

```{r}

# 2. Rolling 3-year beta vs benchmark (ABC + peer avg) ----


# Ra: assets we are analysing (ABC + peer average)
Ra_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>% 
  mutate(across(-date, ~ coalesce(., 0))) %>% 
  tbl_xts(cols_to_xts = c("ABC_Fund", "Peer_Average"))

# Rb: benchmark we compare against
Rb_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(date, Benchmark = Benchmark_Returns) %>% 
  mutate(Benchmark = coalesce(Benchmark, 0)) %>% 
  tbl_xts(cols_to_xts = "Benchmark")

# Rolling 3-year betas (36 months) vs benchmark
PerformanceAnalytics::chart.RollingRegression(
  Ra         = Ra_xts,
  Rb         = Rb_xts,
  width      = 36,
  attribute  = "Beta",
  legend.loc = "topleft",
  main       = "Rolling 3-year beta vs benchmark: ABC vs peer average"
)

# Full sample betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = Ra_xts,
  Rb = Rb_xts
)
```



### 3. Boxplot of 3-year rolling betas 

```{r}
# 3. Boxplot of 3-year rolling betas (Active vs ABC) ----


# 3.1 3-year rolling annualised returns for the benchmark
Benchmark_roll <-
  benchmark_common %>%
  arrange(date) %>%
  group_by(Name) %>%   # only one benchmark name, but keep for consistency
  mutate(
    bench_roll = RcppRoll::roll_prod(1 + Returns,
                                     n     = 36,
                                     fill  = NA,
                                     align = "right")^(12 / 36) - 1
  ) %>%
  group_by(date, Name) %>%
  filter(any(!is.na(bench_roll))) %>%
  ungroup() %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, bench_roll)

# 3.2 3-year rolling annualised returns for all active managers
Active_Managers_roll <-
  active_managers_common %>%
  arrange(date) %>%
  group_by(Name) %>%
  mutate(
    active_roll = RcppRoll::roll_prod(1 + Returns,
                                      n     = 36,
                                      fill  = NA,
                                      align = "right")^(12 / 36) - 1
  ) %>%
  group_by(date, Name) %>%
  filter(any(!is.na(active_roll))) %>%
  ungroup() %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, Name, active_roll)

# 3.3 3-year rolling annualised returns for the ABC Fund
ABC_Fund_roll <-
  ABC_Fund %>%
  filter(date >= start_date, date <= end_date) %>%
  arrange(date) %>%
  group_by(Name) %>%
  mutate(
    ABC_roll = RcppRoll::roll_prod(1 + Returns,
                                   n     = 36,
                                   fill  = NA,
                                   align = "right")^(12 / 36) - 1
  ) %>%
  group_by(date, Name) %>%
  filter(any(!is.na(ABC_roll))) %>%
  ungroup() %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, ABC_roll)

# 3.4 Betas for Active Managers: regress 3y rolling returns on benchmark, by year and fund
AM_Beta <-
  Active_Managers_roll %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y, Name) %>%
  summarise(
    AM_Beta = coef(lm(active_roll ~ bench_roll))[2],
    .groups = "drop"
  )

# 3.5 Betas for ABC Fund: same but only one fund per year
ABC_Beta <-
  ABC_Fund_roll %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y) %>%
  summarise(
    ABC_Beta = coef(lm(ABC_roll ~ bench_roll))[2],
    .groups  = "drop"
  )

# 3.6 Data for boxplot: peer beta distribution and ABC beta per year
# 3.6 Data for boxplot: peer beta distribution and ABC beta per year
Main_box <-
  AM_Beta %>%
  left_join(ABC_Beta, by = "Y") %>%
  filter(Y >= "2016")   # optional: keep recent years only

# 3.7 Boxplot: Year on x, Beta on y
Main_box_plot <-
  Main_box %>%
  ggplot() +
  geom_boxplot(
    aes(x = factor(Y), y = AM_Beta),
    colour = "darkslategray",
    fill   = "lightblue",
    width  = 0.5
  ) +
  geom_point(
    aes(x = factor(Y), y = ABC_Beta),
    colour = "#FF4500",
    size   = 3,
    shape  = 18
  ) +
  labs(
    title    = "3-year rolling beta: Active managers vs ABC Fund",
    subtitle = "Distribution of active manager betas vs ABC beta (per year)",
    x        = "Year",
    y        = "3-year rolling beta vs benchmark",
    caption  = paste(
      "Note:\n",
      "The orange diamond shows the ABC Fund's 3-year rolling beta.\n",
      "The boxplots show the distribution of active managers' betas for each year."
    )
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  scale_y_continuous(breaks = seq(from = -2, to = 3, by = 0.5))

Main_box_plot
```



### 4. Summary statistics, SFM, CAPM alpha / beta

```{r}

# 4. Stats, trailing performance, downside risk, SFM ----

# Combine benchmark, ABC and peer average into one xts object
returns_xts <- merge(Rb_xts, Ra_xts, join = "inner")
colnames(returns_xts) <- c("Benchmark", "ABC_Fund", "Peer_Average")

# 4.1 Summary statistics
PerformanceAnalytics::table.Stats(returns_xts)

# 4.2 Trailing-period performance (months: 3, 6, 12, 36, 60)
PerformanceAnalytics::table.TrailingPeriods(
  R       = returns_xts,
  periods = c(3, 6, 12, 36, 60)
)

# 4.3 Downside risk measures (Rf = 0)
PerformanceAnalytics::table.DownsideRisk(
  R  = returns_xts,
  Rf = 0
)

# 4.4 Single-factor model vs benchmark
PerformanceAnalytics::table.SFM(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)

# 4.5 Betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)

# 4.6 Annualised alpha for ABC vs benchmark (monthly data)
(1 + PerformanceAnalytics::CAPM.alpha(
  Ra = returns_xts[, "ABC_Fund"],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
))^12 - 1
```



### 5. Cumulative returns: ABC vs peers vs benchmark

```{r}

# 5. Cumulative returns ABC vs Peer Avg vs Benchmark ----

cum_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    ret    = coalesce(ret, 0),
    CumRet = cumprod(1 + ret)   # growth of 1 rand
  ) %>%
  ungroup()

g_cum <-
  cum_df %>%
  ggplot(aes(x = date, y = CumRet, colour = Series)) +
  geom_line() +
  labs(
    title = "Cumulative returns: ABC Fund vs peer average and benchmark",
    x     = "",
    y     = "Growth of 1 rand"
  ) +
  theme_fmx()

finplot(
  g_cum,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)

# Log cumulative plot
g_cum +
  coord_trans(y = "log10") +
  labs(
    title = paste0(g_cum$labels$title, "\nLog Scaled"),
    y     = "Log scaled cumulative returns"
  )
```

---

### 6. Rolling 3 year annualised returns

```{r}
# 6. Rolling 3-year annualised returns ----


roll_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    ret      = coalesce(ret, 0),
    RollRets = RcppRoll::roll_prod(1 + ret, 36,
                                   fill  = NA,
                                   align = "right")^(12 / 36) - 1
  ) %>%
  group_by(date) %>%
  filter(any(!is.na(RollRets))) %>%
  ungroup()

g_roll <-
  roll_df %>%
  ggplot(aes(x = date, y = RollRets, colour = Series)) +
  geom_line(alpha = 0.7, linewidth = 1.1) +
  labs(
    title = "Rolling 3-year annualised returns",
    x     = "",
    y     = "Rolling 3-year return (annualised)"
  ) +
  theme_fmx()

finplot(
  g_roll,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)
```

---

### 7. Multi horizon annualised return bar plot
```{r}

# 7. 6m / 1y / 3y / 5y annualised return bar chart ----


idx <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  arrange(date)

dfplot <-
  bind_rows(
    # 6 months (not annualised)
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>%
      mutate(Freq = "A"),
    # 1 year
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 12) - 1) %>%
      mutate(Freq = "B"),
    # 3 years
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 36) - 1) %>%
      mutate(Freq = "C"),
    # 5 years
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 60) - 1) %>%
      mutate(Freq = "D")
  )

to_string <- as_labeller(
  c(
    `A` = "6 Months",
    `B` = "1 Year",
    `C` = "3 Years",
    `D` = "5 Years"
  )
)

g_ann <-
  dfplot %>%
  ggplot() +
  geom_bar(aes(Series, mu, fill = Series), stat = "identity") +
  facet_wrap(~Freq, labeller = to_string, nrow = 1) +
  labs(
    x       = "",
    y       = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualised terms."
  ) +
  fmx_fills() +
  geom_label(
    aes(Series, mu, label = paste0(round(mu, 4) * 100, "%")),
    size     = ggpts(8),
    alpha    = 0.35,
    fontface = "bold",
    nudge_y  = 0.002
  ) +
  theme_fmx(
    CustomCaption = TRUE,
    title.size    = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size  = ggpts(30),
    axis.size     = ggpts(37),
    legend.size   = ggpts(35),
    legend.pos    = "top"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.title.y = element_text(vjust = 2),
    strip.text.x = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann
```



### 8. Rolling 3 year volatility (annualised)
```{r}

# 8. Rolling 3-year annualised volatility ----
plot_dlog <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    logret = log(1 + ret),
    RollSD = RcppRoll::roll_sd(logret, 36,
                               fill  = NA,
                               align = "right") * sqrt(12)
  ) %>%
  filter(!is.na(RollSD)) %>%
  ungroup()

g_vol <-
  plot_dlog %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Series),
            alpha = 0.7, linewidth = 1.25) +
  labs(
    title    = "Rolling 3-year annualised volatility",
    subtitle = "",
    x        = "",
    y        = "Rolling 3-year SD (annualised)",
    caption  = "Note:\nVolatility from 36-month log returns, annualised."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(
  g_vol,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)
```

