
---
title: "ABC Fund Performance Analysis"
author: "Financial Analytics Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,    # show code
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos = "H",
  out.width = "90%"
)
```

# Abstract

This analysis examines ABC Fund's performance relative to its benchmark and industry peers across multiple dimensions including risk exposure, return generation, and market timing.

## Data and Methodology

```{r libraries_data, echo=TRUE}
# Load required libraries
pacman::p_load(
  readr, dplyr, tidyr, ggplot2, lubridate, tbl2xts,
  PerformanceAnalytics, RcppRoll, roll, zoo, fmxdat, kableExtra
)

# Load data
# Load the data
Benchmark <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/Capped_SWIX.rds")
Active_Managers <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/Active_Managers.rds")
ABC_Fund <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/ABC_Fund.rds")
# Set common sample period
start_date <- as.Date("2012-07-31")
end_date <- as.Date("2025-10-31")

# Filter datasets to common period
benchmark_common <- Benchmark %>% filter(date >= start_date, date <= end_date)
active_managers_common <- Active_Managers %>% filter(date >= start_date, date <= end_date)
abc_data <- ABC_Fund %>% filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# Prepare analysis dataset
bench_data <- benchmark_common %>% select(date, Benchmark_Returns = Returns)

peers_data <- active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count = n(),
    .groups = "drop"
  )

analysis_data <- abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data, by = "date")

cat("Analysis period:", format(start_date, "%Y-%m-%d"), "to", format(end_date, "%Y-%m-%d"), "\n")
cat("Total observations:", nrow(analysis_data), "\n")
```

# Performance Overview

## Summary Statistics

```{r summary_stats, echo=TRUE}
# Prepare returns data
returns_xts <- analysis_data %>%
  select(date, Benchmark = Benchmark_Returns, ABC_Fund = ABC_Returns, 
         Peer_Average = Peers_Avg_Returns) %>%
  mutate(across(-date, ~ coalesce(., 0))) %>%
  tbl_xts(cols_to_xts = c("Benchmark", "ABC_Fund", "Peer_Average"))

# Calculate summary statistics
stats_table <- table.Stats(returns_xts)
stats_table_clean <- round(stats_table, 4)

kable(stats_table_clean, caption = "Summary Statistics of Returns", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Trailing Period Performance

```{r trailing_performance, echo=TRUE}
trailing_table <- table.TrailingPeriods(R = returns_xts, periods = c(3, 6, 12, 36, 60))
trailing_table_clean <- round(trailing_table, 4)

kable(trailing_table_clean, caption = "Trailing Period Returns", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

# Risk Analysis

## Downside Risk Measures

```{r downside_risk, echo=TRUE}
downside_table <- table.DownsideRisk(R = returns_xts, Rf = 0)
downside_table_clean <- round(downside_table, 4)

kable(downside_table_clean, caption = "Downside Risk Measures", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Single Factor Model Analysis

```{r single_factor, echo=TRUE}
sfm_table <- table.SFM(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)
sfm_table_clean <- round(sfm_table, 4)

kable(sfm_table_clean, caption = "Single Factor Model vs Benchmark", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

# Rolling Beta Analysis

## Time-Varying Market Exposure

```{r rolling_beta_plot, echo=TRUE, fig.height=6, fig.cap="Rolling 3-year beta vs benchmark: ABC Fund vs Peer Average"}
# Calculate rolling betas
Ra_xts <- returns_xts[, c("ABC_Fund", "Peer_Average")]
Rb_xts <- returns_xts[, "Benchmark"]

chart.RollingRegression(
  Ra = Ra_xts,
  Rb = Rb_xts,
  width = 36,
  attribute = "Beta",
  legend.loc = "topleft",
  main = "Rolling 3-year beta vs benchmark: ABC vs peer average"
)
```

## Beta Distribution Comparison

```{r beta_boxplot, echo=TRUE, fig.height=6, fig.cap="3-year rolling beta distribution: Active managers vs ABC Fund"}
# Prepare data for beta boxplot
Benchmark_roll <- benchmark_common %>%
  arrange(date) %>%
  group_by(Name) %>%
  mutate(
    bench_roll = RcppRoll::roll_prod(1 + Returns, n = 36, fill = NA, align = "right")^(12/36) - 1
  ) %>%
  ungroup() %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, bench_roll)

Active_Managers_roll <- active_managers_common %>%
  arrange(date) %>%
  group_by(Name) %>%
  mutate(
    active_roll = RcppRoll::roll_prod(1 + Returns, n = 36, fill = NA, align = "right")^(12/36) - 1
  ) %>%
  ungroup() %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, Name, active_roll)

ABC_Fund_roll <- abc_data %>%
  arrange(date) %>%
  mutate(
    ABC_roll = RcppRoll::roll_prod(1 + ABC_Returns, n = 36, fill = NA, align = "right")^(12/36) - 1
  ) %>%
  mutate(Y = format(date, "%Y")) %>%
  select(date, Y, ABC_roll)

# Calculate betas
AM_Beta <- Active_Managers_roll %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y, Name) %>%
  summarise(AM_Beta = coef(lm(active_roll ~ bench_roll))[2], .groups = "drop")

ABC_Beta <- ABC_Fund_roll %>%
  left_join(Benchmark_roll, by = c("date", "Y")) %>%
  group_by(Y) %>%
  summarise(ABC_Beta = coef(lm(ABC_roll ~ bench_roll))[2], .groups = "drop")

# Create boxplot
Main_box <- AM_Beta %>% left_join(ABC_Beta, by = "Y") %>% filter(Y >= "2016")

ggplot(Main_box) +
  geom_boxplot(aes(x = factor(Y), y = AM_Beta), 
               colour = "darkslategray", fill = "lightblue", width = 0.5) +
  geom_point(aes(x = factor(Y), y = ABC_Beta), 
             colour = "#FF4500", size = 3, shape = 18) +
  labs(
    title = "3-year rolling beta: Active managers vs ABC Fund",
    subtitle = "Distribution of active manager betas vs ABC beta (per year)",
    x = "Year", 
    y = "3-year rolling beta vs benchmark"
  ) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(from = -2, to = 3, by = 0.5))
```

# Return Analysis

## Cumulative Performance

```{r cumulative_returns, echo=TRUE, fig.height=6, fig.cap="Cumulative returns: ABC Fund vs peer average and benchmark"}
cum_df <- analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark = Benchmark_Returns,
    ABC_Fund = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  group_by(Series) %>%
  mutate(
    ret = coalesce(ret, 0),
    CumRet = cumprod(1 + ret)
  ) %>%
  ungroup()

ggplot(cum_df, aes(x = date, y = CumRet, colour = Series)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Cumulative returns: ABC Fund vs peer average and benchmark",
    x = "", 
    y = "Growth of 1 rand"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Benchmark" = "blue", "ABC_Fund" = "red", "Peer_Average" = "green"))
```

## Multi-Horizon Performance

```{r multi_horizon, echo=TRUE, fig.height=6, fig.cap="Annualized returns across different time horizons"}
# Calculate returns for different horizons
idx <- analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark = Benchmark_Returns,
    ABC_Fund = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  arrange(date)

dfplot <- bind_rows(
  idx %>% filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>%
    group_by(Series) %>% summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>% mutate(Freq = "6 Months"),
  idx %>% filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>%
    group_by(Series) %>% summarise(mu = prod(1 + ret, na.rm = TRUE)^(12/12) - 1) %>% mutate(Freq = "1 Year"),
  idx %>% filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>%
    group_by(Series) %>% summarise(mu = prod(1 + ret, na.rm = TRUE)^(12/36) - 1) %>% mutate(Freq = "3 Years"),
  idx %>% filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>%
    group_by(Series) %>% summarise(mu = prod(1 + ret, na.rm = TRUE)^(12/60) - 1) %>% mutate(Freq = "5 Years")
)

ggplot(dfplot) +
  geom_bar(aes(Series, mu, fill = Series), stat = "identity") +
  facet_wrap(~Freq, nrow = 1) +
  labs(x = "", y = "Returns (Annualized)") +
  theme_minimal() +
  geom_text(aes(Series, mu, label = paste0(round(mu * 100, 2), "%")), 
            vjust = -0.5, size = 3)
```

# Risk-Return Profile

## Rolling Volatility

```{r rolling_volatility, echo=TRUE, fig.height=6, fig.cap="Rolling 3-year annualized volatility"}
plot_dlog <- analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark = Benchmark_Returns,
    ABC_Fund = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(cols = -date, names_to = "Series", values_to = "ret") %>%
  group_by(Series) %>%
  mutate(
    logret = log(1 + ret),
    RollSD = RcppRoll::roll_sd(logret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  filter(!is.na(RollSD)) %>%
  ungroup()

ggplot(plot_dlog) +
  geom_line(aes(date, RollSD, colour = Series), alpha = 0.7, linewidth = 1.25) +
  labs(
    title = "Rolling 3-year annualized volatility",
    x = "", 
    y = "Rolling 3-year SD (annualized)"
  ) +
  theme_minimal()
```

## Risk-Return Scatter

```{r risk_return_scatter, echo=TRUE, fig.height=6, fig.cap="Risk-Return scatter plot"}
# Calculate annualized returns and volatility
annual_stats <- analysis_data %>%
  summarise(
    ABC_Return = prod(1 + ABC_Returns, na.rm = TRUE)^(12/n()) - 1,
    ABC_Vol = sd(ABC_Returns, na.rm = TRUE) * sqrt(12),
    Peer_Return = prod(1 + Peers_Avg_Returns, na.rm = TRUE)^(12/n()) - 1,
    Peer_Vol = sd(Peers_Avg_Returns, na.rm = TRUE) * sqrt(12),
    Benchmark_Return = prod(1 + Benchmark_Returns, na.rm = TRUE)^(12/n()) - 1,
    Benchmark_Vol = sd(Benchmark_Returns, na.rm = TRUE) * sqrt(12)
  )

scatter_data <- data.frame(
  Series = c("ABC Fund", "Peer Average", "Benchmark"),
  Return = c(annual_stats$ABC_Return, annual_stats$Peer_Return, annual_stats$Benchmark_Return),
  Risk = c(annual_stats$ABC_Vol, annual_stats$Peer_Vol, annual_stats$Benchmark_Vol)
)

ggplot(scatter_data, aes(x = Risk, y = Return, color = Series)) +
  geom_point(size = 4) +
  geom_text(aes(label = Series), vjust = -0.5, hjust = 0.5) +
  labs(
    title = "Risk-Return Profile",
    x = "Annualized Volatility",
    y = "Annualized Return"
  ) +
  theme_minimal() +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent)
```

# Conclusion

## Key Findings

```{r conclusion_stats, echo=FALSE}
# Calculate key performance metrics
full_sample_betas <- CAPM.beta(Ra = Ra_xts, Rb = Rb_xts)
abc_alpha_annual <- (1 + CAPM.alpha(Ra = Ra_xts[, "ABC_Fund"], Rb = Rb_xts, Rf = 0))^12 - 1

conclusion_table <- data.frame(
  Metric = c("ABC Fund Beta", "Peer Average Beta", "ABC Annual Alpha", "Information Ratio"),
  Value = c(
    round(full_sample_betas["ABC_Fund"], 3),
    round(full_sample_betas["Peer_Average"], 3),
    paste0(round(abc_alpha_annual * 100, 2), "%"),
    round(mean(returns_xts[, "ABC_Fund"] - returns_xts[, "Benchmark"], na.rm = TRUE) / 
          sd(returns_xts[, "ABC_Fund"] - returns_xts[, "Benchmark"], na.rm = TRUE), 3)
  )
)

kable(conclusion_table, caption = "Key Performance Metrics", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

