
---
title: "Question 1: Portfolio Return Comparison"
author: "Precious Nhamo (Student Number: …)"
date: "`r Sys.Date()`"
bibliography: Tex/ref.bib
output:
  pagedown::html_paged:
    css: ["Template/default-fonts-Texevier.css",
          "Template/default-page-Texevier.css",
          "Template/default-Texevier.css"]
    csl: Template/harvard-stellenbosch-university.csl
    template: ["Template/paged-Texevier.html"]
    toc: true
    self_contained: true
abstract: |
    In this report I provide an overview of the performance of a South
    African equity fund, ABC Fund, relative to its industry peers and the
    Capped SWIX benchmark. I focus on rolling three year beta estimates,
    relative performance visualisations and the interaction between
    cross sectional return dispersion and the fund’s performance. The
    objective is to help a non quantitative portfolio manager
    communicate where his fund offers something different to pure market
    beta and how this difference links to periods of high dispersion.
---
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  error   = TRUE
)

pacman::p_load(
  tidyverse,
  lubridate,
  zoo,
  PerformanceAnalytics,
  ggthemes,
  patchwork,
  broom,
  knitr,
  scales
)
```

\newpage

# Introduction {-}

In this section I briefly explain the setting:

* ABC Fund operates in an industry where the **Capped SWIX** is the primary benchmark.
* The manager wants to understand and communicate:

  * How typical **rolling three year beta** looks for active managers.
  * Whether ABC Fund offers **something other than pure market beta**.
  * How ABC Fund has performed relative to peers over time.
  * Whether its performance improves when **cross sectional dispersion** is high.

I state the aims clearly:

1. Summarise peer and benchmark returns and place ABC Fund in that context.
2. Quantify and visualise the **rolling three year beta** of the fund relative to peers.
3. Show the **relative performance path** of ABC Fund versus peers and benchmark.
4. Explore the claim that **high dispersion rewards contrarian positioning**.

# Data {-}

Here I describe the three data sets:

* **Benchmark**: `Capped_SWIX.rds`

  * Monthly benchmark returns and dates.

* **Active managers**: `Active_Managers.rds`

  * Panel of peer fund returns (multiple managers over time).

* **ABC Fund**: `ABC_Fund.rds`

  * Return history for ABC Fund.

I discuss:

* Sample period used and common date intersection.
* Frequency and treatment of missing observations.
* Any basic summary statistics:

  * Mean and volatility of benchmark, peers and ABC.
  * Number of peers through time.

# Methodology {-}

In this section I outline the main analytical steps.

## Common sample and alignment {-}

* I align all three data sets on a **common sample period**.
* I construct a **peer average return series** (for example equally weighted across managers).
* I ensure returns for ABC Fund, peer average and benchmark are aligned in time.

## Rolling three year beta estimation {-}

* I define **three year rolling windows** (36 months) and estimate CAPM style regressions of:

  * ABC Fund returns on benchmark returns.
  * Peer average returns on benchmark returns.
* I obtain **rolling beta series** for ABC and the peer group.
* These series are later used both for:

  * Time series plots.
  * Cross sectional summaries (for example box and scatter by year).

## Relative performance and attribution views {-}

* I compute cumulative performance:

  * ABC Fund versus peer average versus benchmark.
* I construct visuals that:

  * Show cumulative return paths over time.
  * Highlight periods where ABC materially outperforms or underperforms.

Optionally I may add simple Sharpe ratios or excess return statistics.

## Return dispersion and “contrarian” payoff {-}

* I construct a **dispersion measure** for each period, for example:

  * Cross sectional standard deviation of peer returns.
* I then examine:

  * Relationship between dispersion and **ABC Fund relative performance**.
  * Possibly conditional averages (high dispersion versus low dispersion periods).

This addresses the manager’s claim that high dispersion rewards differentiated positioning.

# Results {-}

In this section I describe what the figures and tables show.

## Rolling beta behaviour {-}

* I show the time series of rolling three year beta for ABC and peers.
* I discuss:

  * Whether ABC Fund’s beta is consistently above, below or near one.
  * How ABC compares to the typical peer beta distribution.
  * Whether ABC genuinely provides exposure different from market beta.

## Relative performance paths {-}

* I present a cumulative performance chart:

  * ABC Fund versus peer average versus benchmark.
* I comment on:

  * Periods of sustained outperformance or underperformance.
  * Whether ABC adds value net of benchmark over the sample.

## Dispersion and fund performance {-}

* I summarise how ABC’s relative returns behave conditional on:

  * High versus low dispersion periods.
* I comment on whether the data support the manager’s claim that:

  * High dispersion tends to coincide with better results for ABC Fund.

# Conclusion {-}

Here I summarise the main findings in plain language:

* Whether ABC Fund provides exposure meaningfully different from market beta.
* Whether ABC has delivered attractive relative performance compared to peers and benchmark.
* Whether the evidence supports the narrative around dispersion and contrarian positioning.

I also note any caveats (sample length, survivorship, simplifying assumptions).

# References {-}

<div id="refs"></div>

# Appendix {-}

## Appendix A: Additional tables {-}

Additional summary statistics or regression tables.

## Appendix B: Additional figures {-}

Extra visuals such as box scatter views of rolling betas by year.

---

---
title: "Q1"
output: html_document
---


```{r load-data-Q1, echo = TRUE}
# Load readr for rds data sets 
pacman::p_load(readr, fmxdat)



# Load the data
Benchmark <- read_rds("data/Capped_SWIX.rds")
Active_Managers <- read_rds("data/Active_Managers.rds")
ABC_Fund <- read_rds("data/ABC_Fund.rds")

```

Since we have a benchmark , ABC fund industry peer we can't analyse all individual industry peers - we will calculate the mean or average of the peers 

```{r}
# Load required libraries with pacman
pacman::p_load(
  dplyr,
  ggplot2,
  lubridate,
  tidyr,
  roll,
  zoo
)

# 1. Set common sample based on ABC Fund dates
start_date <- as.Date("2012-07-31")
end_date   <- as.Date("2025-10-31")

# 2. Filter each data set to the common period
benchmark_common <- 
  Benchmark %>% 
  filter(date >= start_date, date <= end_date)

active_managers_common <- 
  Active_Managers %>% 
  filter(date >= start_date, date <= end_date)

abc_data <- 
  ABC_Fund %>% 
  filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# 3. Prepare benchmark data
bench_data <- 
  benchmark_common %>% 
  select(date, Benchmark_Returns = Returns)

# 4. Prepare peer summary by date
peers_data <- 
  active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns    = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count          = n(),
    .groups              = "drop"
  )

# 5. Merge into a single analysis data set
analysis_data <- 
  abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data,  by = "date")

# Optional quick check
dplyr::glimpse(analysis_data)


```

We will be using analysis data to continue with our questions 



```{r}
pacman::p_load(
  dplyr,
  tbl2xts,
  PerformanceAnalytics
)

# Ra: assets we are analysing (ABC + peer average)
Ra_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>% 
  mutate(across(-date, ~ coalesce(., 0))) %>% 
  tbl_xts(
    cols_to_xts = c("ABC_Fund", "Peer_Average")
  )

# Rb: benchmark we compare against
Rb_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    Benchmark = Benchmark_Returns
  ) %>% 
  mutate(Benchmark = coalesce(Benchmark, 0)) %>% 
  tbl_xts(
    cols_to_xts = "Benchmark"
  )

```

```{r}
# Rolling 3 year betas (36 months) vs benchmark
PerformanceAnalytics::chart.RollingRegression(
  Ra         = Ra_xts,
  Rb         = Rb_xts,
  width      = 36,
  attribute  = "Beta",
  legend.loc = "topleft",
  main       = "Rolling 3 year beta vs benchmark: ABC vs peer average"
)

# Full sample betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = Ra_xts,
  Rb = Rb_xts
)


```

Box-Scatter Plot ( I struggled with this part)

```{r}
pacman::p_load(
  dplyr,
  tidyr,
  ggplot2,
  roll,
  fmxdat
)

# 1. Compute rolling 3 year betas (36 months) vs benchmark

# make sure data is ordered
beta_data <- 
  analysis_data %>% 
  arrange(date)

# rolling regression: y on x with width = 36
# ABC vs Benchmark
roll_abc <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$ABC_Returns,
    width = 36
  )

# Peer average vs Benchmark
roll_peer <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$Peers_Avg_Returns,
    width = 36
  )

# extract slope (beta) from coefficients[, 2]
rolling_betas <- 
  tibble(
    date        = beta_data$date,
    ABC_Fund    = as.numeric(roll_abc$coefficients[, 2]),
    Peer_Average = as.numeric(roll_peer$coefficients[, 2])
  ) %>% 
  pivot_longer(
    cols      = c(ABC_Fund, Peer_Average),
    names_to  = "Series",
    values_to = "Beta"
  ) %>% 
  filter(!is.na(Beta))   # drop early windows with no full 36 months

# 2. Boxplot with scatter (jitter) on top

ggplot(rolling_betas, aes(x = Series, y = Beta)) +
  geom_jitter(width = 0.1, alpha = 0.4) +   # scatter points
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +  # box on top
  labs(
    title = "Distribution of rolling 3 year betas vs benchmark",
    x     = NULL,
    y     = "Rolling 3 year beta"
  ) +
  theme_fmx()

```

```{r}

# Combine benchmark, ABC and peer average into one xts object
returns_xts <- merge(Rb_xts, Ra_xts, join = "inner")
colnames(returns_xts) <- c("Benchmark", "ABC_Fund", "Peer_Average")

# 1. Summary statistics
PerformanceAnalytics::table.Stats(returns_xts)

# 2. Trailing-period performance (in months: 3, 6, 12, 36, 60)
PerformanceAnalytics::table.TrailingPeriods(
  R       = returns_xts,
  periods = c(3, 6, 12, 36, 60)
)

# 3. Downside risk measures (using 0 as risk-free rate here)
PerformanceAnalytics::table.DownsideRisk(
  R  = returns_xts,
  Rf = 0
)

# 4. Single-factor model and CAPM-style alpha/beta vs benchmark
# (a) SFM summary for ABC and Peer_Average vs Benchmark
PerformanceAnalytics::table.SFM(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)

# (b) Betas vs benchmark
PerformanceAnalytics::CAPM.beta(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)

# (c) Annualised alpha for ABC vs benchmark (since data is monthly)
(1 + PerformanceAnalytics::CAPM.alpha(
  Ra = returns_xts[, "ABC_Fund"],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
))^12 - 1

```


We will do other interesting things (the above tables answer how much we differe from the rest of the industry)

```{r}
pacman::p_load(dplyr, tidyr, ggplot2, fmxdat)

# Build long data set with a 0% risk-free series
cum_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    ret    = coalesce(ret, 0),
    CumRet = cumprod(1 + ret)   # growth of 1 rand
  ) %>%
  ungroup()

g_cum <-
  cum_df %>%
  ggplot(aes(x = date, y = CumRet, colour = Series)) +
  geom_line() +
  labs(
    title = "Cumulative returns: ABC Fund vs Peers(Average), Benchmark ",
    x     = "",
    y     = "Growth of 1 rand"
  ) +
  theme_fmx()

# Nice date axis + optional % formatting on y
finplot(
  g_cum,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)


# Log cumulative plot:
g_cum  + coord_trans(y = 'log10') + labs(title = paste0( g_cum$labels$title, "\nLog Scaled"), y = "Log Scaled Cumulative Returns")


```




```{r}
pacman::p_load(dplyr, tidyr, ggplot2, RcppRoll, fmxdat)

roll_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  # make sure NAs don’t break the product
  mutate(
    ret      = coalesce(ret, 0),
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right")^(12 / 36) - 1
  ) %>%
  # keep only dates where at least one series has a rolling return
  group_by(date) %>%
  filter(any(!is.na(RollRets))) %>%
  ungroup()

g_roll <-
  roll_df %>%
  ggplot(aes(x = date, y = RollRets, colour = Series)) +
  geom_line(alpha = 0.7, linewidth = 1.1) +
  labs(
    title = "Rolling 3-year annualised returns",
    x     = "",
    y     = "Rolling 3-year return (annualised)"
  ) +
  theme_fmx()

finplot(
  g_roll,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)

```



```{r}

options(dplyr.summarise.inform = FALSE)
pacman::p_load(dplyr, tidyr, ggplot2, lubridate, tbl2xts, PerformanceAnalytics, fmxdat)

#========================================================
# 1. Build "idx" style object from our monthly return data
#    (Benchmark, ABC_Fund, Peer_Average)
#========================================================

idx <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  # already monthly data, so no need for YM collapsing
  arrange(date) %>%
  group_by(Series) %>%
  ungroup()

#========================================================
# 2. Manual calculation of 6m, 12m, 3y, 5y annualised returns
#========================================================

dfplot <-
  bind_rows(
    # 6 months (not annualised in Nico's example)
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>%
      mutate(Freq = "A"),
    # 12 months (annualised over 1 year => exponent 12/12)
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 12) - 1) %>%
      mutate(Freq = "B"),
    # 3 years (36 months, annualised)
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 36) - 1) %>%
      mutate(Freq = "C"),
    # 5 years (60 months, annualised)
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 60) - 1) %>%
      mutate(Freq = "D")
  )

#========================================================
# 3. Same thing via PerformanceAnalytics for cross-check
#========================================================

idxxts <-
  tbl_xts(idx, cols_to_xts = ret, spread_by = Series)

dfplotxts <-
  bind_rows(
    idxxts %>%
      tail(6) %>%
      PerformanceAnalytics::Return.annualized(scale = 12) %>%
      data.frame() %>%
      mutate(Freq = "A"),
    idxxts %>%
      tail(12) %>%
      PerformanceAnalytics::Return.annualized(scale = 12) %>%
      data.frame() %>%
      mutate(Freq = "B"),
    idxxts %>%
      tail(36) %>%
      PerformanceAnalytics::Return.annualized(scale = 12) %>%
      data.frame() %>%
      mutate(Freq = "C"),
    idxxts %>%
      tail(60) %>%
      PerformanceAnalytics::Return.annualized(scale = 12) %>%
      data.frame() %>%
      mutate(Freq = "D")
  ) %>%
  data.frame() %>%
  tidyr::gather(Series, mu, -Freq)

#========================================================
# 4. Barplot: annualised returns over different horizons
#========================================================

to_string <- as_labeller(
  c(
    `A` = "6 Months",
    `B` = "1 Year",
    `C` = "3 Years",
    `D` = "5 Years"
  )
)

g_ann <-
  dfplot %>%
  ggplot() +
  geom_bar(aes(Series, mu, fill = Series), stat = "identity") +
  facet_wrap(~Freq, labeller = to_string, nrow = 1) +
  labs(
    x       = "",
    y       = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualised terms."
  ) +
  fmx_fills() +
  geom_label(
    aes(Series, mu, label = paste0(round(mu, 4) * 100, "%")),
    size        = ggpts(8),
    alpha       = 0.35,
    fontface    = "bold",
    nudge_y     = 0.002
  ) +
  theme_fmx(
    CustomCaption = TRUE,
    title.size    = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size  = ggpts(30),
    axis.size     = ggpts(37),
    legend.size   = ggpts(35),
    legend.pos    = "top"
  ) +
  theme(
    axis.text.x   = element_blank(),
    axis.title.y  = element_text(vjust = 2),
    strip.text.x  = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann

```


```{r}

pacman::p_load(dplyr, tidyr, ggplot2, RcppRoll, fmxdat)

plot_dlog <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    logret = log(1 + ret),
    # Rolling 36-month SD of log returns, annualised
    RollSD = RcppRoll::roll_sd(logret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  filter(!is.na(RollSD)) %>%
  ungroup()

g <-
  plot_dlog %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Series), alpha = 0.7, linewidth = 1.25) +
  labs(
    title    = "Rolling 3-year annualised volatility",
    subtitle = "",
    x        = "",
    y        = "Rolling 3-year SD (annualised)",
    caption  = "Note:\nVolatility estimated from 36-month log returns, annualised."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(
  g,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)

```

let's do this : ## Or check out this risk-return scatter for several stocks since 2010...
chart.RiskReturnScatter(R=xts.data.dailyreturns['2003-01-01/'][,c("AGL", "AMS", "ANG", "AOD

# Table (I want to do a replicate from bonus practical 2 Financial Statistics I am having issues)


Market dispersion











We begin by graphing the rolling 3-year beta for the active manager then we do the boxplot with the scatterplot and lastly how it differs to answer the first question 



