---
title: "Untitled"
output: html_document
---


In this report I provide an overview of the performance of a South African equity fund, ABC Fund, relative to its industry peers and the Capped SWIX benchmark. I focus on rolling three year beta estimates,relative performance visualisations and the interaction between
cross sectional return dispersion and the fund's performance. The objective is to help a non quantitative portfolio manager communicate where his fund offers something different to pure market beta and how this difference links to periods of high dispersion.


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  error   = TRUE
)

pacman::p_load(
  tidyverse,
  lubridate,
  zoo,
  PerformanceAnalytics,
  ggthemes,
  patchwork,
  broom,
  knitr,
  scales
)
```

\newpage

# Data Preparation

```{r load-data-Q1, error=TRUE}
# Load required packages and data
pacman::p_load(readr, fmxdat, dplyr, ggplot2, lubridate, tidyr, roll, zoo, tbl2xts, PerformanceAnalytics, RcppRoll)

# Load the data
Benchmark <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/Capped_SWIX.rds")
Active_Managers <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/Active_Managers.rds")
ABC_Fund <- read_rds("C:/Users/pmnha/OneDrive/Desktop/22660348/data/ABC_Fund.rds")

# Set common sample based on ABC Fund dates
start_date <- as.Date("2012-07-31")
end_date   <- as.Date("2025-10-31")

# Filter each data set to the common period
benchmark_common <- 
  Benchmark %>% 
  filter(date >= start_date, date <= end_date)

active_managers_common <- 
  Active_Managers %>% 
  filter(date >= start_date, date <= end_date)

abc_data <- 
  ABC_Fund %>% 
  filter(date >= start_date, date <= end_date) %>% 
  select(date, ABC_Returns = Returns)

# Prepare benchmark data
bench_data <- 
  benchmark_common %>% 
  select(date, Benchmark_Returns = Returns)

# Prepare peer summary by date
peers_data <- 
  active_managers_common %>% 
  group_by(date) %>% 
  summarise(
    Peers_Avg_Returns    = mean(Returns, na.rm = TRUE),
    Peers_Median_Returns = median(Returns, na.rm = TRUE),
    Peers_Count          = n(),
    .groups              = "drop"
  )

# Merge into a single analysis data set
analysis_data <- 
  abc_data %>% 
  left_join(bench_data, by = "date") %>% 
  left_join(peers_data,  by = "date")
```

# Rolling Beta Analysis

```{r rolling-beta, error=TRUE}
# Prepare data for PerformanceAnalytics
Ra_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>% 
  mutate(across(-date, ~ coalesce(., 0))) %>% 
  tbl_xts(
    cols_to_xts = c("ABC_Fund", "Peer_Average")
  )

Rb_xts <- 
  analysis_data %>% 
  arrange(date) %>% 
  select(
    date,
    Benchmark = Benchmark_Returns
  ) %>% 
  mutate(Benchmark = coalesce(Benchmark, 0)) %>% 
  tbl_xts(
    cols_to_xts = "Benchmark"
  )

# Rolling 3 year betas (36 months) vs benchmark
PerformanceAnalytics::chart.RollingRegression(
  Ra         = Ra_xts,
  Rb         = Rb_xts,
  width      = 36,
  attribute  = "Beta",
  legend.loc = "topleft",
  main       = "Rolling 3 year beta vs benchmark: ABC vs peer average"
)
```

```{r boxplot-beta, error=TRUE}
# Compute rolling 3 year betas (36 months) vs benchmark
beta_data <- analysis_data %>% arrange(date)

# ABC vs Benchmark
roll_abc <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$ABC_Returns,
    width = 36
  )

# Peer average vs Benchmark
roll_peer <- 
  roll::roll_lm(
    x     = as.matrix(beta_data$Benchmark_Returns),
    y     = beta_data$Peers_Avg_Returns,
    width = 36
  )

# Extract slope (beta) from coefficients[, 2]
rolling_betas <- 
  tibble(
    date        = beta_data$date,
    ABC_Fund    = as.numeric(roll_abc$coefficients[, 2]),
    Peer_Average = as.numeric(roll_peer$coefficients[, 2])
  ) %>% 
  pivot_longer(
    cols      = c(ABC_Fund, Peer_Average),
    names_to  = "Series",
    values_to = "Beta"
  ) %>% 
  filter(!is.na(Beta))

# Boxplot with scatter (jitter) on top
ggplot(rolling_betas, aes(x = Series, y = Beta)) +
  geom_jitter(width = 0.1, alpha = 0.4) +
  geom_boxplot(width = 0.3, alpha = 0.3, outlier.shape = NA) +
  labs(
    title = "Distribution of rolling 3 year betas vs benchmark",
    x     = NULL,
    y     = "Rolling 3 year beta"
  ) +
  theme_fmx()
```

# Performance Analysis

```{r performance-tables, error=TRUE}
# Combine benchmark, ABC and peer average into one xts object
returns_xts <- merge(Rb_xts, Ra_xts, join = "inner")
colnames(returns_xts) <- c("Benchmark", "ABC_Fund", "Peer_Average")

# Summary statistics
PerformanceAnalytics::table.Stats(returns_xts)

# Trailing-period performance
PerformanceAnalytics::table.TrailingPeriods(
  R       = returns_xts,
  periods = c(3, 6, 12, 36, 60)
)

# Single-factor model and CAPM-style alpha/beta vs benchmark
PerformanceAnalytics::table.SFM(
  Ra = returns_xts[, c("ABC_Fund", "Peer_Average")],
  Rb = returns_xts[, "Benchmark"],
  Rf = 0
)
```

```{r cumulative-returns, error=TRUE}
# Cumulative returns plot
cum_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    ret    = coalesce(ret, 0),
    CumRet = cumprod(1 + ret)
  ) %>%
  ungroup()

g_cum <-
  cum_df %>%
  ggplot(aes(x = date, y = CumRet, colour = Series)) +
  geom_line() +
  labs(
    title = "Cumulative returns: ABC Fund vs Peers(Average), Benchmark",
    x     = "",
    y     = "Growth of 1 rand"
  ) +
  theme_fmx()

finplot(
  g_cum,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = FALSE
)
```

```{r rolling-returns, error=TRUE}
# Rolling 3-year returns
roll_df <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    ret      = coalesce(ret, 0),
    RollRets = RcppRoll::roll_prod(1 + ret, 36, fill = NA, align = "right")^(12 / 36) - 1
  ) %>%
  group_by(date) %>%
  filter(any(!is.na(RollRets))) %>%
  ungroup()

g_roll <-
  roll_df %>%
  ggplot(aes(x = date, y = RollRets, colour = Series)) +
  geom_line(alpha = 0.7, linewidth = 1.1) +
  labs(
    title = "Rolling 3-year annualised returns",
    x     = "",
    y     = "Rolling 3-year return (annualised)"
  ) +
  theme_fmx()

finplot(
  g_roll,
  x.date.dist = "2 years",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)
```

```{r annualised-returns, error=TRUE}
# Annualised returns across different horizons
idx <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  arrange(date) %>%
  group_by(Series) %>%
  ungroup()

dfplot <-
  bind_rows(
    # 6 months
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 6)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE) - 1) %>%
      mutate(Freq = "A"),
    # 12 months
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 12)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 12) - 1) %>%
      mutate(Freq = "B"),
    # 3 years
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 36)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 36) - 1) %>%
      mutate(Freq = "C"),
    # 5 years
    idx %>%
      filter(date >= fmxdat::safe_month_min(last(date), N = 60)) %>%
      group_by(Series) %>%
      summarise(mu = prod(1 + ret, na.rm = TRUE)^(12 / 60) - 1) %>%
      mutate(Freq = "D")
  )

to_string <- as_labeller(
  c(
    `A` = "6 Months",
    `B` = "1 Year",
    `C` = "3 Years",
    `D` = "5 Years"
  )
)

g_ann <-
  dfplot %>%
  ggplot() +
  geom_bar(aes(Series, mu, fill = Series), stat = "identity") +
  facet_wrap(~Freq, labeller = to_string, nrow = 1) +
  labs(
    x       = "",
    y       = "Returns (Ann.)",
    caption = "Note:\nReturns in excess of a year are in annualised terms."
  ) +
  fmx_fills() +
  geom_label(
    aes(Series, mu, label = paste0(round(mu, 4) * 100, "%")),
    size        = ggpts(8),
    alpha       = 0.35,
    fontface    = "bold",
    nudge_y     = 0.002
  ) +
  theme_fmx(
    CustomCaption = TRUE,
    title.size    = ggpts(43),
    subtitle.size = ggpts(38),
    caption.size  = ggpts(30),
    axis.size     = ggpts(37),
    legend.size   = ggpts(35),
    legend.pos    = "top"
  ) +
  theme(
    axis.text.x   = element_blank(),
    axis.title.y  = element_text(vjust = 2),
    strip.text.x  = element_text(
      face   = "bold",
      size   = ggpts(35),
      margin = margin(.1, 0, .1, 0, "cm")
    )
  )

g_ann
```

```{r volatility, error=TRUE}
# Rolling volatility
plot_dlog <-
  analysis_data %>%
  arrange(date) %>%
  transmute(
    date,
    Benchmark    = Benchmark_Returns,
    ABC_Fund     = ABC_Returns,
    Peer_Average = Peers_Avg_Returns
  ) %>%
  pivot_longer(
    cols      = -date,
    names_to  = "Series",
    values_to = "ret"
  ) %>%
  group_by(Series) %>%
  mutate(
    logret = log(1 + ret),
    RollSD = RcppRoll::roll_sd(logret, 36, fill = NA, align = "right") * sqrt(12)
  ) %>%
  filter(!is.na(RollSD)) %>%
  ungroup()

g <-
  plot_dlog %>%
  ggplot() +
  geom_line(aes(date, RollSD, colour = Series), alpha = 0.7, linewidth = 1.25) +
  labs(
    title    = "Rolling 3-year annualised volatility",
    subtitle = "",
    x        = "",
    y        = "Rolling 3-year SD (annualised)",
    caption  = "Note:\nVolatility estimated from 36-month log returns, annualised."
  ) +
  theme_fmx(
    title.size    = ggpts(30),
    subtitle.size = ggpts(5),
    caption.size  = ggpts(25),
    CustomCaption = TRUE
  ) +
  fmx_cols()

finplot(
  g,
  x.date.dist = "1 year",
  x.date.type = "%Y",
  x.vert      = TRUE,
  y.pct       = TRUE,
  y.pct_acc   = 1
)
```

